### 2016-10-18 paper.tex
Initial commit

\documentclass{article}

\title{Boutiques: a flexible framework for automated pipeline integration in computing platforms}

\begin{document}

\author{Tristan Glatard, Tristan Aumentado-Armstrong, Natacha Beck, \\
        Pierre Bellec, Remi Bernard, John Flavin, Krzysztof J. Gorgolewski, \\
        Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux, \\
        Gunnar Schaeffer, Marc-\'Etienne Rousseau, Alan C. Evans}

\maketitle

\section{Introduction}

\section{System description}

\subsection{Command-line description}

\subsection{Configuration files}

\subsection{Containers}

\subsection{Repository}

\section{Evaluation}

\subsection{Implementation}

\subsection{Ported pipelines}

% cbrain-plugins-neuro
% PSOM
% MICCAI Challenges
% exports from Nipype
% SPM?

\section{Related work}

% CWL

\section{Discussion}

\section{Conclusion}

\end{document}

### 2016-10-18 paper.tex
More text

@@
 \documentclass{article}
+\usepackage{minted}
 
-\title{Boutiques: a flexible framework for automated pipeline integration in computing platforms}
+\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
+
+\title{Boutiques: a flexible framework for automated application integration in computing platforms}
 
@@
 \author{Tristan Glatard, Tristan Aumentado-Armstrong, Natacha Beck, \\
-        Pierre Bellec, Remi Bernard, John Flavin, Krzysztof J. Gorgolewski, \\
+  Pierre Bellec, Remi Bernard, Sorina Camarasu-Pop, \\
+  Fr\'ed\'eric Cervenansky, Rafael Ferreira da Silva, \\ 
+  John Flavin, Pascal Girard, Krzysztof J. Gorgolewski, \\
   Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux, \\
@@
 
+Why Linux. Define application. Define computing platforms. Explain the problem and why it is difficult.
+
 \section{System description}
+
+\begin{itemize}
+\item JSON manifest
+\item Link to container
+\item Invocation schema
+\end{itemize}
 
@@
 
+The core component of the tool JSON manifest is a flexible
+command-line description represented as a simple string complying to
+the syntax of the \texttt{bash} Linux shell interpreter. The
+command-line string may contain placeholders for input and output
+values. The placeholders may be any string, and the command line may
+actually encompass several commands, e.g., separated by semicolumns,
+pipes (\texttt{|}) or ampersands (\texttt{\&}). Such a flexible
+command-line specification is meant to facilitate minor adjustments on
+the command-line, for instance input decompression and output
+archival, avoiding the use of multiple ``shims'' to align mismatching
+applications in a pipeline~\cite{hull2004treating}.
+
+Here is an example of a typical command-line description:
+\begin{verbatim}
+exampleTool_1 [STRING_INPUT] [FILE_INPUT] [ENUM_INPUT] | \
+exampleTool_2 [FLAG_INPUT] [NUMBER_INPUT] >> [LOG].txt
+\end{verbatim}
+The command line description contains six command-line keys for input
+and output values. When the application is executed, such keys will be
+replaced by numerical values and file names according to the user
+input. No particular syntax is imposed for the command-line keys, they
+just have to be unique. Flags will also be added wherever
+appropriate. Note the use of the pipe operator to chain tools, and of
+the \texttt{>>} bash operator that allows to redirect the standard
+output in a particular file.
+
+\subsection{Input description}
+
+\paragraph{General properties.} Inputs must have a name, a unique
+identifier and a type. They may be optional, have a description, a
+command-line key (placeholder), a flag and flag separator, and a
+default value. Inputs may also be lists.
+
+\paragraph{Types.} Inputs may be of type \texttt{String},
+\texttt{Number}, \texttt{Flag} or \texttt{File}. Types can be
+restricted to a specific set of values (\texttt{String} and
+\texttt{Number} only), to a specific range (\texttt{Number} only) to
+integers or booleans (\texttt{Number} only).
+
+\paragraph{Groups and dependencies.} Groups of inputs may be defined
+from a identifier, name and list of input identifiers. Groups may be
+used for presentation purposes in a graphical user interface and to
+specify the following constraints among inputs: (1)
+\texttt{mutually-exclusive}: only one input in the group may have a
+value (2) \texttt{one-is-required}: at least one of the inputs in the
+group must have a value. Dependencies among inputs may also be defined
+regardless of a particular group: a specific input may (1) require a
+list of other inputs to be active to be available and (2) define a
+list of inputs that are disabled when it is active.
+
+Listing~\ref{listing:input-example} shows the definition of an input
+in the command line exemplified above. According to this definition
+and assuming that the input value entered by the user is 0.3, the
+string \texttt{[NUMBER\_INPUT]} will be replaced by \texttt{-n=0.3} on
+the command line at execution time.
+\begin{listing}
+\begin{minted}[frame=single,
+               framesep=3mm,
+               linenos=false,
+               xleftmargin=21pt,
+               tabsize=4]{js}
+{
+    ``id'' : ``num_input'',
+    ``name'' : ``A number input'',
+    ``type'' : ``Number'',
+    ``command-line-key'' : ``[NUMBER_INPUT]'',
+    ``optional'' : true,
+    ``command-line-flag'' : ``-n'',
+    ``command-line-separator'' : ``='',
+    ``minimum'' : 0,
+    ``maximum'' : 1,
+    ``exclusive-minimum'' : true,
+    ``exclusive-maximum'' : false
+  }
+\end{minted}
+\caption{Input example} 
+\label{listing:input-example}
+\end{listing}
+
+\subsection{Output description}
+
 \subsection{Configuration files}
+
+\subsection{Workflow support}
 
@@
 
+The implementation of the application is defined in a container image
+in the Docker, Singularity or rootfs format. We intentionally support
+multiple container formats as we anticipate that they will be used for
+different purposes. For instance, Docker is well suited for
+application developers and users who want to use applications on their
+local workstation. It is very well documented, maintained and it has a
+rich ecosystem of tools to help build and run containers on most
+platforms, including Windows, MacOS and Linux. Singularity is more
+suited for users and platforms that need to run applications on shared
+computing clusters. Bridges exist among these containers formats to
+convert container images across frameworks. For instance, a platform
+dedicated to high-performance computing may accept manifests referring
+to Docker containers to facilitate application integration by
+developers, and internally convert container images to Singularity to
+run applications efficiently on clusters.
+
+Container images are defined from their URL (Singularity and rootfs)
+or name in a Docker index. Container images may specify a working
+directory where the application has to be run. A hash may also be
+reported to accurately identify container images and detect updates.
+
+\subsection{Custom properties}
+
+\section{Implementation and Evaluation}
+
+\subsection{Core tools} Validator, local executor, invocation schema generator.
+
+\subsection{Platforms}
+
+\paragraph{CBRAIN}
+
+\paragraph{Pegasus}
+
+\paragraph{VIP}
+
 \subsection{Repository}
-
-\section{Evaluation}
-
-\subsection{Implementation}
 
@@
 
+A table summarizing which Boutiques features are used in which pipelines.
+
 \section{Related work}
@@
 
+\section{Acknowledgments}
+
+Amazon grant.
+
+\section{References}
+\bibliographystyle{plain}
+\bibliography{biblio}
+
 \end{document}

### 2016-10-19 paper.tex
More text

@@
 \usepackage{minted}
+\usepackage{pdfcomment}
 
 \newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
+\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
 
@@
 
-Why Linux. Define application. Define computing platforms. Explain the problem and why it is difficult.
+Why Linux. Define application. Define computing platforms. Explain the
+problem and why it is difficult. Approach: start simple and enrich the
+spec when new applications come. Refer to computing infrastructures
+and the need for parallelization. Define task vs. invocation. Who are
+the developers. Refer to neuroinformatics field.
 
@@
 command-line key (placeholder), a flag and flag separator, and a
-default value. Inputs may also be lists.
+default value. Inputs may also be ordered lists.
 
 \paragraph{Types.} Inputs may be of type \texttt{String},
-\texttt{Number}, \texttt{Flag} or \texttt{File}. Types can be
-restricted to a specific set of values (\texttt{String} and
-\texttt{Number} only), to a specific range (\texttt{Number} only) to
-integers or booleans (\texttt{Number} only).
+\texttt{Number}, \texttt{Flag} or \texttt{File}. \texttt{File} may
+also represent a directory. Types can be restricted to a specific set
+of values (\texttt{String} and \texttt{Number} only), to a specific
+range (\texttt{Number} only) to integers or booleans (\texttt{Number}
+only).
 
@@
 
+Application outputs are the files or directories that need to be
+transferred and delivered to the user once the execution is
+complete. While it may not be required to distinguish inputs and
+outputs to build the application command line, this information is
+needed by computing platforms to determine what needs to be saved from
+the execution.
+
+In Boutiques, output files are defined from a unique identifier, a
+name and a path template that specifies the file or directory
+name. Path templates may include input command-line keys when output
+files are named after the input values. Input values may be stripped
+from specific strings before being substituted in the path template,
+e.g., file extensions. Output files may also have a description, a
+command-line flag, a flag separator, and a command-line key in case
+they appear on the command line. They may be optional in case the file
+is not always produced by the application (for instance when it is
+produced only when a particular flag is activated). They may also be
+unordered lists; in this case, the path template must contain a
+wildcard (\texttt{*}) standing for any string of characters (as in
+\texttt{bash}) and defining the pattern used to match the output files
+in the list.
+
+Listing~\ref{listing:output-example} shows the definition of an output
+file in the command line exemplified before. According to this
+definition and assuming that the string input value entered by the
+user is "foo.csv", the string \texttt{[LOG]} will be
+replaced by \texttt{log-foo} on the command line at execution time.
+
+\begin{listing}
+\begin{minted}[frame=single,
+               framesep=3mm,
+               linenos=false,
+               xleftmargin=21pt,
+               tabsize=4]{js}
+
+{
+    "id" : "logfile",
+    "name" : "Log file",
+    "description" : "The output log file from the example tool",
+    "path-template" : "log-[STRING_INPUT]",
+    "command-line-key" : "[LOG]",
+    "path-template-stripped-extensions" : [".txt", ".csv"],
+    "optional" : false
+}
+\end{minted}
+\caption{Output file example} 
+\label{listing:output-example}
+\end{listing}
+
+\subsection{Invocation schema}
+\label{sec:invocation-schema}
+
+Input validation is an important motivation for common application
+descriptions. The obvious solution in our context is to rely on an
+application-specific JSON schema, called ``invocation schema'' to
+specify the correct input values that an application
+accepts. Platforms could leverage the invocation schema to validate
+inputs automatically and without having to develop specific code,
+using any JSON validator.
+
+Invocation schemas, however, are complex JSON objects. Basically, they
+must represent the properties described above in a formal way,
+including dependencies between inputs. To illustrate such complexity,
+listing~\ref{listing:invocation-schema-example} displays
+\todo{Tristan}{...}. % must support dependent invocation ids to allow for workflow support.
+
+It does not seem realistic to assume that
+developers will easily write correct JSON schemas for their
+applications. In addition, invocation schemas can actually be
+generated automatically from the basic application description. Thus,
+the invocation schema is an optional property of the Boutiques schema,
+that platforms or developers may generate automatically using a core
+tool provided by Boutiques.
+
 \subsection{Configuration files}
 
+A large number of applications actually rely on configuration files
+rather than command-line options to define their input and output
+parameters. Indeed, when the number of parameters increases,
+command-line may become long and cumbersome while configuration files
+allow for better structure and documentation.
+
+Configuration files may be complex though, and specified in any
+language.  As it does not seem reasonable to specify a particular
+template for configuration files, the Boutiques specification allows
+application developers to specify their own template, referring to
+input and output command-line keys \todo{Tristan}{Then they are not
+  command-line keys any more} to specify placeholders for input and
+output values. Listing~\ref{listing:configuration-file-example} shows
+an example. Configuration files also have a path template that defines
+how they must be named and where they must be written. They may also
+have a command line key and flag in case they need to be passed on the
+command-line. 
+\todo{Tristan}{See Github issue \#17}
+
 \subsection{Workflow support}
+
+Boutiques intentionally does not support the definition of complete
+workflows from individual applications. Indeed, we believe that
+workflow engines are specific tools that need complex languages to
+implement the particular functions that they target. Efforts to
+specify common workflow languages include the Common Workflow Language
+(CWL)~\cite{cwl} and the Interoperable Workflow Intermediate
+Representation (IWIR)~\cite{plankensteiner2011iwir}.
+
+Instead, Boutiques allows to wrap complete workflows as simple
+applications depending on particular workflow engine(s) installed in
+container images. In this model, workflows are executed as any other
+task on the infrastructure, which allows for basic data parallelism by
+iterating the workflow on a set of independent files (for instance:
+subjects in an analysis). Data parallelism, however, may not be
+sufficient when the analysis of a particular file has to be
+parallelized. To address this issue, a specific property of the
+Boutiques schema specifies that an application may submit new
+tasks. New tasks are created by the application by writing a JSON
+object in the execution directory, complying to the invocation schema
+described in Section~\ref{sec:invocation-schema}. The platform must
+read this object, submit the corresponding task, and write back the
+invocation identifier in the execution directory. The invocation
+identifier may be used by the workflow to specify dependencies between
+tasks in subsequent task submissions.
+
+The adopted ``workflow encapsulation'' model is very simple but it
+allows for a scalable and reliable execution of workflows expressed in
+a variety of languages, as detailed in ~\cite{glatard2016fgcs}.
 
@@
 
+Containers are adopted because they allow for an automated and
+lightweight integration of application implementations in
+platforms. They are extremely useful to improve the reproducibility of
+analyses as variations in the software environment may have an
+important impact on the computed results. They also have limitations,
+in particular they do not specify the hardware architecture required
+to execute an application, which can be an issue in some cases.
+
 \subsection{Custom properties}
+
+Custom properties may be added to the Boutiques specification without
+restriction. They may be useful to implement platform-specific
+features but they should be used with care to avoid making tools
+dependent on a particular platform. The following custom properties
+have been used so far:
+\begin{itemize}
+\item \texttt{cbrain:can-submit-new-tasks}: a boolean, true if the tool may submit new tasks to the platform. \todo{Tristan}{related to the workflow support section}
+\item \texttt{cbrain:inherits-from-class}: a string that defines the Ruby class that should be used as parent class for the
+ tool in CBRAIN. Used to define a progress bar for PSOM tools in CBRAIN.
+\item \texttt{vip:miccai-challenger-email} and
+  \texttt{vip:miccai-challenge-team-name}: strings helping VIP
+  categorize tools for the 2016 MICCAI challenges ``MSSEG'' and
+  ``PETSEG''.
+\end{itemize}
+
+\subsection{Command-line construction}
 
@@
 % SPM?
+% BIDS Apps
 

### 2016-10-20 paper.tex
More text

@@
   Pierre Bellec, Remi Bernard, Sorina Camarasu-Pop, \\
-  Fr\'ed\'eric Cervenansky, Rafael Ferreira da Silva, \\ 
+  Fr\'ed\'eric Cervenansky, Samir Das, Rafael Ferreira da Silva, \\ 
   John Flavin, Pascal Girard, Krzysztof J. Gorgolewski, \\
@@
 spec when new applications come. Refer to computing infrastructures
-and the need for parallelization. Define task vs. invocation. Who are
+and the need for parallelization. Who are
 the developers. Refer to neuroinformatics field.
+
+Vocabulary (check consistency across the paper):
+\begin{itemize}
+\item tool vs. application.
+\item descriptor vs. manifest.
+\item task vs. invocation.
+\end{itemize}
 
@@
 
-\subsection{Workflow support}
+\subsection{Parallelization support}
 
@@
 a variety of languages, as detailed in ~\cite{glatard2016fgcs}.
+
+%\todo{rewrite this section to focus more on parallelism. Start by mentioning data parallelism.}
 
@@
 
-\subsection{Core tools} Validator, local executor, invocation schema generator.
+\subsection{Core tools}
+
+The Boutiques validator checks conformance of JSON manifests to the
+Boutiques schema using a basic JSON validator. It also performs an
+additional set of checks that cannot be easily implemented in JSON
+schema: command-line keys are unique among inputs, input and output
+identifiers are unique, input and output command-line keys are all
+included in the command-line, command-line keys are not contained
+within each other (which would puzzle substitution), output path
+templates are unique (to avoid results to overwrite each other),
+inputs of type Flag have a command-line flag, are optional and are not
+lists, number and list constraints are sensible (e.g. min is lower
+than max), the default value of enum types is part of the enum, an
+input cannot both require and disable another input, required inputs
+cannot require or disable other parameters, group member identifiers
+must correspond to existing inputs and cannot appear in different
+groups, mutually exclusive groups cannot have members requiring other
+members, and that one-is-required groups should never have required
+members. The Boutiques validator is meant for application developers.
+
+The local executor can generate and execute command line from a
+Boutiques manifest and a set of input values represented in a CSV or
+JSON file complying with the invocation schema. It runs the
+command in a container provided that the required framework
+(e.g. Docker) is already installed. It can also generate hypothetical
+command lines from random values, for debugging purposes. The local
+executor can be used by application users, to run tools locally, or by
+platforms, to generate command lines to be run on the execution
+infrastructure.
+
+The invocation schema generator creates an invocation schema (see
+Section~\ref{sec:invocation-schema}) from a Boutiques manifest and
+validates data against it. It can be used by platform administrators
+to add invocation schemas to existing manifests, and by platforms to
+validate input data. Validation is done using a basic JSON
+validator.
 
@@
 
-\subsection{Ported pipelines}
+\subsection{Ported applications}
 
 % cbrain-plugins-neuro
+
+The Pipeline System for Octave and Matlab
+(PSOM)~\cite{bellec2012pipeline} has been integrated in CBRAIN using
+Boutiques. More precisely, a manifest has been written for a Niak
+pre-processing pipeline and PSOM workers. The integration uses the
+parallelization mechanism described in
+Section~\ref{sec:parallelization} so that even the tasks processing a
+given subject can be parallelized. As shown in~\cite{GLAT-16}, this
+allows CBRAIN to leverage the efficient agent model used in PSOM. The
+integration required some work in PSOM to facilitate its invocation as
+a non-interactive command-line application. The resulting CBRAIN
+plugin is available at
+\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It could in
+principle be used in other platforms that support the parallelization
+property, although this has not been tested.
+
 % PSOM
@@
 
+\begin{tabular}{ccccccccccccc}
+  Application & \multicolumn{3}{c}{Containers} & \multicolumn{7}{c}{Input/Output properties}                         & \multicolumn{2}{c}{Other properties}  \\
+              & Docker & Singularity & Rootfs  & Dependencies & Groups & Lists & Optional & Default & Enum & Min/max & Env vars & Min/Max \\
+
+PSOM & 
+\end{tabular}
+
 \section{Related work}
@@
 % CWL
+The Common Workflow Language (CWL) is the closest related work to
+Boutiques, in the sense that it also allows the description of command
+lines executed in containers. According to Github, the CWL project
+started 6 months before Boutiques. The main conceptual differences
+between CWL and Boutiques are the following:
+\begin{itemize}
+\item CWL has a workflow language, Boutiques does not. CWL also has a
+  command line tool description language, but its specification is
+  influenced by the fact that command line tools are included in a
+  specific workflow languages. In Boutiques, the only available
+  workflow construct is sub-tasking (``a task may submit other
+  tasks''). This is intentional as we envisage that Boutiques tools
+  may be composed through any type of workflow engine and language. For instance, this leads to have multiple output types in CWL.
+\item CWL specifies command lines formally whereas in Boutiques they
+  are assembled from simple string templates. In short, a command line
+  in CWL is specified by an executable and an ordered list of
+  parameters. While this representation has the advantage of being
+  more formal, for instance the executable can be identified
+  automatically, it also removes flexibility by imposing that a tool
+  only runs a single command. On the contrary, Boutiques allows any
+  type of bash operators to be combined on the command line. This
+  level of flexibility is important and greatly reduces the ``shim
+  syndrome''. Boutiques' template approach allows for configuration files. 
+\item CWL is validated using SALAD, which makes the ecosystem slightly
+  more complex. For instance, CWL has ``import'' and ``include''
+  directives. Boutiques is validated using plain JSON schema. \todo{ontology layer}.
+\item In CWL, the applications need to be instrumented as follows, which is not the case in Boutiques:
+\begin{itemize}
+\item Applications must write in specific directories.
+\item Applications must produce a JSON object describing the output structure.
+\end{itemize}
+\end{itemize}
+
+\todo{Tristan}{Update from comparison summary on Google doc}
 

### 2016-10-28 paper.tex
Worked on configuration files and parallelization

@@
 \usepackage{pdfcomment}
+\usepackage[margin=1in]{geometry}
 
@@
 
-\author{Tristan Glatard, Tristan Aumentado-Armstrong, Natacha Beck, \\
-  Pierre Bellec, Remi Bernard, Sorina Camarasu-Pop, \\
-  Fr\'ed\'eric Cervenansky, Samir Das, Rafael Ferreira da Silva, \\ 
-  John Flavin, Pascal Girard, Krzysztof J. Gorgolewski, \\
-  Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux, \\
-  Gunnar Schaeffer, Marc-\'Etienne Rousseau, Alan C. Evans}
+\author{Tristan Glatard, Tristan Aumentado-Armstrong, Natacha Beck, Pierre Bellec,\\
+        Remi Bernard, Sorina Camarasu-Pop, Fr\'ed\'eric Cervenansky, Samir Das, \\
+        Rafael Ferreira da Silva,John Flavin, Pascal Girard, Krzysztof J. Gorgolewski, \\
+        Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux, Gunnar Schaeffer,\\
+        Marc-\'Etienne Rousseau, Alan C. Evans}
 
@@
 
+Define ``the platform'', i.e. the entity that will run applications.
 \begin{itemize}
@@
 
-The core component of the tool JSON manifest is a flexible
-command-line description represented as a simple string complying to
-the syntax of the \texttt{bash} Linux shell interpreter. The
-command-line string may contain placeholders for input and output
-values. The placeholders may be any string, and the command line may
-actually encompass several commands, e.g., separated by semicolumns,
-pipes (\texttt{|}) or ampersands (\texttt{\&}). Such a flexible
-command-line specification is meant to facilitate minor adjustments on
-the command-line, for instance input decompression and output
-archival, avoiding the use of multiple ``shims'' to align mismatching
-applications in a pipeline~\cite{hull2004treating}.
+The core component of the JSON manifest is a flexible command-line
+description represented as a simple string template complying to the
+syntax of the \texttt{bash} Linux shell interpreter. The command line
+template may contain placeholders for input and output values, called
+value keys. The value keys may be any string, and the
+command line may actually encompass several commands, e.g., separated
+by semicolumns, pipes (\texttt{|}) or ampersands (\texttt{\&}). Such a
+flexible command-line specification is meant to facilitate the
+embedding of minor operations on the command line, for instance input
+decompression and output archival, avoiding the use of multiple
+``shims'' to align mismatching applications in a
+pipeline~\cite{hull2004treating}.
 
-Here is an example of a typical command-line description:
+Here is an example of a typical command-line template:
 \begin{verbatim}
@@
 \end{verbatim}
-The command line description contains six command-line keys for input
-and output values. When the application is executed, such keys will be
-replaced by numerical values and file names according to the user
-input. No particular syntax is imposed for the command-line keys, they
-just have to be unique. Flags will also be added wherever
-appropriate. Note the use of the pipe operator to chain tools, and of
-the \texttt{>>} bash operator that allows to redirect the standard
-output in a particular file.
+The template contains six value keys for input and output
+values. When the application is executed, such keys will be replaced
+by values and file names according to the user input. Flags will also
+be added wherever appropriate, with customizable separators. No
+particular syntax is imposed for the value keys, they just have
+to be unique.  Note the use of the pipe (\texttt{|}) operator to chain
+tools, and of the \texttt{>>} operator to redirect the standard output
+to a file.
 
@@
 identifier and a type. They may be optional, have a description, a
-command-line key (placeholder), a flag and flag separator, and a
-default value. Inputs may also be ordered lists.
+value key, a flag and flag separator, and a default
+value. Inputs may also be ordered lists; in this case, value
+keys are substituted by the space-separated list of input values.
 
@@
 of values (\texttt{String} and \texttt{Number} only), to a specific
-range (\texttt{Number} only) to integers or booleans (\texttt{Number}
+range (\texttt{Number} only), to integers or to booleans (\texttt{Number}
 only).
@@
 from a identifier, name and list of input identifiers. Groups may be
-used for presentation purposes in a graphical user interface and to
+used to improve the presentation in a graphical user interface and to
 specify the following constraints among inputs: (1)
 \texttt{mutually-exclusive}: only one input in the group may have a
-value (2) \texttt{one-is-required}: at least one of the inputs in the
-group must have a value. Dependencies among inputs may also be defined
-regardless of a particular group: a specific input may (1) require a
-list of other inputs to be active to be available and (2) define a
-list of inputs that are disabled when it is active.
+value; (2) \texttt{one-is-required}: at least one input in the group
+must have a value. Dependencies among inputs may also be defined
+regardless of a particular group: an input may (1) require a list of
+inputs (2) disable a list of inputs.
 
@@
 {
-    ``id'' : ``num_input'',
-    ``name'' : ``A number input'',
-    ``type'' : ``Number'',
-    ``command-line-key'' : ``[NUMBER_INPUT]'',
-    ``optional'' : true,
-    ``command-line-flag'' : ``-n'',
-    ``command-line-separator'' : ``='',
-    ``minimum'' : 0,
-    ``maximum'' : 1,
-    ``exclusive-minimum'' : true,
-    ``exclusive-maximum'' : false
+    "id" : "num_input",
+    "name" : "A number input",
+    "type" : "Number",
+    "value-key" : "[NUMBER_INPUT]",
+    "optional" : true,
+    "command-line-flag" : "-n",
+    "command-line-separator" : "=",
+    "minimum" : 0,
+    "maximum" : 1,
+    "exclusive-minimum" : true,
+    "exclusive-maximum" : false
   }
 \end{minted}
-\caption{Input example} 
+\caption{Input example.} 
 \label{listing:input-example}
@@
 transferred and delivered to the user once the execution is
-complete. While it may not be required to distinguish inputs and
+complete. Although it may not be required to distinguish inputs from
 outputs to build the application command line, this information is
-needed by computing platforms to determine what needs to be saved from
-the execution.
+needed by computing platforms to identify the files that must be
+saved after the execution.
 
@@
 name and a path template that specifies the file or directory
-name. Path templates may include input command-line keys when output
-files are named after the input values. Input values may be stripped
-from specific strings before being substituted in the path template,
-e.g., file extensions. Output files may also have a description, a
-command-line flag, a flag separator, and a command-line key in case
-they appear on the command line. They may be optional in case the file
-is not always produced by the application (for instance when it is
-produced only when a particular flag is activated). They may also be
-unordered lists; in this case, the path template must contain a
-wildcard (\texttt{*}) standing for any string of characters (as in
-\texttt{bash}) and defining the pattern used to match the output files
-in the list.
+name. Path templates may include input value keys in case
+output files are named after the input values. Input values may be
+stripped from specific strings, e.g., file extensions before being
+substituted in the path template. Output files may also have a
+description, a command-line flag, a flag separator, and a value
+key in case they appear on the command line. They may be optional in
+case the file is not always produced by the application, for instance
+when it is produced only when a particular flag is activated. They
+may also be (un-ordered) lists; in this case, the path template must
+contain a wildcard (\texttt{*}) matching any string of characters
+and defining the pattern used to match the
+output files in the list.
 
@@
     "path-template" : "log-[STRING_INPUT]",
-    "command-line-key" : "[LOG]",
+    "value-key" : "[LOG]",
     "path-template-stripped-extensions" : [".txt", ".csv"],
@@
 
+\subsection{Configuration files}
+
+A large number of applications rely on configuration files rather than
+command-line options to define their input and output
+parameters. Indeed, as the number of parameters increases, command
+lines rapidly become long and cumbersome whereas configuration files
+allow for better structure and documentation.
+
+Configuration files may be complex though, and specified in any
+language.  As it does not seem reasonable to adopt a particular
+template for configuration files, the Boutiques specification allows
+application developers to specify their own template, referring to
+input and output value keys. Configuration files are specific types of
+output files that have a
+ file template defined. As output files, 
+they must have a path template that defines
+how they must be named and where they must be written. They may also
+have a value key and a flag in case they need to be passed on the
+command line.
+
+Listing~\ref{listing:configuration-file-example} shows an example.
+\begin{listing}
+\begin{minted}[frame=single,
+               framesep=3mm,
+               linenos=false,
+               xleftmargin=21pt,
+               tabsize=4]{js}
+               {
+                   "id": "config_file",
+                   "name": "Configuration file",
+                   "type": "Configuration File",
+                   "value-key": "[CONFIG_FILE]",
+                   "path-template": "config.txt",
+                   "file-template": [
+                   "# This input is hard-coded",
+                   "stringInput=foo",
+                   "fileInput=[FILE-INPUT]",
+                   "# And here is the result",
+                   "fileOutput=[OUTPUT-FILE]",
+                   ""
+                   ]
+                 }
+\end{minted}
+\caption{Configuration file example. The file template is defined as
+  an array of strings to allow for multi-line strings in JSON.}
+\label{listing:configuration-file-example}
+\end{listing}
+
 \subsection{Invocation schema}
@@
 
-Input validation is an important motivation for common application
-descriptions. The obvious solution in our context is to rely on an
-application-specific JSON schema, called ``invocation schema'' to
+Rigorous input validation is an important motivation for
+Boutiques. The obvious solution in our context is to rely on an
+application-specific JSON schema, called ``invocation schema'', to
 specify the correct input values that an application
-accepts. Platforms could leverage the invocation schema to validate
-inputs automatically and without having to develop specific code,
-using any JSON validator.
+accepts. Platforms could rely on the invocation schema to validate
+inputs automatically using any JSON validator, without having to
+develop specific code.
 
@@
 must represent the properties described above in a formal way,
-including dependencies between inputs. To illustrate such complexity,
-listing~\ref{listing:invocation-schema-example} displays
-\todo{Tristan}{...}. % must support dependent invocation ids to allow for workflow support.
+including dependencies between
+inputs. Listing~\ref{listing:invocation-schema-example} shows an
+example of how dependencies between mutually exclusive parameters are
+defined in the invocation schema.
 
-It does not seem realistic to assume that
-developers will easily write correct JSON schemas for their
-applications. In addition, invocation schemas can actually be
-generated automatically from the basic application description. Thus,
-the invocation schema is an optional property of the Boutiques schema,
-that platforms or developers may generate automatically using a core
-tool provided by Boutiques.
+The invocation schema also has a \texttt{task-dependencies} array that
+specifies that the new invocation to start only when all the
+invocations in this array have completed. This property is used in the
+parallelization model described hereafter.
 
-\subsection{Configuration files}
+\begin{listing}
+\begin{minted}[frame=single,
+               framesep=3mm,
+               linenos=false,
+               xleftmargin=21pt,
+               tabsize=4]{js}
+"dependencies" : {
+      "num_input" : {
+         "properties" : {
+            "str_input" : {
+               "not" : {}
+            }
+         }
+      },
+      "str_input" : {
+         "properties" : {
+            "num_input" : {
+               "not" : {}
+            }
+         }
+      }
+   }
+\end{minted}
+\caption{Excerpt from invocation schema showing dependencies between
+  two mutually exclusive parameters \texttt{num\_input} and
+  \texttt{str\_input}.}
+\label{listing:invocation-schema-example}
+\end{listing}
 
-A large number of applications actually rely on configuration files
-rather than command-line options to define their input and output
-parameters. Indeed, when the number of parameters increases,
-command-line may become long and cumbersome while configuration files
-allow for better structure and documentation.
-
-Configuration files may be complex though, and specified in any
-language.  As it does not seem reasonable to specify a particular
-template for configuration files, the Boutiques specification allows
-application developers to specify their own template, referring to
-input and output command-line keys \todo{Tristan}{Then they are not
-  command-line keys any more} to specify placeholders for input and
-output values. Listing~\ref{listing:configuration-file-example} shows
-an example. Configuration files also have a path template that defines
-how they must be named and where they must be written. They may also
-have a command line key and flag in case they need to be passed on the
-command-line. 
-\todo{Tristan}{See Github issue \#17}
+It does not seem realistic to assume that developers will easily write
+correct JSON schemas for their applications. Instead, invocation
+schemas should be generated automatically using a core Boutiques
+tool. Accordingly, invocation schema is an optional property of the
+Boutiques manifest, that platforms or developers may generate using
+the appropriate tool.
 
 \subsection{Parallelization support}
+\label{sec:parallelization}
+Boutiques support parallelism by allowing applications to submit new
+tasks through the following protocol:
+\begin{enumerate}
+\item Applications write a JSON object complying to the invocation schema in the directory where there were launched.
+\item The platform reads the JSON object, creates the corresponding task, and writes back the task id in the same directory.
+% The platform could also support status requests.
+\end{enumerate}
+Several invocations can be submitted at once, possibly using the
+\texttt{task-dependencies} array to specify dependencies among
+them. This mechanism is implemented in the Boutiques schema using a
+single property, \texttt{can-submit-tasks}.
 
-Boutiques intentionally does not support the definition of complete
-workflows from individual applications. Indeed, we believe that
-workflow engines are specific tools that need complex languages to
-implement the particular functions that they target. Efforts to
-specify common workflow languages include the Common Workflow Language
-(CWL)~\cite{cwl} and the Interoperable Workflow Intermediate
-Representation (IWIR)~\cite{plankensteiner2011iwir}.
+Although very simple, this model allows wrapping complete
+workflows. Workflows are wrapped as any other application, and they
+can be executed using any engine such as Nipype, PSOM and others,
+provided that such engines are installed in the container image
+referred in the application manifest.
 
-Instead, Boutiques allows to wrap complete workflows as simple
-applications depending on particular workflow engine(s) installed in
-container images. In this model, workflows are executed as any other
-task on the infrastructure, which allows for basic data parallelism by
-iterating the workflow on a set of independent files (for instance:
-subjects in an analysis). Data parallelism, however, may not be
-sufficient when the analysis of a particular file has to be
-parallelized. To address this issue, a specific property of the
-Boutiques schema specifies that an application may submit new
-tasks. New tasks are created by the application by writing a JSON
-object in the execution directory, complying to the invocation schema
-described in Section~\ref{sec:invocation-schema}. The platform must
-read this object, submit the corresponding task, and write back the
-invocation identifier in the execution directory. The invocation
-identifier may be used by the workflow to specify dependencies between
-tasks in subsequent task submissions.
-
-The adopted ``workflow encapsulation'' model is very simple but it
+The adopted ``task encapsulation'' model is very simple but it
 allows for a scalable and reliable execution of workflows expressed in
 a variety of languages, as detailed in ~\cite{glatard2016fgcs}.
-
-%\todo{rewrite this section to focus more on parallelism. Start by mentioning data parallelism.}
 
@@
 \begin{itemize}
-\item \texttt{cbrain:can-submit-new-tasks}: a boolean, true if the tool may submit new tasks to the platform. \todo{Tristan}{related to the workflow support section}
 \item \texttt{cbrain:inherits-from-class}: a string that defines the Ruby class that should be used as parent class for the
@@
 additional set of checks that cannot be easily implemented in JSON
-schema: command-line keys are unique among inputs, input and output
-identifiers are unique, input and output command-line keys are all
-included in the command-line, command-line keys are not contained
+schema: value keys are unique among inputs, input and output
+identifiers are unique, input and output value keys are all
+included in the command line, value keys are not contained
 within each other (which would puzzle substitution), output path
@@
 
+Boutiques intentionally does not support the definition of complete
+workflows from individual applications. Indeed, we believe that
+workflow engines are specific tools that need complex languages to
+implement the particular functions that they target. Efforts to
+specify common workflow languages include the Common Workflow Language
+(CWL)~\cite{cwl} and the Interoperable Workflow Intermediate
+Representation (IWIR)~\cite{plankensteiner2011iwir}.
+
 % CWL
@@
 
+% How to build modular container images for workflows.
+
 \section{Conclusion}
@@
 
-\section{References}
 \bibliographystyle{plain}

### 2016-10-31 paper.tex
Added application table

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{lscape}
+\usepackage[table]{xcolor}
 
@@
 \begin{enumerate}
-\item Applications write a JSON object complying to the invocation schema in the directory where there were launched.
-\item The platform reads the JSON object, creates the corresponding task, and writes back the task id in the same directory.
+\item Applications write a JSON object complying to the invocation
+  schema in the directory where there were launched.
+\item The platform reads the JSON object, creates the corresponding
+  task, and writes back the task id in the same directory.
 % The platform could also support status requests.
@@
 
+Containers are not mandatory though. If an application manifest does
+not contain any reference to a container, the platform will assume
+that the application is pre-installed on the execution infrastructure.
+
 \subsection{Custom properties}
@@
 \begin{itemize}
-\item \texttt{cbrain:inherits-from-class}: a string that defines the Ruby class that should be used as parent class for the
- tool in CBRAIN. Used to define a progress bar for PSOM tools in CBRAIN.
+\item \texttt{cbrain:inherits-from-class}: a string that defines the
+  Ruby class that should be used as parent class for the tool in
+  CBRAIN. Used to define a progress bar for PSOM tools in CBRAIN.
 \item \texttt{vip:miccai-challenger-email} and
@@
 
+New list mechanism.
+
 \paragraph{Pegasus}
@@
 
-A table summarizing which Boutiques features are used in which pipelines.
-
-\begin{tabular}{ccccccccccccc}
-  Application & \multicolumn{3}{c}{Containers} & \multicolumn{7}{c}{Input/Output properties}                         & \multicolumn{2}{c}{Other properties}  \\
-              & Docker & Singularity & Rootfs  & Dependencies & Groups & Lists & Optional & Default & Enum & Min/max & Env vars & Min/Max \\
-
-PSOM & 
+Table~\ref{table:applications} summarizes the applications described
+with Boutiques and the main features used.
+\begin{landscape}
+  \begin{table}
+  \rowcolors{2}{gray!25}{white}
+  \begin{tabular}{l|ccc|ccccccc|cc}
+    Application               &\multicolumn{3}{c|}{Containers} & \multicolumn{7}{c|}{Input/Output properties}                         & \multicolumn{2}{c}{Other properties}  \\
+                              &Docker              & Singularity & Rootfs  & Dependencies      &Groups             & Lists             & Optional          &Default            & Enum              & Min/max           & Config.
+file & Env vars     \\
+    \hline
+    Niak pre-processing       &\cellcolor{gray!75} &            &         &                   &\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&              &\cellcolor{gray!75}\\
+    ICA AROMA                 &\cellcolor{gray!75} &            &         &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&              &\\
+    FSL tools &&&&&&&&&&&&\\
+    \multicolumn{1}{r|}{anat} &                    &            &         &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&              &\\
+    \multicolumn{1}{r|}{bet}  &                    &            &         &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}&              &\\
+    \multicolumn{1}{r|}{fast} &                    &            &         &\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&              &\\
+    \multicolumn{1}{r|}{first}&                    &            &         &                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   &              &\\
+    FSL Nipype interfaces     \\
+    HCP PreFreesurfer         &\cellcolor{gray!75} &            &         &                   &                   &                   &                   &                   &                   &                   &              &\\
+    MSSEG MICCAI pipelines    \\
+    PETSEG MICCAI pipelines   \\
 \end{tabular}
+\caption{Applications available in Boutiques with manifest features used.}
+\label{table:applications}
+  \end{table}
+  \end{landscape}
 

### 2016-11-01 paper.tex
Wrote command line construction

@@
 \documentclass{article}
-\usepackage{minted}
-\usepackage{pdfcomment}
-\usepackage[margin=1in]{geometry}
-\usepackage{lscape}
-\usepackage[table]{xcolor}
+
+%%%% Packages %%%%
+
+\usepackage{minted} % Used for JSON highlighting
+\usepackage{pdfcomment} % Used for notes and todos
+\usepackage[margin=1in]{geometry} % Increases margins
+\usepackage{lscape} % Used to put a section in landscape mode
+\usepackage[table]{xcolor} % Color definitions
+\usepackage{algorithm} % Algorithm float
+\usepackage{algpseudocode} % Algorithmic environment
+
+%%%% Commands %%%%
 
@@
 \newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
+\algrenewcommand{\algorithmiccomment}[1]{\# \textit{#1}} % so that comments in algorithms are preprended by '#' instead of a small right arrow
 
@@
         Rafael Ferreira da Silva,John Flavin, Pascal Girard, Krzysztof J. Gorgolewski, \\
-        Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux, Gunnar Schaeffer,\\
-        Marc-\'Etienne Rousseau, Alan C. Evans}
+        Charles G. Guttmann, Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux,\\
+        Gunnar Schaeffer, Marc-\'Etienne Rousseau, Alan C. Evans}
 
@@
 
+At runtime, a value is assigned to all the mandatory and some of the
+optional inputs. Inputs of type "String" may contain any string,
+inputs of type "Number" must contain a string representing a number,
+inputs of type "File" must contain a string representing a file path
+(absolute or relative to the execution directory) and inputs of type
+"Flag" must contain a boolean. When input is a list, the value
+contains the concatenation of all strings in the list.
+
+Algorithm~\ref{algo:command-line} shows how the application command
+line is constructed from the application manifest and values entered
+by the user. It substitutes all the value keys in the command line,
+output path templates and configuration files, and writes the
+configuration files.
+
+\begin{algorithm}[htb]
+\caption{Command-line construction}
+\label{algo:command-line}
+\begin{algorithmic}
+  \State \Comment{Substitute input value keys in output path templates, configuration files and command line.}
+  \For{\texttt{input} in \texttt{inputs}}
+  \If{\texttt{input} has a \texttt{value-key}}
+  \For{\texttt{output} in \texttt{outputs}}
+  \State \texttt{stripped\_value} = \texttt{input\_value}
+  \If{\texttt{input} type is \texttt{File}}
+  \State In \texttt{stripped\_value}, remove  all elements in \texttt{path-template-stripped-extensions}.
+  \EndIf
+  \State In \texttt{path-template}, replace all occurrences of \texttt{value-key} by \texttt{stripped\_value}.
+  \If{\texttt{output} has a \texttt{file-template}}
+  \State In any line of \texttt{file-template}, replace all occurrences of \texttt{value-key} by \texttt{stripped\_value}.
+  \EndIf
+  \EndFor
+  \State Prepend \texttt{command-line-flag} and \texttt{command-line-flag-separator} to \texttt{input\_value}.
+  \State In \texttt{command-line}, replace all occurrences of \texttt{value-key} by \texttt{input\_value}.
+  \EndIf
+  \EndFor
+
+  \State \Comment{Substitute output value keys in configuration files and command line.}
+  \For{\texttt{output} in \texttt{outputs}}
+  \If{\texttt{output} has a \texttt{value-key}}
+  \For{\texttt{output} in \texttt{outputs}}
+  \If{\texttt{output} has a \texttt{file-template}}
+  \State In any line of \texttt{file-template}, replace all occurrences of \texttt{value-key} by \texttt{path-template}.
+  \State \Comment{Input \texttt{value-key} have been substituted in \texttt{path-template} previously.}
+  \EndIf
+  \EndFor
+  \State Prepend \texttt{command-line-flag} and \texttt{command-line-flag-separator} to \texttt{path-template}.
+  \State In \texttt{command-line}, replace all occurrences of \texttt{value-key} by \texttt{path-template}.
+  \EndIf
+  \EndFor
+
+  \State \Comment{Write all configuration files.}
+  \For{\texttt{output} in \texttt{outputs}}
+  \If{\texttt{output} has a \texttt{file-template}}
+  \State Write \texttt{file-template} in \texttt{path-template}
+  \State \Comment{Value keys have been substituted in \texttt{file-template} previously.}
+  \EndIf
+  \EndFor
+
+\end{algorithmic}
+\end{algorithm}
+
 \section{Implementation and Evaluation}
@@
 
-\subsection{Platforms}
-
-\paragraph{CBRAIN}
-
-New list mechanism.
-
-\paragraph{Pegasus}
-
-\paragraph{VIP}
-
 \subsection{Repository}
@@
 
-% PSOM
-% MICCAI Challenges 
-% exports from Nipype
 % SPM?
@@
   \end{landscape}
+
+\subsection{Platforms}
+
+\paragraph{CBRAIN}
+
+New list mechanism.
+
+\paragraph{Pegasus}
+
+\paragraph{SPINE}
+
+\paragraph{VIP}
 

### 2016-11-01 paper.tex
Added parallelization to the application table

@@
 \usepackage[margin=1in]{geometry} % Increases margins
-\usepackage{lscape} % Used to put a section in landscape mode
 \usepackage[table]{xcolor} % Color definitions
@@
 configuration files.
-
-\begin{algorithm}[htb]
+\begin{algorithm}[h!]
 \caption{Command-line construction}
@@
 with Boutiques and the main features used.
-\begin{landscape}
+
 \begin{table}
-  \rowcolors{2}{gray!25}{white}
-  \begin{tabular}{l|ccc|ccccccc|cc}
-    Application               &\multicolumn{3}{c|}{Containers} & \multicolumn{7}{c|}{Input/Output properties}                         & \multicolumn{2}{c}{Other properties}  \\
-                              &Docker              & Singularity & Rootfs  & Dependencies      &Groups             & Lists             & Optional          &Default            & Enum              & Min/max           & Config.
-file & Env vars     \\
+  \begin{tabular}{l|ccccccc}
+    \rowcolor[gray]{0.9}
+    Application               &\multicolumn{7}{c}{Input/Output properties} \\
+    \rowcolor[gray]{0.9}
+    & Dependencies      &Groups             & Lists             & Optional          &Default            & Enum              & Min/max\\
+
     \hline
-    Niak pre-processing       &\cellcolor{gray!75} &            &         &                   &\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&              &\cellcolor{gray!75}\\
-    ICA AROMA                 &\cellcolor{gray!75} &            &         &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&              &\\
-    FSL tools &&&&&&&&&&&&\\
-    \multicolumn{1}{r|}{anat} &                    &            &         &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&              &\\
-    \multicolumn{1}{r|}{bet}  &                    &            &         &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}&              &\\
-    \multicolumn{1}{r|}{fast} &                    &            &         &\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&              &\\
-    \multicolumn{1}{r|}{first}&                    &            &         &                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   &              &\\
+    Niak pre-processing       &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\\
+    ICA AROMA                 &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
+    FSL tools &&&&&&\\
+    \multicolumn{1}{r|}{anat} &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
+    \multicolumn{1}{r|}{bet}  &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}\\      
+    \multicolumn{1}{r|}{fast} &\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
+    \multicolumn{1}{r|}{first}&                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   \\
     FSL Nipype interfaces     \\
-    HCP PreFreesurfer         &\cellcolor{gray!75} &            &         &                   &                   &                   &                   &                   &                   &                   &              &\\
+    HCP PreFreesurfer         &                   &                   &                   &                   &                   &                   &                   \\
+    MSSEG MICCAI pipelines    \\
+    PETSEG MICCAI pipelines   \\
+    \end{tabular}
+  \vspace*{0.5cm}
+  \begin{tabular}{l|ccc|ccc}
+    \hline
+    \rowcolor[gray]{0.9}
+    Application               &\multicolumn{3}{c|}{Containers} &  \multicolumn{3}{c}{Other properties}  \\
+    \rowcolor[gray]{0.9}
+    &Docker              & Singularity & Rootfs  & Config. file & Env vars  & Parallelization   \\
+    \hline
+    Niak pre-processing       &\cellcolor{gray!75} &            &         &              &\cellcolor{gray!75}&\cellcolor{gray!75}\\
+    ICA AROMA                 &\cellcolor{gray!75} &            &         &              &&\\
+    FSL tools &&&&&\\
+    \multicolumn{1}{r|}{anat} &                    &            &         &              &&\\
+    \multicolumn{1}{r|}{bet}  &                    &            &         &              &&\\
+    \multicolumn{1}{r|}{fast} &                    &            &         &              &&\\
+    \multicolumn{1}{r|}{first}&                    &            &         &              &&\\
+    FSL Nipype interfaces     \\
+    HCP PreFreesurfer         &\cellcolor{gray!75} &            &         &             &&\\
     MSSEG MICCAI pipelines    \\
@@
 \end{table}
-  \end{landscape}
 

### 2016-11-01 paper.tex
Added intro to section 2

@@
 
-Define ``the platform'', i.e. the entity that will run applications.
-\begin{itemize}
-\item JSON manifest
-\item Link to container
-\item Invocation schema
-\end{itemize}
+The application is described with a \emph{manifest} that specifies the
+command-line template, inputs and outputs. The manifest may point to a
+container where the application and all its dependencies are
+installed. It may also contain an invocation schema used for input
+validation. At runtime, the execution \emph{platform} builds the
+application command line from the manifest and values entered by the
+user. The platform runs the command line on the execution
+infrastructure, e.g., a server, a cluster or a cloud, within a
+container whenever available. To build and run the command line, the
+platform may rely on the Boutiques \emph{core tools}, in particular on
+the Boutiques validator and local executor. Application manifests may
+be stored in the \emph{repository}, which provides additional features.
 
@@
 
-\section{Implementation and Evaluation}
-
 \subsection{Core tools}
@@
 \subsection{Repository}
+
+\section{Results}
 
@@
 
-\subsection{Platforms}
+\subsection{Supported platforms}
 

### 2016-11-01 paper.tex
More on ported applications

@@
 \newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
-\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
+\newcommand{\todo}[2]{\color{red}TODO: #1\color{black}}
 \algrenewcommand{\algorithmiccomment}[1]{\# \textit{#1}} % so that comments in algorithms are preprended by '#' instead of a small right arrow
@@
         Remi Bernard, Sorina Camarasu-Pop, Fr\'ed\'eric Cervenansky, Samir Das, \\
-        Rafael Ferreira da Silva,John Flavin, Pascal Girard, Krzysztof J. Gorgolewski, \\
+        Rafael Ferreira da Silva, John Flavin, Pascal Girard, Krzysztof J. Gorgolewski, \\
         Charles G. Guttmann, Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux,\\
@@
 
+\todo{Gunnar, Nathaniel, John}
+
 \section{Results}
-
-\subsection{Ported applications}
-
-% cbrain-plugins-neuro
-
-The Pipeline System for Octave and Matlab
-(PSOM)~\cite{bellec2012pipeline} has been integrated in CBRAIN using
-Boutiques. More precisely, a manifest has been written for a Niak
-pre-processing pipeline and PSOM workers. The integration uses the
-parallelization mechanism described in
-Section~\ref{sec:parallelization} so that even the tasks processing a
-given subject can be parallelized. As shown in~\cite{GLAT-16}, this
-allows CBRAIN to leverage the efficient agent model used in PSOM. The
-integration required some work in PSOM to facilitate its invocation as
-a non-interactive command-line application. The resulting CBRAIN
-plugin is available at
-\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It could in
-principle be used in other platforms that support the parallelization
-property, although this has not been tested.
-
-% SPM?
-% BIDS Apps
-
-Table~\ref{table:applications} summarizes the applications described
-with Boutiques and the main features used.
-
-\begin{table}
-  \begin{tabular}{l|ccccccc}
-    \rowcolor[gray]{0.9}
-    Application               &\multicolumn{7}{c}{Input/Output properties} \\
-    \rowcolor[gray]{0.9}
-    & Dependencies      &Groups             & Lists             & Optional          &Default            & Enum              & Min/max\\
-
-    \hline
-    Niak pre-processing       &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\\
-    ICA AROMA                 &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
-    FSL tools &&&&&&\\
-    \multicolumn{1}{r|}{anat} &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
-    \multicolumn{1}{r|}{bet}  &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}\\      
-    \multicolumn{1}{r|}{fast} &\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
-    \multicolumn{1}{r|}{first}&                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   \\
-    FSL Nipype interfaces     \\
-    HCP PreFreesurfer         &                   &                   &                   &                   &                   &                   &                   \\
-    MSSEG MICCAI pipelines    \\
-    PETSEG MICCAI pipelines   \\
-    \end{tabular}
-  \vspace*{0.5cm}
-  \begin{tabular}{l|ccc|ccc}
-    \hline
-    \rowcolor[gray]{0.9}
-    Application               &\multicolumn{3}{c|}{Containers} &  \multicolumn{3}{c}{Other properties}  \\
-    \rowcolor[gray]{0.9}
-    &Docker              & Singularity & Rootfs  & Config. file & Env vars  & Parallelization   \\
-    \hline
-    Niak pre-processing       &\cellcolor{gray!75} &            &         &              &\cellcolor{gray!75}&\cellcolor{gray!75}\\
-    ICA AROMA                 &\cellcolor{gray!75} &            &         &              &&\\
-    FSL tools &&&&&\\
-    \multicolumn{1}{r|}{anat} &                    &            &         &              &&\\
-    \multicolumn{1}{r|}{bet}  &                    &            &         &              &&\\
-    \multicolumn{1}{r|}{fast} &                    &            &         &              &&\\
-    \multicolumn{1}{r|}{first}&                    &            &         &              &&\\
-    FSL Nipype interfaces     \\
-    HCP PreFreesurfer         &\cellcolor{gray!75} &            &         &             &&\\
-    MSSEG MICCAI pipelines    \\
-    PETSEG MICCAI pipelines   \\
-  \end{tabular} 
-  \caption{Applications available in Boutiques with manifest features used.}
-  \label{table:applications}
-\end{table}
 
@@
 \paragraph{VIP}
+
+\subsection{Ported applications}
+
+% cbrain-plugins-neuro
+
+Table~\ref{table:applications} summarizes the applications described
+with Boutiques and the main features used.
+
+The Niak pre-processing pipeline, executed with the Pipeline System
+for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
+in CBRAIN using Boutiques. The integration uses the parallelization
+mechanism described in Section~\ref{sec:parallelization} so that even
+the tasks processing a single subject can be parallelized. It also
+allows CBRAIN to leverage the efficient agent model used in PSOM, as
+described in~\cite{GLAT-16}. The integration required some work in
+PSOM to facilitate its invocation as a non-interactive command-line
+application. The resulting CBRAIN plugin is available at
+\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It could in
+principle be used in other platforms that support the parallelization
+property, although this has not been tested.
+
+Boutiques was also used to port to the VIP platform the 23 pipelines
+involved in two computational challenges organized by the MICCAI
+conference in 2016, related to the segmentation of multiple-sclerosis
+lesions in MR images and of tumor volumes in PET images. The pipeline
+developers created Boutiques manifests and wrapped their applications
+in Docker containers. The pipelines were ported to VIP and executed on
+205 subjects in a few weeks only. Some pipelines had to be adjusted
+once ported to the platform for the following reasons:
+\begin{enumerate}
+\item A pipeline
+  required a GPU, which we enabled through the
+  \texttt{nvidia-docker}\footnote{\url{https://github.com/NVIDIA/nvidia-docker}}
+  tool not supported in Boutiques by default although it could be a possible extension.
+\item A pipeline required
+  more than 10 GB of data dependencies (atlas data) which exceeded the
+  maximum size allowed for Docker containers in our setup; we solved the
+  issue by installing the data on a directory of the host server that we
+  mounted in the container. This could not possibly be supported by
+  Boutiques since it is very dependent on the platform.
+\item A pipeline wrote more than 10 GB of intermediate data in a
+  temporary directory of the container located on a 2 GB partition. We
+  solved the issue by mounting a host directory in the temporary directory.
+\end{enumerate}
+The pipelines did not use much of the expressiveness features offered
+by Boutiques since the instructions given to the challengers required
+that all the pipeline parameters were fixed, to facilitate execution.
+
+The Nipype workflow engine~\cite{gorgolewski2011nipype} was
+instrumented to export to Boutiques applications described in its
+internal Python description language. \todo{Update nipype\_cmd to
+  support the latest Boutiques schema and complete this paragraph.}
+
+Other neuroinformatics tools were ported to CBRAIN using Boutiques:
+FSL tools and ICA AROMA are available at
+\url{https://github.com/aces/cbrain-plugins-neuro} and some
+pipelines of the Human Connectome Project (HCP) are at
+\url{https://github.com/big-data-lab-team/cbrain-plugins-hcp}.
+
+\todo{SPM}
+
+\todo{Chris, something on BIDS apps?}
+
+\todo{Sorina, a paragraph on other applications in VIP?}
+
+\begin{table}
+  \begin{tabular}{l|ccccccc}
+    \rowcolor[gray]{0.9}
+    Application                      & \multicolumn{7}{c}{Input/Output properties} \\
+    \rowcolor[gray]{0.9}
+                                     & Dependencies      &Groups             & Lists             & Optional          &Default            & Enum              & Min/max\\
+
+    \hline
+    Niak pre-processing              &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\\
+    ICA AROMA                        &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
+    FSL tools &&&&&&\\
+    \multicolumn{1}{r|}{anat}        &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
+    \multicolumn{1}{r|}{bet}         &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}\\      
+    \multicolumn{1}{r|}{fast}        &\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
+    \multicolumn{1}{r|}{first}       &                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   \\
+    FSL Nipype interfaces     \\
+    HCP PreFreesurfer                &                   &                   &                   &                   &                   &                   &                   \\
+    MICCAI challenge (23 pipelines)\\&                   &                   &                   &                   &                   &                   & \\
+    \end{tabular}
+  \vspace*{0.5cm}
+  \begin{tabular}{l|ccc|ccc}
+    \hline
+    \rowcolor[gray]{0.9}
+    Application                     &\multicolumn{3}{c|}{Containers}            &  \multicolumn{3}{c}{Other properties}  \\
+    \rowcolor[gray]{0.9}
+                                    &Docker              & Singularity& Rootfs  & Config. file & Env vars  & Parallelization   \\
+    \hline
+    Niak pre-processing             &\cellcolor{gray!75} &            &         &              &\cellcolor{gray!75}&\cellcolor{gray!75}\\
+    ICA AROMA                       &\cellcolor{gray!75} &            &         &              &&\\
+    FSL tools &&&&&\\
+    \multicolumn{1}{r|}{anat}       &                    &            &         &              &&\\
+    \multicolumn{1}{r|}{bet}        &                    &            &         &              &&\\
+    \multicolumn{1}{r|}{fast}       &                    &            &         &              &&\\
+    \multicolumn{1}{r|}{first}      &                    &            &         &              &&\\
+    FSL Nipype interfaces     \\
+    HCP PreFreesurfer               &\cellcolor{gray!75} &            &         &             &&\\
+    MICCAI challenge (23 pipelines) &\cellcolor{gray!75} &            &         &             &&\\
+  \end{tabular} 
+  \caption{Applications available in Boutiques with manifest features used.}
+  \label{table:applications}
+\end{table}
+
+\section{Discussion}
+
+\begin{itemize}
+\item Security: containers may be updated after being integrated in the platforms.
+\item The manifest specification is getting increasingly complex, but
+  this complexity is required to describe real-world
+  applications. When something cannot be described using the manifest,
+  it is always possible to change the command line format by writing a
+  wrapper script for the application. However, when containers are
+  used, the wrapper script needs to be added to the container, which
+  has some overhead (in particular, the containers provided by the
+  developers cannot be used directly).
+\item Testing is critical.
+  \item Workflow engines can be platforms (they can run Boutiques tools), workflows can be applications (they can be described in Boutiques). 
+\end{itemize}
 
@@
 
-\todo{Tristan}{Update from comparison summary on Google doc}
-
-\section{Discussion}
+\todo{Update from comparison summary on Google doc}
 

### 2016-11-01 paper.tex
Added text on CBRAIN

@@
 features but they should be used with care to avoid making tools
-dependent on a particular platform. The following custom properties
-have been used so far:
-\begin{itemize}
-\item \texttt{cbrain:inherits-from-class}: a string that defines the
-  Ruby class that should be used as parent class for the tool in
-  CBRAIN. Used to define a progress bar for PSOM tools in CBRAIN.
-\item \texttt{vip:miccai-challenger-email} and
-  \texttt{vip:miccai-challenge-team-name}: strings helping VIP
-  categorize tools for the 2016 MICCAI challenges ``MSSEG'' and
-  ``PETSEG''.
-\end{itemize}
+dependent on a particular platform.
 
@@
 
-\paragraph{CBRAIN}
+The following platforms are able to integrate and execute Boutiques
+applications.
 
-New list mechanism.
+\subsubsection{CBRAIN}
+
+CBRAIN~\cite{SHER-14} is a web platform that can process data
+distributed into multiple storage locations on computing clusters and
+clouds. CBRAIN offers transparent access to remote data sources,
+distributed computing sites, and an array of processing and
+visualization tools within a controlled, secure environment.  The
+CBRAIN service deployed at the Montreal Neurological Institute relies
+on the infrastructure provided by Compute Canada~\cite{das2016mni}. It
+currently provides 460+ collaborators in 20 countries with web access
+to several systems, including 6 clusters of the Compute Canada
+high-performance computing infrastructure (totalling more than 100,000
+computing cores and 40PB of disk storage) and Amazon EC2. CBRAIN
+transiently stores about 10 million files representing over 50TB
+distributed in 42 servers. 56 data processing tools are integrated and
+over 340,000 processing batches have been submitted since 2010.
+
+Applications in CBRAIN are integrated as Ruby classes that create web
+forms, validation and execution code. The Boutiques support in CBRAIN
+is based on a set of templates that generate such classes from the
+application manifest. Two integration modes are available to integrate
+Boutiques applications: (1) the application manifest is stored in a
+CBRAIN plugin, and the Ruby classes are generated on-the-fly when
+CBRAIN starts. This mode allows CBRAIN developers to add new features
+to all Boutiques applications (e.g. improved error message reporting)
+by updating the Ruby templates. It is, however, constrained to the
+Boutiques manifest schema: features not supported in Boutiques, e.g.,
+cluster-specific adjustments, cannot be supported. (2) CBRAIN Ruby
+classes are generated from the Boutiques manifest in an offline
+process and integrated in CBRAIN as any other application. This mode
+allows developers to customize applications by editing the generated
+Ruby classes, but the resulting applications are difficult to maintain
+in the long term, in particular when the manifests are updated.
+
+A custom property (\texttt{cbrain:inherits-from-class}) was added to
+the Boutiques manifest to define the Ruby class that should be used as
+parent class for the application in CBRAIN. It provides more
+flexibility to the first integration model described previously. For
+instance, it was used to define a progress bar for PSOM pipelines in
+CBRAIN.
+
+CBRAIN was also extended to support the parallelization mechanism
+available in Boutiques: the CBRAIN execution agent periodically scans
+the execution directories of the parallelized applications and it submits
+the invocations defined in the JSON objects written by the
+application in this directory.
+
+A new list mechanism was also introduced in CBRAIN to facilitate the
+iteration of Boutiques applications on large sets of files. CBRAIN
+lists are specific files that contain references to other CBRAIN
+files. When a list is passed to a Boutiques application, the elements
+in the list are either concatenated in a single command line (when the
+corresponding Boutiques input is a list), or a new command line is
+generated for every element in the list (when the input is not a
+list). Supporting lists as a specific CBRAIN file type allows for
+improved validation, e.g. lists that contain references to
+non-existent or deleted files can be detected. It also allows users to
+edit lists using their own tools such as scripts or spreadsheet
+applications.
 
@@
 
+\todo{Rafael}
+
 \paragraph{SPINE}
+
+\todo{Charles}
 
@@
 by Boutiques since the instructions given to the challengers required
-that all the pipeline parameters were fixed, to facilitate execution.
+that all the pipeline parameters were fixed, to facilitate
+execution. Two custom properties (\texttt{vip:miccai-challenger-email}
+and \texttt{vip:miccai-challenge-team-name}) were added to the
+Boutiques manifest to help VIP handle pipeline results for the 2016
+MICCAI challenges.
 
@@
 \item Workflow engines can be platforms (they can run Boutiques tools), workflows can be applications (they can be described in Boutiques).
+\item With Boutiques, developers can port their application once and execute it in several platforms. It removes the technological dependency to a particular platform and facilitates migration. 
 \end{itemize}

### 2016-11-03 paper.tex
Started SPM text

@@
         Remi Bernard, Sorina Camarasu-Pop, Fr\'ed\'eric Cervenansky, Samir Das, \\
-        Rafael Ferreira da Silva, John Flavin, Pascal Girard, Krzysztof J. Gorgolewski, \\
-        Charles G. Guttmann, Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux,\\
-        Gunnar Schaeffer, Marc-\'Etienne Rousseau, Alan C. Evans}
+        Rafael Ferreira da Silva, Guillaume Flandin, John Flavin, Pascal Girard, \\
+         Krzysztof J. Gorgolewski, Charles G. Guttmann, Nathaniel Kofalt, Pierre-Olivier Quirion,\\
+         Pierre Rioux, Gunnar Schaeffer, Marc-\'Etienne Rousseau, Alan C. Evans}
 
@@
 
-The Niak pre-processing pipeline, executed with the Pipeline System
+\paragraph{Niak.} The Niak pre-processing pipeline, executed with the Pipeline System
 for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
@@
 
-Boutiques was also used to port to the VIP platform the 23 pipelines
+\paragraph{Nipype.} The Nipype workflow engine~\cite{gorgolewski2011nipype} was
+instrumented to export to Boutiques applications described in its
+internal Python description language. \todo{Update nipype\_cmd to
+  support the latest Boutiques schema and complete this paragraph.}
+
+\paragraph{SPM.} The Statistical Parameter Mapping toolbox (SPM)~\cite{penny2011statistical}
+was exported to Boutiques through a specific \texttt{spm2boutiques}
+tool that writes Boutiques manifests from configuration files of SPM
+functions (object declarations). \todo{More on that when it is
+  completed.}
+
+\paragraph{MICCAI challenges.} Boutiques was also used to port to the VIP platform the 23 pipelines
 involved in two computational challenges organized by the MICCAI
@@
 
-The Nipype workflow engine~\cite{gorgolewski2011nipype} was
-instrumented to export to Boutiques applications described in its
-internal Python description language. \todo{Update nipype\_cmd to
-  support the latest Boutiques schema and complete this paragraph.}
-
-Other neuroinformatics tools were ported to CBRAIN using Boutiques:
+\paragraph{Other tools.} Other neuroinformatics tools were ported to CBRAIN using Boutiques:
 FSL tools and ICA AROMA are available at
-\url{https://github.com/aces/cbrain-plugins-neuro} and some
-pipelines of the Human Connectome Project (HCP) are at
+\url{https://github.com/aces/cbrain-plugins-neuro} and some pipelines
+of the Human Connectome Project (HCP) are at
 \url{https://github.com/big-data-lab-team/cbrain-plugins-hcp}.
-
-\todo{SPM}
 

### 2016-11-03 paper.tex
First introduction draft

@@
 \usepackage{algpseudocode} % Algorithmic environment
+\usepackage{xspace}
 
@@
 \newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
-\newcommand{\todo}[2]{\color{red}TODO: #1\color{black}}
+\newcommand{\todo}[1]{\color{red}TODO: #1\color{black}}
+\newcommand{\boutiques}{Boutiques\xspace}
 \algrenewcommand{\algorithmiccomment}[1]{\# \textit{#1}} % so that comments in algorithms are preprended by '#' instead of a small right arrow
@@
 
+\abstract{Porting applications to execution platforms is a critical
+  step to enable their sharing among scientific communities and their
+  efficient execution on computing infrastructures. However,
+  application porting remains a costly human effort that consists of
+  1)~installing the application on the target infrastructure,
+  2)~describing the application in a format compatible with the
+  execution platform, and 3)~generating proper user interfaces. Due to
+  the variety of platforms, application porting efforts are often
+  replicated several times while mutualization would save cost and
+  improve the quality of the ported applications. We describe
+  \emph{boutiques}, a system that allows for automatic import and
+  exchange of applications across platforms. Boutiques relies on Linux
+  containers to solve the problem of application installation in a
+  lightweight manner. It also adopts a rich and flexible application
+  description format which is versatile enough to be used in various
+  platforms. Boutiques is currently supported by \todo{X} platforms
+  and was used to describe \todo{Y} applications.
+}
+
 \section{Introduction}
 
-Why Linux. Define application. Define computing platforms. Explain the
-problem and why it is difficult. Approach: start simple and enrich the
-spec when new applications come. Refer to computing infrastructures
-and the need for parallelization. Who are
-the developers. Refer to neuroinformatics field.
+Platforms such as web services, portals, virtual research
+environments, science gateways and workflow engines often rely on
+third-party data processing and simulation applications to deliver
+services to their users. Such applications are often repeatedly and
+manually ported to various platforms whereas this porting effort would
+greatly benefit from being mutualized. Virtual container systems such
+as Docker and Singularity greatly facilitate the sharing of
+applications executed in Linux systems, and improve the
+reproducibility of analyses by defining immutable, reusable execution
+environments. However, containers need to be properly annotated to
+allow platforms to leverage the applications that they
+contain. Basically, the command lines that can be run in a container
+must be formally described so that they can be automatically created
+and executed by platforms. Describing command lines is not
+straightforward since they often contain implicit semantics specified
+in human-readable documents such as manuals.
 
-Vocabulary (check consistency across the paper):
-\begin{itemize}
-\item tool vs. application.
-\item descriptor vs. manifest.
-\item task vs. invocation.
-\end{itemize}
+In Section~\ref{sec:system}, we describe \boutiques, a system to share
+command-line applications across platforms. In \boutiques, a command
+line is described from a flexible template referring to the inputs it
+requires and the outputs it produces. Inputs may be inter-dependent,
+e.g., input A is available only if input B was defined and passed in
+configuration files or directly on the command line. Such formal
+descriptions, also called \emph{manifests}, allows for advanced
+validation of input values, which is useful error detection. Manifests
+refer to containers where applications and all their dependencies are
+installed. \boutiques also provides a simple parallelization framework
+that allows applications such as workflow engines to execute multiple
+tasks concurrently. A set of core tools facilitates the construction,
+validation and execution of \boutiques manifests. A repository
+orchestrates these tools together.
+
+\boutiques was created using a bottom-up approach in which new
+features were progressively added as applications and platforms
+required. It was used to describe \todo{Y} applications and it is
+supported by \todo{X} platforms, which we report in
+Section~\ref{sec:results}.
+
+\boutiques manifests are intented to be produced by scientific
+application developers, to be published in common repositories, and to
+be consumed by execution platforms. Although the motivating use cases
+were taken from neuroinformatics, our primary field of interest,
+nothing prevents the system from being used in a variety of other
+fields.
+
+\note{Tristan}{Vocabulary (check consistency across the paper): tool vs. application; descriptor vs. manifest; task vs. invocation}
+\note{Tristan}{Targeted journals: Scientific Data, Giga Science, Frontiers in Neuroinformatics, Future Generation Computer Systems}
 
 \section{System description}
+\label{sec:system}
 
@@
 
+\todo{json editor}
+
 \subsection{Repository}
@@
 \section{Results}
+\label{sec:results}
 
@@
 
-\paragraph{Pegasus}
+\subsection{Flywheel}
+
+\todo{Gunnar, Nathanael, John}
+
+\subsection{Pegasus}
 
@@
 
-\paragraph{SPINE}
+\subsection{SPINE}
 
@@
 
-\paragraph{VIP}
+\subsection{VIP}
 

### 2016-11-10 paper.tex
Proof-read sections I and II

@@
 \newcommand{\boutiques}{Boutiques\xspace}
+\newcommand{\notimplementedyet}[1]{\color{blue}\emph{#1}\footnote{Still needs to be implemented}\color{black}\xspace}
 \algrenewcommand{\algorithmiccomment}[1]{\# \textit{#1}} % so that comments in algorithms are preprended by '#' instead of a small right arrow
@@
 
-\abstract{Porting applications to execution platforms is a critical
-  step to enable their sharing among scientific communities and their
-  efficient execution on computing infrastructures. However,
-  application porting remains a costly human effort that consists of
-  1)~installing the application on the target infrastructure,
-  2)~describing the application in a format compatible with the
-  execution platform, and 3)~generating proper user interfaces. Due to
-  the variety of platforms, application porting efforts are often
-  replicated several times while mutualization would save cost and
-  improve the quality of the ported applications. We describe
-  \emph{boutiques}, a system that allows for automatic import and
-  exchange of applications across platforms. Boutiques relies on Linux
-  containers to solve the problem of application installation in a
-  lightweight manner. It also adopts a rich and flexible application
-  description format which is versatile enough to be used in various
-  platforms. Boutiques is currently supported by \todo{X} platforms
-  and was used to describe \todo{Y} applications.
-}
+\abstract{Porting applications to execution platforms such as web
+  portals and workflow engines is critical to enable their sharing
+  among scientific communities and their efficient execution on
+  computing infrastructures. However, application porting remains a
+  costly human effort that consists of 1)~installing the application
+  on the target infrastructure, 2)~describing the application in a
+  format compatible with the execution platform, and 3)~generating
+  proper user interfaces. Due to the variety of platforms, application
+  porting efforts are often replicated several times while
+  mutualization would save cost and improve the quality of the ported
+  applications. We describe \boutiques, a system that allows for
+  automatic import and exchange of applications across
+  platforms. \boutiques adopts a rich, flexible application
+  description format and it relies on Linux containers to install
+  applications across platforms in a lightweight manner. It is
+  currently supported by \todo{X} platforms and was used to describe
+  \todo{Y} applications.  }
 
@@
 manually ported to various platforms whereas this porting effort would
-greatly benefit from being mutualized. Virtual container systems such
-as Docker and Singularity greatly facilitate the sharing of
-applications executed in Linux systems, and improve the
-reproducibility of analyses by defining immutable, reusable execution
-environments. However, containers need to be properly annotated to
-allow platforms to leverage the applications that they
-contain. Basically, the command lines that can be run in a container
-must be formally described so that they can be automatically created
-and executed by platforms. Describing command lines is not
+greatly benefit from being mutualized. Meanwhile, virtual container
+systems such as Docker and Singularity greatly facilitate the sharing
+of software and improve the reproducibility of analyses by defining
+immutable, reusable execution environments. However, containers need
+to be properly annotated to allow platforms to leverage the software
+they contain. Basically, the command lines that can be run in a
+container must be formally described so that they can be automatically
+created and executed by platforms. Describing command lines is not
 straightforward since they often contain implicit semantics specified
@@
 
-In Section~\ref{sec:system}, we describe \boutiques, a system to share
-command-line applications across platforms. In \boutiques, a command
-line is described from a flexible template referring to the inputs it
-requires and the outputs it produces. Inputs may be inter-dependent,
-e.g., input A is available only if input B was defined and passed in
-configuration files or directly on the command line. Such formal
-descriptions, also called \emph{manifests}, allows for advanced
-validation of input values, which is useful error detection. Manifests
-refer to containers where applications and all their dependencies are
-installed. \boutiques also provides a simple parallelization framework
-that allows applications such as workflow engines to execute multiple
-tasks concurrently. A set of core tools facilitates the construction,
-validation and execution of \boutiques manifests. A repository
-orchestrates these tools together.
+\boutiques is a system to share command-line applications across
+platforms. In \boutiques, a command line is described using a flexible
+template comprising the inputs it requires and the outputs it
+produces. Inputs may be inter-dependent, e.g., input A cannot be used
+if input B is used, and passed in configuration files or directly on
+the command line. Such formal descriptions, also called
+\emph{manifests}, allow for advanced validation of input values to
+prevent errors. Manifests refer to containers where applications and
+all their dependencies are installed. \boutiques also provides a
+simple parallelization framework that allows applications such as
+workflow engines to concurrently execute multiple tasks (in the
+remainder, tasks are called \emph{invocations}). A set of core tools
+facilitate the construction, validation and execution of \boutiques
+manifests. A repository orchestrates these tools together. \boutiques
+manifests are intented to be produced by scientific application
+developers, to be published in common repositories and to be consumed
+by execution platforms.
 
-\boutiques was created using a bottom-up approach in which new
-features were progressively added as applications and platforms
-required. It was used to describe \todo{Y} applications and it is
-supported by \todo{X} platforms, which we report in
-Section~\ref{sec:results}.
+The remainder of this paper describes the \boutiques system
+(Section~\ref{sec:system}) and reports on its adoption by platforms
+and applications in the neuroinformatics domain, our primary field of
+interest (Section~\ref{sec:results}). It closes on a discussion and
+comparison with related systems.
 
-\boutiques manifests are intented to be produced by scientific
-application developers, to be published in common repositories, and to
-be consumed by execution platforms. Although the motivating use cases
-were taken from neuroinformatics, our primary field of interest,
-nothing prevents the system from being used in a variety of other
-fields.
-
-\note{Tristan}{Vocabulary (check consistency across the paper): tool vs. application; descriptor vs. manifest; task vs. invocation}
+\note{Tristan}{Vocabulary (check consistency across the paper): tool vs. application}
 \note{Tristan}{Targeted journals: Scientific Data, Giga Science, Frontiers in Neuroinformatics, Future Generation Computer Systems}
@@
 
-The application is described with a \emph{manifest} that specifies the
+Applications are described with a JSON manifest that specifies the
 command-line template, inputs and outputs. The manifest may point to a
@@
 installed. It may also contain an invocation schema used for input
-validation. At runtime, the execution \emph{platform} builds the
-application command line from the manifest and values entered by the
-user. The platform runs the command line on the execution
-infrastructure, e.g., a server, a cluster or a cloud, within a
-container whenever available. To build and run the command line, the
-platform may rely on the Boutiques \emph{core tools}, in particular on
-the Boutiques validator and local executor. Application manifests may
-be stored in the \emph{repository}, which provides additional features.
+validation. At runtime, the execution platform builds the command line
+from the manifest and values entered by the user. The platform runs
+the command line on the execution infrastructure, e.g., a server, a
+cluster or a cloud, within a container whenever available. To build
+and run the command line, the platform may rely on the \boutiques core
+tools, in particular on the validator and local executor. Application
+manifests may be stored in the repository which provides additional
+features.
 
@@
 
-The core component of the JSON manifest is a flexible command-line
+The core component of the manifest is a flexible command-line
 description represented as a simple string template complying to the
-syntax of the \texttt{bash} Linux shell interpreter. The command line
+syntax of the \texttt{bash} UNIX shell. \texttt{bash} is the default
+shell on the major Linux distributions and OS X. The command-line
 template may contain placeholders for input and output values, called
-value keys. The value keys may be any string, and the
-command line may actually encompass several commands, e.g., separated
-by semicolumns, pipes (\texttt{|}) or ampersands (\texttt{\&}). Such a
-flexible command-line specification is meant to facilitate the
-embedding of minor operations on the command line, for instance input
-decompression and output archival, avoiding the use of multiple
-``shims'' to align mismatching applications in a
-pipeline~\cite{hull2004treating}.
+value keys. The command line may actually
+encompass several commands separated, e.g., by semicolumns, pipes
+(\texttt{|}) or ampersands (\texttt{\&}). Such a flexible command-line
+specification is meant to facilitate the embedding of minor mundane
+operations on the command line, for instance directory creation, input
+decompression or output archival.
 
@@
 \end{verbatim}
-The template contains six value keys for input and output
-values. When the application is executed, such keys will be replaced
-by values and file names according to the user input. Flags will also
-be added wherever appropriate, with customizable separators. No
-particular syntax is imposed for the value keys, they just have
-to be unique.  Note the use of the pipe (\texttt{|}) operator to chain
-tools, and of the \texttt{>>} operator to redirect the standard output
-to a file.
+The template contains six value keys for input and output values, that
+will be replaced by values and file names according to the user input
+when the application is executed. Flags will also be added wherever
+appropriate, with customizable separators. No particular syntax is
+imposed for the value keys, they just have to be unique.  Note the use
+of the pipe (\texttt{|}) operator to chain applications, and of the
+\texttt{>>} operator to redirect the standard output to a file.
 
@@
 identifier and a type. They may be optional, have a description, a
-value key, a flag and flag separator, and a default
-value. Inputs may also be ordered lists; in this case, value
-keys are substituted by the space-separated list of input values.
+value key, a flag and flag separator, and a default value. Inputs may
+also be ordered lists: in this case, value keys are substituted by the
+space-separated list of input values.
 
@@
 \texttt{Number}, \texttt{Flag} or \texttt{File}. \texttt{File} may
-also represent a directory. Types can be restricted to a specific set
+also represents a directory. Types can be restricted to a specific set
 of values (\texttt{String} and \texttt{Number} only), to a specific
@@
 \paragraph{Groups and dependencies.} Groups of inputs may be defined
-from a identifier, name and list of input identifiers. Groups may be
+with a identifier, name and list of input identifiers. Groups may be
 used to improve the presentation in a graphical user interface and to
@@
 regardless of a particular group: an input may (1) require a list of
-inputs (2) disable a list of inputs.
+inputs and (2) disable a list of inputs.
 
@@
 
-Application outputs are the files or directories that need to be
+Application outputs are files and directories that need to be
 transferred and delivered to the user once the execution is
@@
 outputs to build the application command line, this information is
-needed by computing platforms to identify the files that must be
-saved after the execution.
+needed by computing platforms to identify the files that must be saved
+after the execution.
 
-In Boutiques, output files are defined from a unique identifier, a
-name and a path template that specifies the file or directory
-name. Path templates may include input value keys in case
-output files are named after the input values. Input values may be
-stripped from specific strings, e.g., file extensions before being
-substituted in the path template. Output files may also have a
-description, a command-line flag, a flag separator, and a value
-key in case they appear on the command line. They may be optional in
-case the file is not always produced by the application, for instance
-when it is produced only when a particular flag is activated. They
-may also be (un-ordered) lists; in this case, the path template must
-contain a wildcard (\texttt{*}) matching any string of characters
-and defining the pattern used to match the
-output files in the list.
+In \boutiques, output files must have a unique identifier, a name and
+a path template that specifies the file or directory name. Path
+templates may include input value keys in case output files are named
+after the input values. In this case, input values may be stripped
+from specific strings, e.g., file extensions before being substituted
+in the path template. Output files may also have a description, a
+command-line flag, a flag separator, and a value key in case they
+appear on the command line. They may be optional in case the file is
+not always produced by the application, for instance when it is
+produced only when a particular flag is activated. They may also be
+(un-ordered) lists: in this case, the path template must contain a
+wildcard (\texttt{*}) matching any string of characters and defining
+the pattern used to match the output files in the list.
 
@@
     "name" : "Log file",
-    "description" : "The output log file from the example tool",
+    "description" : "The output log file from the example application",
     "path-template" : "log-[STRING_INPUT]",
@@
 language.  As it does not seem reasonable to adopt a particular
-template for configuration files, the Boutiques specification allows
-application developers to specify their own template, referring to
-input and output value keys. Configuration files are specific types of
-output files that have a
- file template defined. As output files, 
-they must have a path template that defines
-how they must be named and where they must be written. They may also
-have a value key and a flag in case they need to be passed on the
-command line.
-
-Listing~\ref{listing:configuration-file-example} shows an example.
+template for configuration files, the \boutiques specification allows
+application developers to specify their own template containing input
+and output value keys. Configuration files are specific types of
+output files that must have a file template in addition to a path
+template that defines how they will be named and where they will be
+written. They may also have a value key and a flag in case they need
+to be passed on the command
+line. Listing~\ref{listing:configuration-file-example} shows an
+example.
 \begin{listing}
@@
 
-\subsection{Invocation schema}
-\label{sec:invocation-schema}
-
-Rigorous input validation is an important motivation for
-Boutiques. The obvious solution in our context is to rely on an
-application-specific JSON schema, called ``invocation schema'', to
-specify the correct input values that an application
-accepts. Platforms could rely on the invocation schema to validate
-inputs automatically using any JSON validator, without having to
-develop specific code.
-
-Invocation schemas, however, are complex JSON objects. Basically, they
-must represent the properties described above in a formal way,
-including dependencies between
-inputs. Listing~\ref{listing:invocation-schema-example} shows an
-example of how dependencies between mutually exclusive parameters are
-defined in the invocation schema.
-
-The invocation schema also has a \texttt{task-dependencies} array that
-specifies that the new invocation to start only when all the
-invocations in this array have completed. This property is used in the
-parallelization model described hereafter.
-
-\begin{listing}
-\begin{minted}[frame=single,
-               framesep=3mm,
-               linenos=false,
-               xleftmargin=21pt,
-               tabsize=4]{js}
-"dependencies" : {
-      "num_input" : {
-         "properties" : {
-            "str_input" : {
-               "not" : {}
-            }
-         }
-      },
-      "str_input" : {
-         "properties" : {
-            "num_input" : {
-               "not" : {}
-            }
-         }
-      }
-   }
-\end{minted}
-\caption{Excerpt from invocation schema showing dependencies between
-  two mutually exclusive parameters \texttt{num\_input} and
-  \texttt{str\_input}.}
-\label{listing:invocation-schema-example}
-\end{listing}
-
-It does not seem realistic to assume that developers will easily write
-correct JSON schemas for their applications. Instead, invocation
-schemas should be generated automatically using a core Boutiques
-tool. Accordingly, invocation schema is an optional property of the
-Boutiques manifest, that platforms or developers may generate using
-the appropriate tool.
-
-\subsection{Parallelization support}
-\label{sec:parallelization}
-Boutiques support parallelism by allowing applications to submit new
-tasks through the following protocol:
-\begin{enumerate}
-\item Applications write a JSON object complying to the invocation
-  schema in the directory where there were launched.
-\item The platform reads the JSON object, creates the corresponding
-  task, and writes back the task id in the same directory.
-% The platform could also support status requests.
-\end{enumerate}
-Several invocations can be submitted at once, possibly using the
-\texttt{task-dependencies} array to specify dependencies among
-them. This mechanism is implemented in the Boutiques schema using a
-single property, \texttt{can-submit-tasks}.
-
-Although very simple, this model allows wrapping complete
-workflows. Workflows are wrapped as any other application, and they
-can be executed using any engine such as Nipype, PSOM and others,
-provided that such engines are installed in the container image
-referred in the application manifest.
-
-The adopted ``task encapsulation'' model is very simple but it
-allows for a scalable and reliable execution of workflows expressed in
-a variety of languages, as detailed in ~\cite{glatard2016fgcs}.
-
-\subsection{Containers}
-
-The implementation of the application is defined in a container image
-in the Docker, Singularity or rootfs format. We intentionally support
-multiple container formats as we anticipate that they will be used for
-different purposes. For instance, Docker is well suited for
-application developers and users who want to use applications on their
-local workstation. It is very well documented, maintained and it has a
-rich ecosystem of tools to help build and run containers on most
-platforms, including Windows, MacOS and Linux. Singularity is more
-suited for users and platforms that need to run applications on shared
-computing clusters. Bridges exist among these containers formats to
-convert container images across frameworks. For instance, a platform
-dedicated to high-performance computing may accept manifests referring
-to Docker containers to facilitate application integration by
-developers, and internally convert container images to Singularity to
-run applications efficiently on clusters.
-
-Container images are defined from their URL (Singularity and rootfs)
-or name in a Docker index. Container images may specify a working
-directory where the application has to be run. A hash may also be
-reported to accurately identify container images and detect updates.
-
-Containers are adopted because they allow for an automated and
-lightweight integration of application implementations in
-platforms. They are extremely useful to improve the reproducibility of
-analyses as variations in the software environment may have an
-important impact on the computed results. They also have limitations,
-in particular they do not specify the hardware architecture required
-to execute an application, which can be an issue in some cases.
-
-Containers are not mandatory though. If an application manifest does
-not contain any reference to a container, the platform will assume
-that the application is pre-installed on the execution infrastructure.
-
-\subsection{Custom properties}
-
-Custom properties may be added to the Boutiques specification without
-restriction. They may be useful to implement platform-specific
-features but they should be used with care to avoid making tools
-dependent on a particular platform.
-
 \subsection{Command-line construction}
@@
 At runtime, a value is assigned to all the mandatory and some of the
-optional inputs. Inputs of type "String" may contain any string,
-inputs of type "Number" must contain a string representing a number,
-inputs of type "File" must contain a string representing a file path
-(absolute or relative to the execution directory) and inputs of type
-"Flag" must contain a boolean. When input is a list, the value
-contains the concatenation of all strings in the list.
-
-Algorithm~\ref{algo:command-line} shows how the application command
-line is constructed from the application manifest and values entered
-by the user. It substitutes all the value keys in the command line,
-output path templates and configuration files, and writes the
-configuration files.
+optional inputs. %% Inputs of type ``String'' may contain any string,
+%% inputs of type "Number" must contain a string representing a number,
+%% inputs of type "File" must contain a string representing a file path
+%% (absolute or relative to the execution directory) and inputs of type
+%% "Flag" must contain a boolean. 
+%% When input is a list, the value contains the concatenation of all the
+%%values in the list.
+Algorithm~\ref{algo:command-line} shows how the command line is
+constructed from the manifest and the values entered by the user. It
+substitutes all the value keys in the command line, output path
+templates and configuration files, and writes the configuration files.
 \begin{algorithm}[h!]
@@
 
+\subsection{Invocation schema}
+\label{sec:invocation-schema}
+
+Rigorous input validation is an important motivation for
+\boutiques. The obvious solution for JSON documents is to rely on an
+application-specific JSON schema, called \emph{invocation schema}, to
+specify the input values accepted by an application. Platforms can
+rely on invocation schemas to validate inputs using
+any JSON validator, without having to develop specific code.
+
+Invocation schemas, however, are complex JSON objects. Basically, they
+must represent the properties described above in a formal way,
+including dependencies between
+inputs. Listing~\ref{listing:invocation-schema-example} shows an
+example of how dependencies between mutually exclusive parameters are
+defined in the invocation schema. As it does not seem realistic to
+assume that developers will easily write correct invocations schemas
+even for their own applications, invocation schemas can be generated
+automatically using a core \boutiques tool. The invocation schema is
+stored as an optional property of the Boutiques manifest.
+
+\notimplementedyet{The invocation schema also has a
+  \texttt{task-dependencies} array that allows specifying that the new
+  invocation may start only when previously-submitted invocations have
+  completed. This property is used in the parallelization model
+  described hereafter.}
+
+\begin{listing}
+\begin{minted}[frame=single,
+               framesep=3mm,
+               linenos=false,
+               xleftmargin=21pt,
+               tabsize=4]{js}
+"dependencies" : {
+      "num_input" : {
+         "properties" : {
+            "str_input" : {
+               "not" : {}
+            }
+         }
+      },
+      "str_input" : {
+         "properties" : {
+            "num_input" : {
+               "not" : {}
+            }
+         }
+      }
+   }
+\end{minted}
+\caption{Excerpt from invocation schema showing dependencies between
+  two mutually exclusive parameters \texttt{num\_input} and
+  \texttt{str\_input}.}
+\label{listing:invocation-schema-example}
+\end{listing}
+
+\subsection{Parallelization support}
+\label{sec:parallelization}
+\boutiques supports parallelism by allowing applications to submit new
+invocations through the following protocol \todo{Needs to be
+  documented on Github}:
+\begin{enumerate}
+\item Applications write a JSON object complying to the invocation
+  schema in their execution directory.
+\item The platform reads the JSON object, creates the corresponding
+  invocation, and writes back the invocation id in the same directory.
+% The platform could also support status requests.
+\end{enumerate}
+Several invocations can be submitted at once, possibly using the
+\texttt{task-dependencies} array to specify dependencies among
+them. This mechanism is implemented in the Boutiques schema using a
+single property, \texttt{can-submit-tasks}.
+
+Although very simple, this model allows wrapping complete
+workflows. Workflows are wrapped as any other application, and they
+can be executed using any engine provided that it is installed in the
+container image referred in the application manifest. Such a ``task
+encapsulation'' model allows for a scalable and reliable execution of
+workflows expressed in a variety of languages, as detailed in
+~\cite{glatard2016fgcs}.
+
+\subsection{Containers}
+
+Applications are installed in a container image complying to the
+Docker, Singularity or rootfs format. We intentionally support
+multiple container formats as we anticipate that they will be used for
+different purposes. For instance, Docker is well suited for developers
+and users who manipulate applications on their local workstation. It
+is well documented, maintained and it has a rich ecosystem of tools to
+help build and run containers on most operating systems. Singularity
+is more suited for users and platforms that need to run applications
+on shared computing clusters. Bridges exist among these containers
+formats to convert container images across frameworks. For instance, a
+platform dedicated to high-performance computing may accept manifests
+referring to Docker containers to facilitate application integration
+by developers, and internally convert container images to Singularity
+to run applications efficiently on clusters.
+
+Container images are defined from their URL (Singularity and rootfs)
+or name in a Docker index. Container images may specify a working
+directory where the application has to be run. A hash may also be
+reported to accurately identify container images and detect updates.
+
+Containers were adopted because they allow for an automated and
+lightweight integration of application implementations in
+platforms. They are extremely useful to improve the reproducibility of
+analyses as variations in the software environment may have an
+important impact on the computed results. They also have limitations,
+in particular they do not specify the hardware architecture required
+to execute an application, which can be an issue in some cases.
+
+Containers are not mandatory though. If an application manifest does
+not contain any reference to a container, the platform will assume
+that the application is pre-installed on the execution infrastructure.
+
+\subsection{Custom properties}
+
+Custom properties may be added to the Boutiques specification without
+restriction. Custom properties are corralled in a specific JSON object
+to facilitate validation. They may be useful to implement
+platform-specific features but they should be used with care to avoid
+making applications dependent on a particular platform.
+
 \subsection{Core tools}
@@
 The Boutiques validator checks conformance of JSON manifests to the
-Boutiques schema using a basic JSON validator. It also performs an
-additional set of checks that cannot be easily implemented in JSON
-schema: value keys are unique among inputs, input and output
-identifiers are unique, input and output value keys are all
-included in the command line, value keys are not contained
-within each other (which would puzzle substitution), output path
-templates are unique (to avoid results to overwrite each other),
-inputs of type Flag have a command-line flag, are optional and are not
-lists, number and list constraints are sensible (e.g. min is lower
-than max), the default value of enum types is part of the enum, an
-input cannot both require and disable another input, required inputs
-cannot require or disable other parameters, group member identifiers
-must correspond to existing inputs and cannot appear in different
-groups, mutually exclusive groups cannot have members requiring other
-members, and that one-is-required groups should never have required
-members. The Boutiques validator is meant for application developers.
+Boutiques schema using a basic JSON validator. It also performs the
+following checks that cannot be easily implemented in JSON schema:
+value keys are unique among inputs, input and output identifiers are
+unique, input and output value keys are all included in the command
+line, value keys are not contained within each other (which would
+puzzle substitution), output path templates are unique (to avoid
+results to overwrite each other), inputs of type Flag have a
+command-line flag, are optional and are not lists, number and list
+constraints are sensible (e.g. min is lower than max), the default
+value of enum types is part of the enum, an input cannot both require
+and disable another input, required inputs cannot require or disable
+other parameters, group member identifiers must correspond to existing
+inputs and cannot appear in different groups, mutually exclusive
+groups cannot have members requiring other members, and
+one-is-required groups should never have required members. The
+Boutiques validator is meant for application developers; platforms
+should rely on regular JSON validation to avoid depending on a
+third-party tool.
 
@@
 Boutiques manifest and a set of input values represented in a CSV or
-JSON file complying with the invocation schema. It runs the
-command in a container provided that the required framework
-(e.g. Docker) is already installed. It can also generate hypothetical
-command lines from random values, for debugging purposes. The local
-executor can be used by application users, to run tools locally, or by
-platforms, to generate command lines to be run on the execution
-infrastructure.
+JSON file complying with the invocation schema. It runs the command in
+a container provided that the required framework (e.g. Docker) is
+already installed. It can also generate hypothetical command lines
+from random values for debugging purposes. The local executor can be
+used by application users to run applications locally, or by platforms
+to generate command lines to be run on the execution infrastructure.
 
-The invocation schema generator creates an invocation schema (see
-Section~\ref{sec:invocation-schema}) from a Boutiques manifest and
-validates data against it. It can be used by platform administrators
-to add invocation schemas to existing manifests, and by platforms to
-validate input data. Validation is done using a basic JSON
-validator.
+The invocation schema generator can create an invocation schema (see
+Section~\ref{sec:invocation-schema}) from a \boutiques manifest and
+validate input data against it using a regular JSON validator. It can
+be used to add invocation schemas to existing manifests.
 
-\todo{json editor}
+Finally, a JSON editor facilitates the writing of \boutiques manifests
+through a web interface.
 
@@
 transiently stores about 10 million files representing over 50TB
-distributed in 42 servers. 56 data processing tools are integrated and
+distributed in 42 servers. 56 data processing applications are integrated and
 over 340,000 processing batches have been submitted since 2010.
@@
 mechanism described in Section~\ref{sec:parallelization} so that even
-the tasks processing a single subject can be parallelized. It also
+the invocations processing a single subject can be parallelized. It also
 allows CBRAIN to leverage the efficient agent model used in PSOM, as
@@
 
-\paragraph{Other tools.} Other neuroinformatics tools were ported to CBRAIN using Boutiques:
+\paragraph{Other applications.} Other neuroinformatics applications were ported to CBRAIN using Boutiques:
 FSL tools and ICA AROMA are available at
@@
 \item Testing is critical.
-\item Workflow engines can be platforms (they can run Boutiques tools), workflows can be applications (they can be described in Boutiques).
+\item Workflow engines can be platforms (they can run Boutiques applications), workflows can be applications (they can be described in Boutiques).
 \item With Boutiques, developers can port their application once and execute it in several platforms. It removes the technological dependency to a particular platform and facilitates migration. 
 \end{itemize}
+
+The \boutiques specification adopts a bottom-up approach where new
+features are progressively added when applications or platforms
+require them. It is currently supported by \todo{X} platforms and  \todo{Y} applications
+were described.
+
+ Although the motivating use cases
+were taken from neuroinformatics, our primary field of interest,
+nothing prevents the system from being used in a variety of other
+fields.
 
@@
 workflows from individual applications. Indeed, we believe that
-workflow engines are specific tools that need complex languages to
+workflow engines are specific applications that need complex languages to
 implement the particular functions that they target. Efforts to

### 2016-11-11 paper.tex
Corrections in 3.1.1

@@
 to several systems, including 6 clusters of the Compute Canada
-high-performance computing infrastructure (totalling more than 100,000
+high-performance computing infrastructure (totaling more than 100,000
 computing cores and 40PB of disk storage) and Amazon EC2. CBRAIN
@@
 Applications in CBRAIN are integrated as Ruby classes that create web
-forms, validation and execution code. The Boutiques support in CBRAIN
-is based on a set of templates that generate such classes from the
-application manifest. Two integration modes are available to integrate
-Boutiques applications: (1) the application manifest is stored in a
-CBRAIN plugin, and the Ruby classes are generated on-the-fly when
-CBRAIN starts. This mode allows CBRAIN developers to add new features
-to all Boutiques applications (e.g. improved error message reporting)
-by updating the Ruby templates. It is, however, constrained to the
-Boutiques manifest schema: features not supported in Boutiques, e.g.,
-cluster-specific adjustments, cannot be supported. (2) CBRAIN Ruby
-classes are generated from the Boutiques manifest in an offline
-process and integrated in CBRAIN as any other application. This mode
-allows developers to customize applications by editing the generated
-Ruby classes, but the resulting applications are difficult to maintain
-in the long term, in particular when the manifests are updated.
+forms, validate parameters and run command lines on computing
+resources. \boutiques is supported through a set of templates that
+generate such classes from the application manifest. Two application
+integration modes are available:
+\begin{enumerate}
+  \item The manifest is stored in a CBRAIN plugin and the Ruby classes
+    are generated on-the-fly when CBRAIN starts. This mode allows
+    CBRAIN developers update all Boutiques applications at once by
+    editing the templates. However, it does not allow for
+    customization beyond the Boutiques schema.
+  \item Ruby classes are generated from manifests
+    through an offline process and integrated in CBRAIN as any other
+    application. This mode allows developers to customize applications
+    by editing the generated Ruby classes, but the resulting
+    applications are difficult to maintain in the long term, in
+    particular when the manifests are updated.
+\end{enumerate}
 
@@
 parent class for the application in CBRAIN. It provides more
-flexibility to the first integration model described previously. For
-instance, it was used to define a progress bar for PSOM pipelines in
-CBRAIN.
+flexibility to the first integration model described previously.
 
@@
 list). Supporting lists as a specific CBRAIN file type allows for
-improved validation, e.g. lists that contain references to
+improved validation. For instance, lists that contain references to
 non-existent or deleted files can be detected. It also allows users to
@@
 
-\subsection{Flywheel}
+\subsubsection{Flywheel}
 
@@
 
-\subsection{Pegasus}
+\subsubsection{Pegasus}
 
@@
 
-\subsection{SPINE}
+\subsubsection{SPINE}
 
@@
 
-\subsection{VIP}
+\subsubsection{VIP}
 

### 2016-11-11 paper.tex
First part of the discussion

@@
 
+\todo{Sorina, Tristan}
+
 \subsection{Ported applications}
@@
 for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
-in CBRAIN using Boutiques. The integration uses the parallelization
+in CBRAIN through Boutiques. The integration uses the parallelization
 mechanism described in Section~\ref{sec:parallelization} so that even
@@
 \paragraph{SPM.} The Statistical Parameter Mapping toolbox (SPM)~\cite{penny2011statistical}
-was exported to Boutiques through a specific \texttt{spm2boutiques}
-tool that writes Boutiques manifests from configuration files of SPM
-functions (object declarations). \todo{More on that when it is
-  completed.}
+was exported to Boutiques \ldots \todo{Guillaume}
 
-\paragraph{MICCAI challenges.} Boutiques was also used to port to the VIP platform the 23 pipelines
-involved in two computational challenges organized by the MICCAI
-conference in 2016, related to the segmentation of multiple-sclerosis
-lesions in MR images and of tumor volumes in PET images. The pipeline
-developers created Boutiques manifests and wrapped their applications
-in Docker containers. The pipelines were ported to VIP and executed on
-205 subjects in a few weeks only. Some pipelines had to be adjusted
-once ported to the platform for the following reasons:
+\paragraph{MICCAI challenges.}
+Boutiques was used to integrate 23 pipelines in the VIP platform in
+the context of two challenges organized by the MICCAI conference in
+2016, related to the segmentation of multiple-sclerosis lesions in MR
+images (MSSEG
+challenge\footnote{\url{https://portal.fli-iam.irisa.fr/msseg-challenge/overview}})
+and of tumor volumes in PET images (PETSEG
+challenge\footnote{\url{https://portal.fli-iam.irisa.fr/petseg-challenge/overview}}). The
+pipelines were ported to VIP and executed on 205 subjects in a few
+weeks only, which would not have been possible without
+\boutiques. Some pipelines had to be adjusted manually once ported to
+the platform for the following reasons:
 \begin{enumerate}
@@
   tool not supported in Boutiques by default although it could be a possible extension.
-\item A pipeline required
-  more than 10 GB of data dependencies (atlas data) which exceeded the
-  maximum size allowed for Docker containers in our setup; we solved the
-  issue by installing the data on a directory of the host server that we
-  mounted in the container. This could not possibly be supported by
-  Boutiques since it is very dependent on the platform.
+\item A pipeline required more than 10 GB of data dependencies (atlas
+  data) which exceeded the maximum size allowed for Docker containers
+  in our setup. We solved the issue by installing the data in a
+  directory of the host server that we mounted in the container. Such
+  platform-specific configurations could not possibly be supported by
+  \boutiques.
 \item A pipeline wrote more than 10 GB of intermediate data in a
@@
 by Boutiques since the instructions given to the challengers required
-that all the pipeline parameters were fixed, to facilitate
+that all the pipeline parameters were hardcoded, to facilitate
 execution. Two custom properties (\texttt{vip:miccai-challenger-email}
 and \texttt{vip:miccai-challenge-team-name}) were added to the
-Boutiques manifest to help VIP handle pipeline results for the 2016
-MICCAI challenges.
+\boutiques manifest to help post-process results in the specific
+context of MICCAI challenges.
 
@@
 
+With Boutiques, developers can port their application once and execute
+it in several platforms. It removes the technological dependency to a
+particular platform and facilitates migration.  Although the
+motivating use cases were taken from neuroinformatics, our primary
+field of interest, nothing prevents the system from being used in a
+variety of other fields.
+
+\subsection{\boutiques manifest}
+
+The \boutiques manifest specification allows describing a wide range
+of applications, but it is also getting increasingly complex through
+additions such as invocation schemas and dependencies among inputs. In
+the future, even more features might be added, for instance to allow
+for a better description of when optional outputs are produced, i.e.,
+which combination of input values is required. Extending the
+\boutiques manifest has two main goals. First, accurately describing
+application improves validation: incorrect input values and execution
+results are more precisely detected when the application manifest is
+comprehensive, which is essential to improve the reliability and
+user-friendliness of execution platforms. Secondly, a rich manifest
+schema reduces the need for custom wrappers to integrate applications
+as even convoluted command lines can be described: this is
+particularly interesting for applications installed in containers
+since adding a wrapper to the execution stack requires an update in
+the container image.
+
+Nonetheless, a complex manifest schema has a cost for application
+developers and platforms, which we address as follows. For developers,
+we maintain the set of mandatory manifest properties as small as
+possible so that simple applications can be described in a few lines
+only (see Listing~\ref{listing:minimal}). For platforms, we aim at
+supporting as many features as possible in the local executor so that
+only the following steps need to be implemented to support \boutiques
+in a platform, regardless of the complexity of the manifest:
+\begin{enumerate}
+  \item Input entry: generate the interface to allow users to enter
+    input values.
+  \item Input validation: create a JSON object from the interface,
+    validate it against the invocation schema.
+  \item Execution: generate and run the command line from the local
+    executor and input JSON object.
+  \item Output delivery: from the manifest, identify the output files
+    and deliver them to the user.
+\end{enumerate}
+In particular, advanced validation features such as dependencies
+between inputs are embedded in the regular JSON validation performed
+at step 2, without requiring the platform to support the related
+manifest properties.
+\begin{listing}
+\begin{minted}[frame=single,
+               framesep=3mm,
+               linenos=false,
+               xleftmargin=21pt,
+               tabsize=4]{js}
+{
+    "name": "echo",
+    "tool-version": "1.0",
+    "description": "A simple script to test output files",
+    "command-line": "echo [PARAM] > output.txt",
+    "schema-version": "0.4",
+    "inputs": [{
+	"id": "param",
+	"name": "Parameter",
+        "value-key": "[INPUT_FILE]",
+	"type": "File"
+    }],
+    "output-files": [{
+	"id": "output_file",
+	"name": "Output file",
+	"path-template": "output.txt"
+    }]
+
+}
+\end{minted}
+\caption{A minimal \boutiques manifest.} 
+\label{listing:minimal}
+\end{listing}
+
+\subsection{Workflow support}
+
+The parallelization framework described in~\ref{sec:parallelization}
+supports workflows by considering them as both applications and
+platforms, which we think is a powerful but underexploited
+model. Workflows are applications because they are invoked like any
+other script or executable, possibly through a command line. Workflows
+are also platforms because they may invoke other applications that
+might themselves be described in \boutiques. This model is powerful
+because it shields the \boutiques specification from specific workflow
+constructs and allows a wide range of workflow engines to be described
+uniformly.
+
+\subsection{Limitations}
+
+A few limitations remain that should be addressed in the
+future. When
+
 \begin{itemize}
 \item Security: containers may be updated after being integrated in the platforms.
-\item The manifest specification is getting increasingly complex, but
-  this complexity is required to describe real-world
-  applications. When something cannot be described using the manifest,
-  it is always possible to change the command line format by writing a
-  wrapper script for the application. However, when containers are
-  used, the wrapper script needs to be added to the container, which
-  has some overhead (in particular, the containers provided by the
-  developers cannot be used directly).
-\item Testing is critical.
-\item Workflow engines can be platforms (they can run Boutiques applications), workflows can be applications (they can be described in Boutiques).
-\item With Boutiques, developers can port their application once and execute it in several platforms. It removes the technological dependency to a particular platform and facilitates migration. 
+\item Testing of applications.
 \end{itemize}
-
-The \boutiques specification adopts a bottom-up approach where new
-features are progressively added when applications or platforms
-require them. It is currently supported by \todo{X} platforms and  \todo{Y} applications
-were described.
-
- Although the motivating use cases
-were taken from neuroinformatics, our primary field of interest,
-nothing prevents the system from being used in a variety of other
-fields.
 
@@
 
+The \boutiques specification adopts a bottom-up
+approach where new features are progressively added when applications
+or platforms require them.
+
 \section{Acknowledgments}

### 2016-11-11 paper.tex
Finished discussion

@@
 A few limitations remain that should be addressed in the
-future. When
+future. First, \boutiques moves the application porting bottleneck to
+validation. Using \boutiques, functions can be automatically exported
+from frameworks such as Nipype and SPM, creating hundreds of richly
+described applications potentially usable by end-users. However, how
+to validate that these applications actually execute as expected
+remain unclear. A testing framework could be designed and potentially
+fed by existing frameworks, to address this issue.
 
-\begin{itemize}
-\item Security: containers may be updated after being integrated in the platforms.
-\item Testing of applications.
-\end{itemize}
+Another limitation is related to the security of containerized
+applications. Since containers are usually controlled by application
+developers rather than platforms, which is a good thing to reduce
+application porting bottlenecks, nothing prevents developers to embed
+malicious code in their container at any stage of the process,
+possibly after a platform administrator inspected the
+container. Containers are currently bulky file archives that are
+cumbersome to inspect. Tools need to be developed to allow for an
+easier characterization of container contents, for instance through
+comparison digests with respect to validated base images.
 

### 2016-11-11 paper.tex
Added related work on CWL

@@
 \subsection{Configuration files}
-
+\label{sec:configuration-files}
 A large number of applications rely on configuration files rather than
@@
 
+% How to build modular container images for workflows.
+
 \section{Related work}
 
-Boutiques intentionally does not support the definition of complete
-workflows from individual applications. Indeed, we believe that
-workflow engines are specific applications that need complex languages to
-implement the particular functions that they target. Efforts to
-specify common workflow languages include the Common Workflow Language
-(CWL)~\cite{cwl} and the Interoperable Workflow Intermediate
-Representation (IWIR)~\cite{plankensteiner2011iwir}.
+\subsection{Common Workflow Language}
 
-% CWL
-The Common Workflow Language (CWL) is the closest related work to
-Boutiques, in the sense that it also allows the description of command
-lines executed in containers. According to Github, the CWL project
-started 6 months before Boutiques. The main conceptual differences
-between CWL and Boutiques are the following:
+The Common Workflow Language is the closest piece of related work as
+it supports containers. In particular, the Command Line Tool
+Description of CWL overlaps with the \boutiques manifest. According to
+Github, CWL started 6 months before Boutiques (September 2014 vs. May
+2015). The differences highlighted below were identified from version
+1.0 of the CWL Command Line Tool
+Description\footnote{\url{http://www.commonwl.org/v1.0/CommandLineTool.html}}.
+
+\subsubsection{Conceptual differences}
+
+The following differences are conceptual in the sense that they may
+not be easily addressed by CWL or Boutiques developers.
+
+First, CWL has a workflow language whereas \boutiques does not. In
+\boutiques, workflows are integrated as any other applications, except
+that they may submit other invocations to enable workflow
+parallelism. This fundamental difference has consequences on the
+complexity of CWL application descriptions and on the possibility to
+reuse existing workflows in \boutiques. The adoption of ontologies in
+CWL may also be another consequence (see below).
+
+CWL has a formal command-line definition while \boutiques is more
+flexible. CWL specifies command lines using an array of executable and
+arguments whereas \boutiques only uses a string template. \boutiques'
+template approach may create issues in some cases, but it also allows
+developers to add simple operations to an application without having
+to write a specific wrapper. For instance, a \boutiques command line
+may easily include input decompression using, e.g., the tar command in
+addition to the main tool command. On the contrary, CWL applications
+may only have a single executable. Importantly, \boutiques' template
+system allows supporting configuration files (see
+Section~\ref{sec:configuration-files}).
+
+CWL uses ontologies while \boutiques does not. Ontologies allow for
+richer definitions but they also have an overhead. The main
+consequences are the following:
 \begin{itemize}
-\item CWL has a workflow language, Boutiques does not. CWL also has a
-  command line tool description language, but its specification is
-  influenced by the fact that command line tools are included in a
-  specific workflow languages. In Boutiques, the only available
-  workflow construct is sub-tasking (``a task may submit other
-  tasks''). This is intentional as we envisage that Boutiques tools
-  may be composed through any type of workflow engine and language. For instance, this leads to have multiple output types in CWL.
-\item CWL specifies command lines formally whereas in Boutiques they
-  are assembled from simple string templates. In short, a command line
-  in CWL is specified by an executable and an ordered list of
-  parameters. While this representation has the advantage of being
-  more formal, for instance the executable can be identified
-  automatically, it also removes flexibility by imposing that a tool
-  only runs a single command. On the contrary, Boutiques allows any
-  type of bash operators to be combined on the command line. This
-  level of flexibility is important and greatly reduces the ``shim
-  syndrome''. Boutiques' template approach allows for configuration files. 
-\item CWL is validated using SALAD, which makes the ecosystem slightly
-  more complex. For instance, CWL has ``import'' and ``include''
-  directives. Boutiques is validated using plain JSON schema. \todo{ontology layer}.
-\item In CWL, the applications need to be instrumented as follows, which is not the case in Boutiques:
-\begin{itemize}
-\item Applications must write in specific directories.
-\item Applications must produce a JSON object describing the output structure.
-\end{itemize}
+\item CWL uses a specific framework for
+validation, called SALAD (Semantic Annotations for Linked Avro Data)
+whereas Boutiques uses plain JSON schema. The main goal of SALAD is to
+allow ``working with complex data structures and document formats, such
+as schemas, object references, and namespaces''. \boutiques only relies
+on the basic types required to describe and validate a command line
+syntactically. While the use of SALAD certainly allows for
+higher-level validation, and may simplify the composition and
+validation of complex workflows, it also introduces a substantial
+overhead in the specification, and platforms have to use the validator
+provided by CWL. On the contrary, a regular JSON validator can be used in \boutiques.
+\item  CWL has a rich set of types whereas \boutiques only has simple
+types. Again, this may be seen as a feature or as an overhead
+depending on the context. \boutiques tries to limit the complexity of
+the specification to facilitate its support by platforms where tools
+will be integrated.
 \end{itemize}
 
-\todo{Update from comparison summary on Google doc}
+\subsubsection{Major differences}
 
-% How to build modular container images for workflows.
+The following differences are major but they may be addressed by the
+CWL and \boutiques developers as they do not undermine the application
+description model.
+
+CWL tools have to write in a specific set of directories called
+``designated output directory'', ``designated temporary directory''
+and ``system temporary directory''. Tools are informed of the location
+of such directories through environment variables. Having to write in
+specific directories is problematic because tools have to be modified
+to enable that. In \boutiques, the path of output files is defined
+using a dedicated property.
+
+CWL types are richer, not only
+semantically but also syntactically. For instance, files have
+properties for basename, dirname, location, path, checksum, etc.
+
+\boutiques supports various types of containers (Docker, Singularity,
+rootfs) while CWL supports only Docker. At the same time, CWL’s
+requirements are richer: for instance, they may include RAM and disk
+while Boutiques is limited to walltime estimate. CWL has hints, i.e.,
+recommendations that only lead to warnings when not respected.
+
+In \boutiques, dependencies can be defined among inputs, e.g., to
+specify that an input may be used only when a particular flag is
+activated. This is a very useful feature to improve validation, in
+particular when tools have a lot of options.
+
+In \boutiques, named
+groups of inputs can be defined, which improves the presentation of
+long parameter lists for the user and enables the definition of more
+constraints within groups (e.g. mutually exclusive inputs).
+
+\subsubsection{Other minor differences}
+
+The \boutiques and CWL schemas have other minor differences that we
+summarized in a public
+spreadsheet\footnote{\url{https://docs.google.com/spreadsheets/d/1eLhpxzaaMPCIUFmV5Trs-WtaMpBO-c6QDQsmhwP7QIE/edit\#gid=0}}.
 

### 2016-11-11 paper.tex
A brief conclusion

@@
 
-The \boutiques specification adopts a bottom-up
-approach where new features are progressively added when applications
-or platforms require them.
+\boutiques is available at \url{https://github.com/boutiques}. We
+welcome feedback, issue reporting and pull requests. \boutiques adopts
+a bottom-up approach where new features are progressively added based
+on feedback from applications and platforms. Beyond the technicalities
+discussed in this paper, the availability of a solid core of
+applications and platforms in the framework is key to its success,
+which we plan to continuously enhance in the future.
 

### 2016-11-11 paper.tex
More related work

@@
 
+\subsection{Other frameworks}
+
+A large number of frameworks have been created to facilitate the
+integration of command-line applications in platforms. In
+neuroinformatics, several platforms platforms define a formal
+interface to embed command-line applications. Among them, the Common
+Toolkit\footnote{\url{http://www.commontk.org}} has an
+interoperability focus like \boutiques as its command-line modules are
+supported by several platforms such as 3D Slicer~\cite{pieper20043d},
+NiftyView~\cite{Craddock2016}, GIMIAS~\cite{larrabide2009gimias},
+MedInria~\cite{larrabide2009gimias}, MeVisLab~\cite{heckel2009object}
+and MITK workbench~\cite{nolden2013medical}. The framework, however,
+remains tightly bound to the Common Toolkit's C++ implementation which
+limits its adoption, e.g., in web platforms \todo{Fr\'ed\'eric, maybe
+  you have more to say on this}.
+
+In the distributed computing community, systems were also proposed to
+facilitate the embedding of applications in platforms. The Grid
+Execution Management for Legacy Code Architecture
+(GEMLCA~\cite{delaitre2005gemlca}) was used to wrap applications in
+grid computing systems. Interestingly, it has been used to embed
+workflow engines in the SHIWA
+platform~\cite{terstyanszky2014enabling}, in a similar but different
+way than proposed by \boutiques (Section~\ref{sec:parallelization}).
+
+The recent advent of virtual containers requires a new generation of
+application description frameworks that are independent from any
+programming language and that expose a rich set of properties to
+describe command lines as intended by \boutiques.
+
 \section{Conclusion}

### 2016-11-11 paper.tex
Minor edits

@@
 
-\note{Tristan}{Vocabulary (check consistency across the paper): tool vs. application}
-\note{Tristan}{Targeted journals: Scientific Data, Giga Science, Frontiers in Neuroinformatics, Future Generation Computer Systems}
+%\note{Tristan}{Vocabulary (check consistency across the paper): tool vs. application}
+%\note{Tristan}{Targeted journals: Scientific Data, Giga Science, Frontiers in Neuroinformatics, Future Generation Computer Systems}
 
@@
 \boutiques supports parallelism by allowing applications to submit new
-invocations through the following protocol \todo{Needs to be
-  documented on Github}:
+invocations through \notimplementedyet{the following protocol}:
 \begin{enumerate}
@@
 
-\section{Acknowledgments}
-
-Amazon grant.
-
 \bibliographystyle{plain}

### 2016-11-11 paper.tex
Updates in related work

@@
 
+The Common Workflow Language\footnote{\url{http://www.commonwl.org}}
+is the piece of related work most closely related to
+\boutiques. Several other application description frameworks have been
+developed but they do not support containers.
+
 \subsection{Common Workflow Language}
 
-The Common Workflow Language is the closest piece of related work as
-it supports containers. In particular, the Command Line Tool
-Description of CWL overlaps with the \boutiques manifest. According to
-Github, CWL started 6 months before Boutiques (September 2014 vs. May
-2015). The differences highlighted below were identified from version
-1.0 of the CWL Command Line Tool
+The Command Line Tool Description of the Common Workflow Language
+(CWL) overlaps with the \boutiques manifest. According to the history
+of the CWL and \boutiques Github repositories, CWL started 6 months
+before Boutiques (September 2014 vs. May 2015). The differences
+highlighted below were identified from version 1.0 of the CWL Command
+Line Tool
 Description\footnote{\url{http://www.commonwl.org/v1.0/CommandLineTool.html}}.
@@
 The following differences are conceptual in the sense that they may
-not be easily addressed by CWL or Boutiques developers.
+not be easily addressed by CWL or Boutiques developers without deeply
+refactoring the frameworks.
 
@@
 
-CWL tools have to write in a specific set of directories called
+\begin{itemize}
+\item CWL tools have to write in a specific set of directories called
 ``designated output directory'', ``designated temporary directory''
@@
 using a dedicated property.
-
-CWL types are richer, not only
+\item CWL types are richer, not only
 semantically but also syntactically. For instance, files have
 properties for basename, dirname, location, path, checksum, etc.
-
-\boutiques supports various types of containers (Docker, Singularity,
+\item \boutiques supports various types of containers (Docker, Singularity,
 rootfs) while CWL supports only Docker. At the same time, CWL’s
@@
 recommendations that only lead to warnings when not respected.
-
-In \boutiques, dependencies can be defined among inputs, e.g., to
+\item In \boutiques, dependencies can be defined among inputs, e.g., to
 specify that an input may be used only when a particular flag is
@@
 particular when tools have a lot of options.
-
-In \boutiques, named
+\item In \boutiques, named
 groups of inputs can be defined, which improves the presentation of
@@
 constraints within groups (e.g. mutually exclusive inputs).
+\end{itemize}
 
@@
 
-A large number of frameworks have been created to facilitate the
+Several frameworks have been created to facilitate the
 integration of command-line applications in platforms. In
-neuroinformatics, several platforms platforms define a formal
-interface to embed command-line applications. Among them, the Common
-Toolkit\footnote{\url{http://www.commontk.org}} has an
-interoperability focus like \boutiques as its command-line modules are
-supported by several platforms such as 3D Slicer~\cite{pieper20043d},
+neuroinformatics, many platforms define a formal interface to embed
+command-line applications. Among them, the Common
+Toolkit\footnote{\url{http://www.commontk.org}} interoperates with
+several platforms such as 3D Slicer~\cite{pieper20043d},
 NiftyView~\cite{Craddock2016}, GIMIAS~\cite{larrabide2009gimias},
@@
 platform~\cite{terstyanszky2014enabling}, in a similar but different
-way than proposed by \boutiques (Section~\ref{sec:parallelization}).
+way than proposed by \boutiques (see Section~\ref{sec:parallelization}).
 
@@
 programming language and that expose a rich set of properties to
-describe command lines as intended by \boutiques.
+describe command lines, as intended by \boutiques.
 

### 2016-11-11 paper.tex
Updates in discussion

@@
 
-With Boutiques, developers can port their application once and execute
-it in several platforms. It removes the technological dependency to a
+With Boutiques, developers can port their applications once and execute
+them in several platforms. \boutiques removes the technological dependency to a
 particular platform and facilitates migration.  Although the
@@
 field of interest, nothing prevents the system from being used in a
-variety of other fields.
+variety of other domains.
 
@@
 of applications, but it is also getting increasingly complex through
-additions such as invocation schemas and dependencies among inputs. In
-the future, even more features might be added, for instance to allow
-for a better description of when optional outputs are produced, i.e.,
-which combination of input values is required. Extending the
-\boutiques manifest has two main goals. First, accurately describing
-application improves validation: incorrect input values and execution
-results are more precisely detected when the application manifest is
-comprehensive, which is essential to improve the reliability and
-user-friendliness of execution platforms. Secondly, a rich manifest
-schema reduces the need for custom wrappers to integrate applications
-as even convoluted command lines can be described: this is
-particularly interesting for applications installed in containers
-since adding a wrapper to the execution stack requires an update in
-the container image.
+additions such as invocation schemas and dependencies among
+inputs. Extending the \boutiques manifest has two main goals. First,
+accurately describing applications improves validation: incorrect input
+values and execution results are more precisely detected when the
+application manifest is comprehensive, which is essential to improve
+the reliability and user-friendliness of execution
+platforms. Secondly, a rich manifest schema reduces the need for
+custom wrappers to integrate applications as even convoluted command
+lines can be described: this is particularly interesting for
+applications installed in containers since adding a wrapper to the
+execution stack requires an update in the container image.
 
@@
 A few limitations remain that should be addressed in the
-future. First, \boutiques moves the application porting bottleneck to
-validation. Using \boutiques, functions can be automatically exported
-from frameworks such as Nipype and SPM, creating hundreds of richly
-described applications potentially usable by end-users. However, how
-to validate that these applications actually execute as expected
-remain unclear. A testing framework could be designed and potentially
-fed by existing frameworks, to address this issue.
+future. First, \boutiques moves the application porting bottleneck
+from integration to validation. Using \boutiques, functions can be
+automatically exported from frameworks such as Nipype and SPM,
+creating hundreds of richly described applications potentially usable
+by end-users. However, it remains unclear how to validate that these
+applications actually execute as expected. A testing framework could
+be designed and potentially fed by existing frameworks, to address
+this issue.
 
@@
 applications and platforms in the framework is key to its success,
-which we plan to continuously enhance in the future.
+which we plan to continuously enhance.
 

### 2016-12-05 paper.tex

@@
 
-\todo{Sorina, Tristan}
+The Virtual Imaging Platform (VIP)~\cite{GLAT-13} is a web portal for medical 
+simulation  and image data analysis. VIP allows to make applications 
+available as a service, and also to benefit from the resources available 
+to the biomed Virtual Organisation (VO) of the European Grid Infrastructure 
+\footnote{\url{https://www.egi.eu/}}, supported by approximately 65 computing
+sites world-wide offering a total of 130 computing clusters
+and 5 PB of storage. The VIP service is currently deployed at the Creatis laboratory
+\footnote{\url{https://www.creatis.insa-lyon.fr/site7/en}} and uses the DIRAC 
+French service \footnote{\url{https://dirac.france-grilles.fr}} to submit jobs on 
+EGI grid and cloud resources.  VIP counts, in December 2016, more than 950 
+registered users and a growing number of available applications.
+
+Until recently, applications were integrated in VIP as workflows defined and managed 
+with the MOTEUR\cite{GLAT-08} workflow engine. As of today, Boutiques is supported 
+through a "Boutiques application importer" tool allowing to automatically import
+applications based on their JSON description.
+
+\todo{Sorina, Tristan: develop on the functionalities and current limits of the application importer?}
 

### 2016-12-08 paper.tex
Fixes @johnflavin's comment in #1

--- 2016-11-11 paper.tex
@@
          Krzysztof J. Gorgolewski, Charles G. Guttmann, Nathaniel Kofalt, Pierre-Olivier Quirion,\\
-         Pierre Rioux, Gunnar Schaeffer, Marc-\'Etienne Rousseau, Alan C. Evans}
+         Pierre Rioux, Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
 
@@
 
+\subsection{Reproducibility}
+
+\boutiques helps reproduce\footnote{in Peng's
+  sense~\cite{peng2011reproducible}} executions through containers and
+formal command-line descriptions. With \boutiques, complete sets of
+applications could be easily migrated across execution platforms,
+including high-performance computing platforms and individual laptops,
+to reproduce analyses. The \boutiques manifest describes the
+parameters and implementation of the application, and the invocation
+schema describes the parameter values.
+
+However, reproducibility is a large problem that \boutiques only
+partially addresses. At the command-line execution level, containers
+help freeze a large fraction of the software ecosystem but they do not
+shield against discrepancies arising from different Linux kernel
+versions or hardware platforms. For instance, containers may not
+execute consistently on different x86\_64 CPUs (e.g. Intel and AMD)
+when software compiled with architecture-specific flags is involved
+(see for instance GCC's \texttt{-march} flag).
+
+In addition, important runtime parameters, for instance related to
+multi-threading or available resources (storage, RAM), may be set by
+the execution platform without being specified in the \boutiques
+manifest. Such runtime parameters may impact reproducibility in some
+cases. To properly cover this issue, \boutiques manifests should be
+complemented by a provenance framework that captures a detailed trace
+of the execution.
+
+At a higher level, reproducibility is impacted by factors
+such as data sharing~\cite{poldrack2014making} which are out of the
+scope of \boutiques.
+
 \subsection{Limitations}

### 2016-12-08 paper.tex
A paragraph on other application types, following discussion with @merous

@@
 
+\subsection{Application types}
+
+So far, \boutiques has focused on the description of non-interactive
+command-line applications. While such applications cover a large
+fractions of the tools involved in scientific data processing, other
+types of programs exist such as web services, interactive tools, and
+applications with a graphical user interface (GUI).  Such application
+classes could be described in \boutiques through a command-line
+mapping. For instance, web services may be wrapped as command-line
+applications using tools such as \texttt{curl} or
+\texttt{wget}. Interactive applications may also be transformed to
+non-interactive ones through configuration files. Finally, nothing
+prevents a \boutiques tool from popping up a window for a user to
+provide input through a GUI. This should, however, be specified
+as an extension to the manifest since most platforms would not support
+this feature by default. Graphical output produced by applications
+executed in containers should also be treated specifically.
+
 \subsection{Limitations}

### 2016-12-13 paper.tex
Added VIP importer and applications descriptions

@@
 
-Until recently, applications were integrated in VIP as workflows defined and managed 
-with the MOTEUR\cite{GLAT-08} workflow engine. As of today, Boutiques is supported 
-through a "Boutiques application importer" tool allowing to automatically import
-applications based on their JSON description.
-
-\todo{Sorina, Tristan: develop on the functionalities and current limits of the application importer?}
+Until recently, applications were manually integrated in VIP as workflows written 
+in the Gwendia\cite{MONT-09} language and managed  with the MOTEUR\cite{GLAT-08} workflow engine. 
+As of today, Boutiques is supported through a "Boutiques application importer" tool, which 
+parses the JSON file and automatically generates the corresponding application workflow
+and the wrapper script that handles, among others, the execution of the command line itself. 
+The use of an application workflow allows to (i) iterate on input lists and generate 
+parallel tasks, as well as to (ii) concatenate multiple tools. The applications used in 
+the MICCAI challenges, for example, were followed by other tool allowing to compute 
+challenge-specific metrics. Tool concatenation is currently handeled at the importer 
+level based on pre-defined workflow templates.
+
+\todo{Tristan: validate the description above}
 
@@
 
-\todo{Sorina, a paragraph on other applications in VIP?}
+\paragraph{VIP applications.}A few tools were imported in VIP using Boutiques.
+
+The first one is a phase contrast simulator called CreaPhase and developed at Creatis. 
+The CreaPhase inputs had certain particularies (some nedeed to be enclosed 
+in simple quotes, other were vectors of variable size enclosed in brackets)
+that required the post-processing of the wrapper script generated by the
+Boutiques importer.
+
+The Odin MRI simulator\footnote{\url{http://od1n.sourceforge.net}} was also imported 
+in VIP using Boutiques. Since Odin  may require important amounts of computing resources 
+(the common use-case consists in iterating on a large number of inputs files), Odin is 
+executed by VIP on the EGI grid resources, which currently do not support Docker. Consequently, 
+the Odin executable was exported from the Docker image which allowed for his CentOS compilation
+and the Odin wrapper script was modified so that it makes no "Docker run" call.
+
+A few MRtrix3\footnote{\url{http://www.mrtrix.org/}} tools were also make available in VIP. 
+A script developed at Creatis, which combines MRtrix3 and FSL tools, is currently being
+imported and tested.
+
+\todo{Tristan: validate the descriptions above}
 

### 2016-12-18 paper.tex
updated manuscript; added ndmg bits

@@
         Rafael Ferreira da Silva, Guillaume Flandin, John Flavin, Pascal Girard, \\
-         Krzysztof J. Gorgolewski, Charles G. Guttmann, Nathaniel Kofalt, Pierre-Olivier Quirion,\\
-         Pierre Rioux, Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
+        Krzysztof J. Gorgolewski, Gregory Kiar, Charles G. Guttmann, Nathaniel Kofalt, \\
+        Pierre-Olivier Quirion, Pierre Rioux, Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
 
@@
 value keys. The command line may actually
-encompass several commands separated, e.g., by semicolumns, pipes
+encompass several commands separated, e.g., by semicolons, pipes
 (\texttt{|}) or ampersands (\texttt{\&}). Such a flexible command-line
@@
 \paragraph{Types.} Inputs may be of type \texttt{String},
-\texttt{Number}, \texttt{Flag} or \texttt{File}. \texttt{File} may
-also represents a directory. Types can be restricted to a specific set
-of values (\texttt{String} and \texttt{Number} only), to a specific
+\texttt{Number}, \texttt{Flag} %, \texttt{Enum},
+or \texttt{File}. \texttt{File} may
+also represent a directory. Types can be restricted to a specific set
+of values (\texttt{String} %, \texttt{Enum},
+and \texttt{Number} only), to a specific
 range (\texttt{Number} only), to integers or to booleans (\texttt{Number}
-only).
+only) %, or to an arbitrary but specified set of values (\texttt{Enum} only)
+.
 
@@
 value of enum types is part of the enum, an input cannot both require
+\note{greg}{should Enum be capitalized? It is earlier in the document}
 and disable another input, required inputs cannot require or disable
@@
 
+\paragraph{ndmg.} The NeuroData MRI to Graphs one-click connectome estimation
+pipeline~\cite{kiar2016ndmg}, developed in Python and leveraging FSL, was
+exported to \boutiques and is available at \url{https://github.com/neurodata/boutiques-tools}.
+
 \paragraph{Other applications.} Other neuroinformatics applications were ported to CBRAIN using Boutiques:
@@
     HCP PreFreesurfer                &                   &                   &                   &                   &                   &                   &                   \\
+    ndmg                &                   &       \cellcolor{gray!75}            &                   &          \cellcolor{gray!75}         &      \cellcolor{gray!75}             &                   &                   \\
     MICCAI challenge (23 pipelines)\\&                   &                   &                   &                   &                   &                   & \\
@@
     \multicolumn{1}{r|}{first}      &                    &            &         &              &&\\
-    FSL Nipype interfaces     \\
+    FSL Nipype interfaces           &                    &            &         &              &&\\
     HCP PreFreesurfer               &\cellcolor{gray!75} &            &         &              &&\\
+    ndmg                            &\cellcolor{gray!75} &            &         &              &&\\
     MICCAI challenge (23 pipelines) &\cellcolor{gray!75} &            &         &              &&\\
@@
 \end{table}
-
+\note{greg}{@tristan g: chris helped me verify that ndmg also works with singularity, but we have not shared
+the image anywhere. Would it be worth me publishing that image for this table?}
 \section{Discussion}
@@
 CWL has a formal command-line definition while \boutiques is more
-flexible. CWL specifies command lines using an array of executable and
+flexible. CWL specifies command lines using an array of executables and
 arguments whereas \boutiques only uses a string template. \boutiques'

### 2016-12-19 paper.tex
added affiliation

@@
         Pierre-Olivier Quirion, Pierre Rioux, Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
-
+\note{greg}{gk's affiliation: Department of Biomedical Engineering, Johns Hopkins University, Baltimore, MD}
 \maketitle
@@
 Finally, a JSON editor facilitates the writing of \boutiques manifests
-through a web interface.
+through a web interface and a Jupyter Notebook tutorial facilitates
+new users getting started with Boutiques..
 

### 2016-12-19 paper.tex
Fixes following @gkiar's input.

@@
         Rafael Ferreira da Silva, Guillaume Flandin, John Flavin, Pascal Girard, \\
-        Krzysztof J. Gorgolewski, Gregory Kiar, Charles G. Guttmann, Nathaniel Kofalt, \\
+        Krzysztof J. Gorgolewski, Charles G. Guttmann, Gregory Kiar, Nathaniel Kofalt, \\
         Pierre-Olivier Quirion, Pierre Rioux, Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
+
 \note{greg}{gk's affiliation: Department of Biomedical Engineering, Johns Hopkins University, Baltimore, MD}
+
 \maketitle
@@
 \begin{verbatim}
-exampleTool_1 [STRING_INPUT] [FILE_INPUT] [ENUM_INPUT] | \
+exampleTool_1 [STRING_INPUT] [FILE_INPUT]  | \
                               exampleTool_2 [FLAG_INPUT] [NUMBER_INPUT] >> [LOG].txt
@@
 \paragraph{Types.} Inputs may be of type \texttt{String},
-\texttt{Number}, \texttt{Flag} %, \texttt{Enum},
+\texttt{Number}, \texttt{Flag} 
 or \texttt{File}. \texttt{File} may
 also represent a directory. Types can be restricted to a specific set
-of values (\texttt{String} %, \texttt{Enum},
+of values (\texttt{String} 
 and \texttt{Number} only), to a specific
 range (\texttt{Number} only), to integers or to booleans (\texttt{Number}
-only) %, or to an arbitrary but specified set of values (\texttt{Enum} only)
-.
+only).
 
@@
 constraints are sensible (e.g. min is lower than max), the default
-value of enum types is part of the enum, an input cannot both require
-\note{greg}{should Enum be capitalized? It is earlier in the document}
+value of restricted types is part of the restriction, an input cannot both require
 and disable another input, required inputs cannot require or disable
@@
 through a web interface and a Jupyter Notebook tutorial facilitates
-new users getting started with Boutiques..
+new users getting started with Boutiques.
 
@@
     \rowcolor[gray]{0.9}
-                                     & Dependencies      &Groups             & Lists             & Optional          &Default            & Enum              & Min/max\\
+                                     & Dependencies      &Groups             & Lists             & Optional          &Default            & Type restriction             & Min/max\\
 
@@
 CWL has a formal command-line definition while \boutiques is more
-flexible. CWL specifies command lines using an array of executables and
-arguments whereas \boutiques only uses a string template. \boutiques'
-template approach may create issues in some cases, but it also allows
-developers to add simple operations to an application without having
-to write a specific wrapper. For instance, a \boutiques command line
-may easily include input decompression using, e.g., the tar command in
-addition to the main tool command. On the contrary, CWL applications
-may only have a single executable. Importantly, \boutiques' template
-system allows supporting configuration files (see
-Section~\ref{sec:configuration-files}).
+flexible. CWL specifies command lines using an array containing an
+executable and a set of arguments whereas \boutiques only uses a
+string template. \boutiques' template approach may create issues in
+some cases, but it also allows developers to add simple operations to
+an application without having to write a specific wrapper. For
+instance, a \boutiques command line may easily include input
+decompression using, e.g., the tar command in addition to the main
+tool command. On the contrary, CWL applications may only have a single
+executable. Importantly, \boutiques' template system allows supporting
+configuration files (see Section~\ref{sec:configuration-files}).
 

### 2017-10-04 paper.tex
Added new applications imported into VIP and 2 comments/questions to Tristan

@@
 parallel tasks, as well as to (ii) concatenate multiple tools. The applications used in 
-the MICCAI challenges, for example, were followed by other tool allowing to compute 
+the MICCAI challenges, for example, were followed by other tools allowing to compute 
 challenge-specific metrics. Tool concatenation is currently handeled at the importer 
@@
 
-\todo{Tristan: validate the description above}
+\note{sorina}{@tristan: some info on workflows is available in the paragraph above. Is it OK/enough or do you have other suggestions?}
 
@@
 
-A few MRtrix3\footnote{\url{http://www.mrtrix.org/}} tools were also make available in VIP. 
-A script developed at Creatis, which combines MRtrix3 and FSL tools, is currently being
-imported and tested.
-
-\todo{Tristan: validate the descriptions above}
+A few MRtrix3\footnote{\url{http://www.mrtrix.org/}} tools were also made available in VIP, as well as 
+a script developed at Creatis, which combines MRtrix3 and FSL tools.
+
+A few other applications from the FLI (France Life Imaging) consortium were also imported in
+VIP using Boutiques. One of them is Ginfizz, a FMRI pipeline allowing to execute multiple 
+preprocessing steps and basic analyses such as: (i) realignment (motion registration and correction),
+(ii) slice timing (temporal interpolation to correct for the delay of acquisition between slices), 
+(iii) coregistration (registering functional and structural data), (iv) normalization (registering 
+of functional and anatomical data to a match a template that has to be specified), and (iv)
+smoothing (spatial filtering to increase Signal to Noise ratio).
+The use of Docker containers (\url{https://hub.docker.com/r/fliiam/ginfizz} based on \url{https://hub.docker.com/r/miykael/nipype_level2})
+greatly simplifies its deployment. The animaN4BiasCorrection tool from the Anima  \footnote{\url{https://github.com/Inria-Visages/Anima-Public/wiki}} project was also made available 
+in VIP. The import process was straightforward.
+
+\note{sorina}{@tristan: please check if you agree with the descriptions above}
 

### 2017-10-11 paper.tex
abstract edits

--- 2016-12-19 paper.tex
@@
         Pierre-Olivier Quirion, Pierre Rioux, Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
-\note{greg}{gk's affiliation: Department of Biomedical Engineering, Johns Hopkins University, Baltimore, MD}
 \maketitle
@@
   format compatible with the execution platform, and 3)~generating
-  proper user interfaces. Due to the variety of platforms, application
-  porting efforts are often replicated several times while
-  mutualization would save cost and improve the quality of the ported
-  applications. We describe \boutiques, a system that allows for
-  automatic import and exchange of applications across
-  platforms. \boutiques adopts a rich, flexible application
-  description format and it relies on Linux containers to install
-  applications across platforms in a lightweight manner. It is
-  currently supported by \todo{X} platforms and was used to describe
-  \todo{Y} applications.  }
+  proper user interfaces. Due to the variety of computing platforms,
+  application porting efforts are often replicated several times while
+  interoperability would save cost and improve the quality of the ported
+  applications. We describe \boutiques, a descriptive framework and
+  command line utility that allows for automatic import and exchange of
+  applications across platforms. \boutiques adopts a rich, flexible
+  application description format which relies on Linux container
+  environments such as Docker and Singularity to install applications
+  across platforms in a lightweight manner. Boutiques is currently used
+  by three distinct virtual research platforms, and has been used to
+  describe dozens of neuroimaging tools.}
 

### 2017-10-11 boutiquespaper.tex
minor intro updates, added makefile, renamed paper file, updated instructions`

--- 2017-10-11 paper.tex
@@
   by three distinct virtual research platforms, and has been used to
-  describe dozens of neuroimaging tools.}
+  describe dozens of neuroinformatics tools.}
 
@@
 
-Platforms such as web services, portals, virtual research
-environments, science gateways and workflow engines often rely on
-third-party data processing and simulation applications to deliver
-services to their users. Such applications are often repeatedly and
-manually ported to various platforms whereas this porting effort would
-greatly benefit from being mutualized. Meanwhile, virtual container
-systems such as Docker and Singularity greatly facilitate the sharing
-of software and improve the reproducibility of analyses by defining
+Platforms such as web services, portals, and virtual research
+environments are becoming an common solution for increasing the
+accessibility of high performance computing. These scientific gateways and
+workflow engines often rely on third-party data processing and simulation
+applications to deliver services to their users. Such applications are
+often repeatedly and manually ported to various platforms whereas this
+porting effort would greatly benefit from being mutualized. Meanwhile,
+virtual container systems such as Docker and Singularity greatly facilitate
+the sharing of software and improve the reproducibility of analyses by defining
 immutable, reusable execution environments. However, containers need
 to be properly annotated to allow platforms to leverage the software
-they contain. Basically, the command lines that can be run in a
-container must be formally described so that they can be automatically
-created and executed by platforms. Describing command lines is not
-straightforward since they often contain implicit semantics specified
-in human-readable documents such as manuals.
+they contain. The command lines that can be run in a container must be formally
+described so that they can be automatically created and executed by platforms.
+Describing command lines is not straightforward since they often contain
+implicit semantics specified in human-readable documents such as manuals.
 
@@
 template comprising the inputs it requires and the outputs it
-produces. Inputs may be inter-dependent, e.g., input A cannot be used
-if input B is used, and passed in configuration files or directly on
-the command line. Such formal descriptions, also called
+produces. Inputs may be inter-dependent, e.g., input \emph{A} cannot be used
+if input \emph{B} is used; inputs may be passed in configuration files
+or directly on the command line. Such formal descriptions, also called
 \emph{manifests}, allow for advanced validation of input values to
-prevent errors. Manifests refer to containers where applications and
-all their dependencies are installed. \boutiques also provides a
+prevent errors. A manifest refers to a container where a given application
+and all their dependencies are installed. \boutiques also provides a
 simple parallelization framework that allows applications such as
@@
 remainder, tasks are called \emph{invocations}). A set of core tools
-facilitate the construction, validation and execution of \boutiques
-manifests. A repository orchestrates these tools together. \boutiques
-manifests are intented to be produced by scientific application
-developers, to be published in common repositories and to be consumed
-by execution platforms.
+facilitate the construction, validation, import, execution, and publishing
+of \boutiques manifests. \boutiques manifests are intented to be produced by
+scientific developers and stored alongside their application, indexed by common
+repositories, and to be consumed by execution platforms.
 

### 2017-10-11 boutiquespaper.tex

--- 2017-10-04 paper.tex
@@
         Pierre-Olivier Quirion, Pierre Rioux, Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
-
-\note{greg}{gk's affiliation: Department of Biomedical Engineering, Johns Hopkins University, Baltimore, MD}
-
 \maketitle
@@
   format compatible with the execution platform, and 3)~generating
-  proper user interfaces. Due to the variety of platforms, application
-  porting efforts are often replicated several times while
-  mutualization would save cost and improve the quality of the ported
-  applications. We describe \boutiques, a system that allows for
-  automatic import and exchange of applications across
-  platforms. \boutiques adopts a rich, flexible application
-  description format and it relies on Linux containers to install
-  applications across platforms in a lightweight manner. It is
-  currently supported by \todo{X} platforms and was used to describe
-  \todo{Y} applications.  }
+  proper user interfaces. Due to the variety of computing platforms,
+  application porting efforts are often replicated several times while
+  interoperability would save cost and improve the quality of the ported
+  applications. We describe \boutiques, a descriptive framework and
+  command line utility that allows for automatic import and exchange of
+  applications across platforms. \boutiques adopts a rich, flexible
+  application description format which relies on Linux container
+  environments such as Docker and Singularity to install applications
+  across platforms in a lightweight manner. Boutiques is currently used
+  by three distinct virtual research platforms, and has been used to
+  describe dozens of neuroinformatics tools.}
 

### 2017-10-11 boutiquespaper.tex
minor edits up to configuration files

@@
 installed. It may also contain an invocation schema used for input
-validation. At runtime, the execution platform builds the command line
-from the manifest and values entered by the user. The platform runs
-the command line on the execution infrastructure, e.g., a server, a
-cluster or a cloud, within a container whenever available. To build
-and run the command line, the platform may rely on the \boutiques core
-tools, in particular on the validator and local executor. Application
-manifests may be stored in the repository which provides additional
-features.
+validation (this will be created at runtime if it is not found). At
+runtime, the execution platform builds the command line from the manifest
+and values entered by the user. The platform runs the command line on the
+execution infrastructure, e.g., a server, a cluster or a cloud, within a
+container whenever available. To build and run the command line, the
+platform may rely on the \boutiques core tools, packaged through the 
+\texttt{bosh} command-line utility, such as the validator and executor.
 
@@
 template may contain placeholders for input and output values, called
-value keys. The command line may actually
-encompass several commands separated, e.g., by semicolons, pipes
-(\texttt{|}) or ampersands (\texttt{\&}). Such a flexible command-line
-specification is meant to facilitate the embedding of minor mundane
-operations on the command line, for instance directory creation, input
-decompression or output archival.
+value keys. The command line may actually encompass several commands
+separated by semicolons, pipes (\texttt{|}) or ampersands (\texttt{\&}).
+Such a flexible command-line specification is meant to facilitate the
+embedding of minor mundane operations on the command line, for instance
+directory creation, input decompression or output archival.
 
@@
 \end{verbatim}
-The template contains six value keys for input and output values, that
+The template contains six value keys for input and output values that
 will be replaced by values and file names according to the user input
@@
 \paragraph{General properties.} Inputs must have a name, a unique
-identifier and a type. They may be optional, have a description, a
+identifier, and a type. They may be optional, have a description, a
 value key, a flag and flag separator, and a default value. Inputs may
@@
 
-\paragraph{Types.} Inputs may be of type \texttt{String},
-\texttt{Number}, \texttt{Flag} 
-or \texttt{File}. \texttt{File} may
-also represent a directory. Types can be restricted to a specific set
-of values (\texttt{String} 
-and \texttt{Number} only), to a specific
-range (\texttt{Number} only), to integers or to booleans (\texttt{Number}
-only).
+\paragraph{Types.} Inputs may be of type \texttt{String}, \texttt{Number},
+\texttt{Flag} or \texttt{File}. \texttt{File} may also represent a directory.
+Types can be restricted to a specific set of values (\texttt{String} 
+and \texttt{Number} only), to a specific range (\texttt{Number} only), to
+integers or to booleans (\texttt{Number} only).
 
 \paragraph{Groups and dependencies.} Groups of inputs may be defined
-with a identifier, name and list of input identifiers. Groups may be
+with an identifier, name, and list of input identifiers. Groups may be
 used to improve the presentation in a graphical user interface and to
-specify the following constraints among inputs: (1)
-\texttt{mutually-exclusive}: only one input in the group may have a
-value; (2) \texttt{one-is-required}: at least one input in the group
-must have a value. Dependencies among inputs may also be defined
-regardless of a particular group: an input may (1) require a list of
-inputs and (2) disable a list of inputs.
-
-Listing~\ref{listing:input-example} shows the definition of an input
-in the command line exemplified above. According to this definition
-and assuming that the input value entered by the user is 0.3, the
-string \texttt{[NUMBER\_INPUT]} will be replaced by \texttt{-n=0.3} on
-the command line at execution time.
+specify the following constraints among inputs:
+(1)~\texttt{mutually-exclusive}: only one member in the group may have a
+value; (2)~\texttt{one-is-required}: at least one member in the group
+must have a value; (3)~\texttt{all-or-none}: if any of the members are
+present all members must be present. Dependencies among inputs may also be
+defined regardless of a particular group: an input may (1)~require a list of
+inputs and (2)~disable a list of inputs.
+
+Listing~\ref{listing:input-example} shows the definition of an input in the
+command line exemplified above. According to this definition and assuming
+that the input value entered by the user is 0.3, the string
+\texttt{[NUMBER\_INPUT]} will be replaced by \texttt{-n=0.3} on the command
+line at execution time.
 \begin{listing}
@@
 \end{minted}
-\caption{Input example.} 
+\caption{Example of a \texttt{Number}-type input.} 
 \label{listing:input-example}
@@
 
-In \boutiques, output files must have a unique identifier, a name and
+In \boutiques, output files must have a unique identifier, a name, and
 a path template that specifies the file or directory name. Path
@@
 after the input values. In this case, input values may be stripped
-from specific strings, e.g., file extensions before being substituted
+from specific strings, e.g., file extensions, before being substituted
 in the path template. Output files may also have a description, a
@@
 \end{minted}
-\caption{Output file example} 
+\caption{Example of an output leveraging \texttt{path-template} search-and-replacement.} 
 \label{listing:output-example}
@@
 \end{minted}
-\caption{Configuration file example. The file template is defined as
+\caption{Example of a configuration input file. The file template is defined as
   an array of strings to allow for multi-line strings in JSON.}

### 2017-10-12 boutiquespaper.tex
updated methods section

@@
 %%values in the list.
+% gk: I agree with not including the above
 Algorithm~\ref{algo:command-line} shows how the command line is
@@
 even for their own applications, invocation schemas can be generated
-automatically using a core \boutiques tool. The invocation schema is
-stored as an optional property of the Boutiques manifest.
+automatically by the \boutiques command-line utility. The invocation
+schema is stored as an optional property of the \boutiques manifest.
 
@@
                tabsize=4]{js}
+{
   "dependencies" : {
@@
    }
+}
 \end{minted}
@@
 different purposes. For instance, Docker is well suited for developers
-and users who manipulate applications on their local workstation. It
-is well documented, maintained and it has a rich ecosystem of tools to
-help build and run containers on most operating systems. Singularity
+and users who manipulate applications on their local workstations or the
+cloud. It is well documented, maintained and it has a rich ecosystem of
+tools to help build and run containers on most operating systems. Singularity
 is more suited for users and platforms that need to run applications
@@
 
-Container images are defined from their URL (Singularity and rootfs)
-or name in a Docker index. Container images may specify a working
-directory where the application has to be run. A hash may also be
+Container images are defined from their URL (rootfs) or name in the
+Docker or Singularity indices. Manifests and container images may specify a
+working directory where the application has to be run, and a hash may also be
 reported to accurately identify container images and detect updates.
@@
 
-Containers are not mandatory though. If an application manifest does
-not contain any reference to a container, the platform will assume
-that the application is pre-installed on the execution infrastructure.
+%%Containers are not mandatory though. If an application manifest does
+%%not contain any reference to a container, the platform will assume
+%%that the application is pre-installed on the execution infrastructure. 
+% gk: I believe we only support containers now, correct?
 
@@
 platform-specific features but they should be used with care to avoid
-making applications dependent on a particular platform.
+making applications dependent on a particular platform or replacing
+existing functionality already represented in the \boutiques schema.
 
@@
 
-The Boutiques validator checks conformance of JSON manifests to the
-Boutiques schema using a basic JSON validator. It also performs the
-following checks that cannot be easily implemented in JSON schema:
+\boutiques is available on Python through the PyPi package repository as
+``\texttt{boutiques}''. The \boutiques package exposes a command-line
+utility to the user, \texttt{bosh}, which contains entrypoints for all
+core functions within \boutiques. There is help-text which should guide
+users through interacting with the \texttt{bosh} utility. The core tools
+provided by \boutiques are, the: validator, executor (for both launch and
+simulation), invocation schema handler, importer, and publisher. The tools
+exposed through the command-line API are also available consistently through
+the Python interface by importing the \boutiques package.
+
+\paragraph{Validator.} The \boutiques validator checks conformance of JSON
+manifests to the \boutiques schema using a basic JSON validator. It also
+performs the following checks that cannot be easily implemented in JSON schema:
 value keys are unique among inputs, input and output identifiers are
 unique, input and output value keys are all included in the command
-line, value keys are not contained within each other (which would
-puzzle substitution), output path templates are unique (to avoid
-results to overwrite each other), inputs of type Flag have a
-command-line flag, are optional and are not lists, number and list
-constraints are sensible (e.g. min is lower than max), the default
+line, identifiers with the same value key are mutually exclusive, value keys
+are not contained within each other (which would puzzle substitution), output
+path templates are unique (to avoid results to overwrite each other), inputs of
+type Flag have a command-line flag, are optional and are not lists, the default
 value of restricted types is part of the restriction, an input cannot both require
@@
 inputs and cannot appear in different groups, mutually exclusive
-groups cannot have members requiring other members, and
-one-is-required groups should never have required members. The
-Boutiques validator is meant for application developers; platforms
-should rely on regular JSON validation to avoid depending on a
-third-party tool.
-
-The local executor can generate and execute command line from a
-Boutiques manifest and a set of input values represented in a CSV or
-JSON file complying with the invocation schema. It runs the command in
-a container provided that the required framework (e.g. Docker) is
-already installed. It can also generate hypothetical command lines
-from random values for debugging purposes. The local executor can be
-used by application users to run applications locally, or by platforms
-to generate command lines to be run on the execution infrastructure.
-
-The invocation schema generator can create an invocation schema (see
-Section~\ref{sec:invocation-schema}) from a \boutiques manifest and
-validate input data against it using a regular JSON validator. It can
-be used to add invocation schemas to existing manifests.
-
-Finally, a JSON editor facilitates the writing of \boutiques manifests
-through a web interface and a Jupyter Notebook tutorial facilitates
-new users getting started with Boutiques.
+groups cannot have members requiring other members, one-is-required groups
+should never have required members, and all-or-none group members must not be
+required.
+
+\paragraph{Executor.} The executor has two modes of operation: \emph{launch}, and
+\emph{simulate}. The simulate mode can generate hypothetical command lines from
+random values given the manifest (and corresponding invocation schema) for debugging
+purposes, or display the command that would be executed given a provided valid
+invocation. The launch mode can execute command line from a \boutiques manifest and
+a set of input values represented in JSON file complying with the invocation schema.
+It runs the command in a container provided that the required framework (e.g. Docker)
+is already installed. The executor can be used by application users to run
+applications locally, or by platforms to generate command lines to be run on the
+execution infrastructure.
+
+\paragraph{Invocation Schema Handler.} The invocation schema handler can create an
+invocation schema (see Section~\ref{sec:invocation-schema}) from a \boutiques manifest
+and validate input data against it using a regular JSON validator. It can be used to add
+invocation schemas to existing manifests. It is used by the executor if no invocation
+schema is present in the \boutiques manifest being deployed.
+
+\paragraph{Importer.} The importer takes \boutiques manifests from older versions and
+updates them to be compliant with the most recent version of the schema. This tool can
+also create manifests from select application collections, such as BIDS
+apps~\cite{gorgolewski2017bids}.
+
+\paragraph{Publisher.} As \boutiques has primarily been adopted in the neuroinformatics
+community, the publisher gets a further description (such as author, website, etc.) of
+the described tool, and adds an index to it on
+NeuroLinks\footnote{\url{https://brainhack101.github.io/neurolinks/}}, a repository
+containing links to useful resources and tools. This enables others to find and use the
+descriptor for your tool easily. This functionality could be extended to
+new repositories for other domains, such as Bioconductor for bioinformatics.
+
+Though not a component of the \boutiques tool chain, a Jupyter Notebook tutorial exists
+to facilitate new users getting started with \boutiques.
 
 \subsection{Repository}
-
+%% gk: I think we should remove this and just have the reference in Publisher
 \todo{Gunnar, Nathaniel, John}
@@
 
-The following platforms are able to integrate and execute Boutiques
+The following platforms are currently able to integrate and execute Boutiques
 applications.

### 2017-10-13 boutiquespaper.tex
more minor edits (done for the day)
updated results, discussion, and conclusion

@@
 %%values in the list.
-% gk: I agree with not including the above
+%% gk: I agree with not including the above
 Algorithm~\ref{algo:command-line} shows how the command line is
@@
 %%that the application is pre-installed on the execution infrastructure. 
-% gk: I believe we only support containers now, correct?
+%% gk: I believe we only support containers now, correct?
 
@@
 
-The following platforms are currently able to integrate and execute Boutiques
-applications.
+\boutiques application import and execution is currently supported by several
+platforms which aid in execution and tool management. These are enumerated
+below.
 
@@
 on the infrastructure provided by Compute Canada~\cite{das2016mni}. It
-currently provides 460+ collaborators in 20 countries with web access
+currently provides 500+ collaborators in 22 countries with web access
 to several systems, including 6 clusters of the Compute Canada
@@
 transiently stores about 10 million files representing over 50TB
-distributed in 42 servers. 56 data processing applications are integrated and
-over 340,000 processing batches have been submitted since 2010.
+distributed in 42 servers. 51 public data processing applications are
+integrated and over 340,000 processing batches have been submitted since 2010.
 
@@
     are generated on-the-fly when CBRAIN starts. This mode allows
-    CBRAIN developers update all Boutiques applications at once by
+    CBRAIN developers update all \boutiques applications at once by
     editing the templates. However, it does not allow for
-    customization beyond the Boutiques schema.
+    customization beyond the \boutiques schema.
   \item Ruby classes are generated from manifests
@@
 \end{enumerate}
+%% gk: I think we can skip this last bit, since there is only one type of
+%% \emph{boutiques} integration with cbrain, we needn't mention the other?
 
 A custom property (\texttt{cbrain:inherits-from-class}) was added to
-the Boutiques manifest to define the Ruby class that should be used as
+the \boutiques manifest to define the Ruby class that should be used as
 parent class for the application in CBRAIN. It provides more
@@
 CBRAIN was also extended to support the parallelization mechanism
-available in Boutiques: the CBRAIN execution agent periodically scans
+available in \boutiques: the CBRAIN execution agent periodically scans
 the execution directories of the parallelized applications and it submits
@@
 A new list mechanism was also introduced in CBRAIN to facilitate the
-iteration of Boutiques applications on large sets of files. CBRAIN
+iteration of \boutiques applications on large sets of files. CBRAIN
 lists are specific files that contain references to other CBRAIN
-files. When a list is passed to a Boutiques application, the elements
+files. When a list is passed to a \boutiques application, the elements
 in the list are either concatenated in a single command line (when the
-corresponding Boutiques input is a list), or a new command line is
+corresponding \boutiques input is a list), or a new command line is
 generated for every element in the list (when the input is not a
@@
 \subsubsection{Flywheel}
-
+%% gk: shall we check with them, and if nothing, remove?
 \todo{Gunnar, Nathanael, John}
@@
 \subsubsection{Pegasus}
-
+%% gk: to what extent does this exist?
 \todo{Rafael}
@@
 \subsubsection{SPINE}
-
+%% gk: shall we check with them, and if nothing, remove?
 \todo{Charles}
@@
 \footnote{\url{https://www.egi.eu/}}, supported by approximately 65 computing
-sites world-wide offering a total of 130 computing clusters
-and 5 PB of storage. The VIP service is currently deployed at the Creatis laboratory
+sites world-wide offering a total of 130 computing clusters and 5 PB of storage.
+The VIP service is currently deployed at the Creatis laboratory
 \footnote{\url{https://www.creatis.insa-lyon.fr/site7/en}} and uses the DIRAC 
 French service \footnote{\url{https://dirac.france-grilles.fr}} to submit jobs on 
-EGI grid and cloud resources.  VIP counts, in December 2016, more than 950 
+EGI grid and cloud resources.  VIP counts, in October 2017, more than 1000 
 registered users and a growing number of available applications.
+%% gk: I guessed 50 more in the last year, since 1000 is a nice number, is this accurate?
 
@@
 parses the JSON file and automatically generates the corresponding application workflow
-and the wrapper script that handles, among others, the execution of the command line itself. 
+and the wrapper script that handles, among other things, the execution of the command line itself. 
 The use of an application workflow allows to (i) iterate on input lists and generate 
@@
 Table~\ref{table:applications} summarizes the applications described
-with Boutiques and the main features used.
+with \boutiques and the main features used.
 
@@
 for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
-in CBRAIN through Boutiques. The integration uses the parallelization
+in CBRAIN through \boutiques. The integration uses the parallelization
 mechanism described in Section~\ref{sec:parallelization} so that even
@@
 \paragraph{Nipype.} The Nipype workflow engine~\cite{gorgolewski2011nipype} was
-instrumented to export to Boutiques applications described in its
+instrumented to export to \boutiques applications described in its
 internal Python description language. \todo{Update nipype\_cmd to
-  support the latest Boutiques schema and complete this paragraph.}
+  support the latest \boutiques schema and complete this paragraph.}
 
@@
 \paragraph{MICCAI challenges.}
-Boutiques was used to integrate 23 pipelines in the VIP platform in
+\boutiques was used to integrate 23 pipelines in the VIP platform in
 the context of two challenges organized by the MICCAI conference in
@@
 The pipelines did not use much of the expressiveness features offered
-by Boutiques since the instructions given to the challengers required
+by \boutiques since the instructions given to the challengers required
 that all the pipeline parameters were hardcoded, to facilitate
@@
 \paragraph{ndmg.} The NeuroData MRI to Graphs one-click connectome estimation
-pipeline~\cite{kiar2016ndmg}, developed in Python and leveraging FSL, was
+pipeline~\cite{kiar2017comprehensive}, developed in Python and leveraging FSL, was
 exported to \boutiques and is available at \url{https://github.com/neurodata/boutiques-tools}.
-
-\paragraph{Other applications.} Other neuroinformatics applications were ported to CBRAIN using Boutiques:
+The ndmg pipeline is deployed in CBRAIN via its \boutiques descriptor, and is available
+both through Docker and Singularity container environments.
+
+\paragraph{BIDS apps.} Brain Imaging Data Structure supported
+applications~\cite{gorgolewski2017bids}, a recent effort for the adoption of BIDS support
+in common neuroimaging pipelines, require a standardized set of input and output parameters
+by tools. A \boutiques tool has been developed which generates a manifest for any such
+BIDS app, enabling them to be deployed within a broad range of high performance computing
+infrastructures, such as CBRAIN.
+
+\paragraph{Other applications.} Other neuroinformatics applications were ported to CBRAIN using \boutiques:
 FSL tools and ICA AROMA are available at
@@
 
-\todo{Chris, something on BIDS apps?}
-
-\paragraph{VIP applications.}A few tools were imported in VIP using Boutiques.
-
+\paragraph{VIP applications.} Several tools were imported in VIP using \boutiques. 
 The first one is a phase contrast simulator called CreaPhase and developed at Creatis. 
@@
 The Odin MRI simulator\footnote{\url{http://od1n.sourceforge.net}} was also imported 
-in VIP using Boutiques. Since Odin  may require important amounts of computing resources 
+in VIP using \boutiques. Since Odin  may require important amounts of computing resources 
 (the common use-case consists in iterating on a large number of inputs files), Odin is 
@@
 A few other applications from the FLI (France Life Imaging) consortium were also imported in
-VIP using Boutiques. One of them is Ginfizz, a FMRI pipeline allowing to execute multiple 
+VIP using \boutiques. One of them is Ginfizz, a FMRI pipeline allowing to execute multiple 
 preprocessing steps and basic analyses such as: (i) realignment (motion registration and correction),
@@
 of functional and anatomical data to a match a template that has to be specified), and (iv)
-smoothing (spatial filtering to increase Signal to Noise ratio).
-The use of Docker containers (\url{https://hub.docker.com/r/fliiam/ginfizz} based on \url{https://hub.docker.com/r/miykael/nipype_level2})
-greatly simplifies its deployment. The animaN4BiasCorrection tool from the Anima  \footnote{\url{https://github.com/Inria-Visages/Anima-Public/wiki}} project was also made available 
+smoothing (spatial filtering to increase Signal to Noise ratio). The use of Docker containers
+(\url{https://hub.docker.com/r/fliiam/ginfizz} based on \url{https://hub.docker.com/r/miykael/nipype_level2}) 
+greatly simplifies its deployment. The animaN4BiasCorrection tool from the
+Anima\footnote{\url{https://github.com/Inria-Visages/Anima-Public/wiki}} project was also made available 
 in VIP. The import process was straightforward.
@@
     HCP PreFreesurfer                &                   &                   &                   &                   &                   &                   &                   \\
-    ndmg                &                   &       \cellcolor{gray!75}            &                   &          \cellcolor{gray!75}         &      \cellcolor{gray!75}             &                   &                   \\
+    ndmg                &                   &       \cellcolor{gray!75}            &                   &          \cellcolor{gray!75}         &      \cellcolor{gray!75}             & \cellcolor{gray!75} &                   \\
     MICCAI challenge (23 pipelines)\\&                   &                   &                   &                   &                   &                   & \\
@@
     \hline
-    Niak pre-processing             &\cellcolor{gray!75} &            &         &              &\cellcolor{gray!75}&\cellcolor{gray!75}\\
+    Niak pre-processing             &\cellcolor{gray!75} &\cellcolor{gray!75}&         &              &\cellcolor{gray!75}&\cellcolor{gray!75}\\
     ICA AROMA                       &\cellcolor{gray!75} &            &         &              &&\\
@@
     HCP PreFreesurfer               &\cellcolor{gray!75} &            &         &              &&\\
-    ndmg                            &\cellcolor{gray!75} &            &         &              &&\\
+    ndmg                            &\cellcolor{gray!75} &\cellcolor{gray!75}&         &              &&\\
     MICCAI challenge (23 pipelines) &\cellcolor{gray!75} &            &         &              &&\\
@@
 \end{table}
-\note{greg}{@tristan g: chris helped me verify that ndmg also works with singularity, but we have not shared
-the image anywhere. Would it be worth me publishing that image for this table?}
+
 \section{Discussion}
-
-With Boutiques, developers can port their applications once and execute
+With \boutiques, developers can port their applications once and execute
 them in several platforms. \boutiques removes the technological dependency to a
@@
 only (see Listing~\ref{listing:minimal}). For platforms, we aim at
-supporting as many features as possible in the local executor so that
+supporting as many features as possible in the executor so that
 only the following steps need to be implemented to support \boutiques
@@
 by end-users. However, it remains unclear how to validate that these
-applications actually execute as expected. A testing framework could
-be designed and potentially fed by existing frameworks, to address
-this issue.
+applications actually execute as expected beyond executing packaged tests
+distributed alongside a given tool. A \boutiques-specific testing
+framework could be designed and potentially fed by existing frameworks,
+to address this issue.
 
@@
 easier characterization of container contents, for instance through
-comparison digests with respect to validated base images.
+comparison digests with respect to validated base images. Singularity
+containers have reduced security risks as compared to Docker containers,
+but the issue of content transparency is still not avoided.
 
@@
 of the CWL and \boutiques Github repositories, CWL started 6 months
-before Boutiques (September 2014 vs. May 2015). The differences
+before \boutiques (September 2014 vs. May 2015). The differences
 highlighted below were identified from version 1.0 of the CWL Command
@@
 The following differences are conceptual in the sense that they may
-not be easily addressed by CWL or Boutiques developers without deeply
+not be easily addressed by CWL or \boutiques developers without deeply
 refactoring the frameworks.
@@
 validation, called SALAD (Semantic Annotations for Linked Avro Data)
-whereas Boutiques uses plain JSON schema. The main goal of SALAD is to
+whereas \boutiques uses plain JSON schema. The main goal of SALAD is to
 allow ``working with complex data structures and document formats, such
@@
 \item \boutiques supports various types of containers (Docker, Singularity,
-rootfs) while CWL supports only Docker. At the same time, CWL’s
-requirements are richer: for instance, they may include RAM and disk
-while Boutiques is limited to walltime estimate. CWL has hints, i.e.,
+rootfs) while CWL supports only Docker. Both tools have rich requirements:
+for instance, they may include RAM, disk usage, and walltime estimate; 
+whereas \boutiques also supports number of requested CPU cores. CWL has hints, i.e.,
 recommendations that only lead to warnings when not respected.

### 2017-10-13 paper.tex
Put @gkiar 2nd author

--- 2016-12-19 paper.tex
@@
 
-\author{Tristan Glatard, Tristan Aumentado-Armstrong, Natacha Beck, Pierre Bellec,\\
-        Remi Bernard, Sorina Camarasu-Pop, Fr\'ed\'eric Cervenansky, Samir Das, \\
-        Rafael Ferreira da Silva, Guillaume Flandin, John Flavin, Pascal Girard, \\
-        Krzysztof J. Gorgolewski, Charles G. Guttmann, Gregory Kiar, Nathaniel Kofalt, \\
-        Pierre-Olivier Quirion, Pierre Rioux, Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
+\author{Tristan Glatard, Gregory Kiar, Tristan Aumentado-Armstrong,
+  Natacha Beck, Pierre Bellec,\\ Remi Bernard, Sorina Camarasu-Pop,
+  Fr\'ed\'eric Cervenansky, Samir Das, \\ Rafael Ferreira da Silva,
+  Guillaume Flandin, John Flavin, Pascal Girard, \\ Krzysztof
+  J. Gorgolewski, Charles G. Guttmann, Val\'erie Hayot-Sasson,
+  Nathaniel Kofalt, \\ Pierre-Olivier Quirion, Pierre Rioux,
+  Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
 

### 2017-10-13 boutiquespaper.tex

@@
 
-\author{Tristan Glatard, Tristan Aumentado-Armstrong, Natacha Beck, Pierre Bellec,\\
-        Remi Bernard, Sorina Camarasu-Pop, Fr\'ed\'eric Cervenansky, Samir Das, \\
-        Rafael Ferreira da Silva, Guillaume Flandin, John Flavin, Pascal Girard, \\
-        Krzysztof J. Gorgolewski, Charles G. Guttmann, Gregory Kiar, Nathaniel Kofalt, \\
-        Pierre-Olivier Quirion, Pierre Rioux, Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
+\author{Tristan Glatard, Gregory Kiar, Tristan Aumentado-Armstrong,
+  Natacha Beck, Pierre Bellec,\\ Remi Bernard, Sorina Camarasu-Pop,
+  Fr\'ed\'eric Cervenansky, Samir Das, \\ Rafael Ferreira da Silva,
+  Guillaume Flandin, John Flavin, Pascal Girard, \\ Krzysztof
+  J. Gorgolewski, Charles G. Guttmann, Val\'erie Hayot-Sasson,
+  Nathaniel Kofalt, \\ Pierre-Olivier Quirion, Pierre Rioux,
+  Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
+
+\note{greg}{gk's affiliation: Department of Biomedical Engineering, Johns Hopkins University, Baltimore, MD}
+
 \maketitle
@@
   environments such as Docker and Singularity to install applications
-  across platforms in a lightweight manner. Boutiques is currently used
+  across platforms in a lightweight manner. \boutiques is currently used
   by three distinct virtual research platforms, and has been used to

### 2017-10-19 boutiquespaper.tex

@@
   Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
-
-\note{greg}{gk's affiliation: Department of Biomedical Engineering, Johns Hopkins University, Baltimore, MD}
 
@@
 if input \emph{B} is used; inputs may be passed in configuration files
-or directly on the command line. Such formal descriptions, also called
-\emph{manifests}, allow for advanced validation of input values to
-prevent errors. A manifest refers to a container where a given application
+or directly on the command line. Such formal descriptions, simply referred to
+as \emph{descriptors}, allow for advanced validation of input values to
+prevent errors. A descriptor refers to a container where a given application
 and all their dependencies are installed. \boutiques also provides a
@@
 facilitate the construction, validation, import, execution, and publishing
-of \boutiques manifests. \boutiques manifests are intented to be produced by
+of \boutiques descriptors. \boutiques descriptors are intented to be produced by
 scientific developers and stored alongside their application, indexed by common
@@
 
-Applications are described with a JSON manifest that specifies the
-command-line template, inputs and outputs. The manifest may point to a
+Applications are described with a JSON descriptor that specifies the
+command-line template, inputs and outputs. The descriptor may point to a
 container where the application and all its dependencies are
@@
 validation (this will be created at runtime if it is not found). At
-runtime, the execution platform builds the command line from the manifest
+runtime, the execution platform builds the command line from the descriptor
 and values entered by the user. The platform runs the command line on the
@@
 
-The core component of the manifest is a flexible command-line
+The core component of the descriptor is a flexible command-line
 description represented as a simple string template complying to the
@@
 Algorithm~\ref{algo:command-line} shows how the command line is
-constructed from the manifest and the values entered by the user. It
+constructed from the descriptor and the values entered by the user. It
 substitutes all the value keys in the command line, output path
@@
 automatically by the \boutiques command-line utility. The invocation
-schema is stored as an optional property of the \boutiques manifest.
+schema is stored as an optional property of the \boutiques descriptor.
 
@@
 can be executed using any engine provided that it is installed in the
-container image referred in the application manifest. Such a ``task
+container image referred in the application descriptor. Such a ``task
 encapsulation'' model allows for a scalable and reliable execution of
@@
 formats to convert container images across frameworks. For instance, a
-platform dedicated to high-performance computing may accept manifests
+platform dedicated to high-performance computing may accept descriptors
 referring to Docker containers to facilitate application integration
@@
 
-%%Containers are not mandatory though. If an application manifest does
+%%Containers are not mandatory though. If an application descriptor does
 %%not contain any reference to a container, the platform will assume
@@
 \paragraph{Validator.} The \boutiques validator checks conformance of JSON
-manifests to the \boutiques schema using a basic JSON validator. It also
+descriptors to the \boutiques schema using a basic JSON validator. It also
 performs the following checks that cannot be easily implemented in JSON schema:
@@
 \emph{simulate}. The simulate mode can generate hypothetical command lines from
-random values given the manifest (and corresponding invocation schema) for debugging
+random values given the descriptor (and corresponding invocation schema) for debugging
 purposes, or display the command that would be executed given a provided valid
-invocation. The launch mode can execute command line from a \boutiques manifest and
+invocation. The launch mode can execute command line from a \boutiques descriptor and
 a set of input values represented in JSON file complying with the invocation schema.
@@
 \paragraph{Invocation Schema Handler.} The invocation schema handler can create an
-invocation schema (see Section~\ref{sec:invocation-schema}) from a \boutiques manifest
+invocation schema (see Section~\ref{sec:invocation-schema}) from a \boutiques descriptor
 and validate input data against it using a regular JSON validator. It can be used to add
-invocation schemas to existing manifests. It is used by the executor if no invocation
-schema is present in the \boutiques manifest being deployed.
-
-\paragraph{Importer.} The importer takes \boutiques manifests from older versions and
+invocation schemas to existing descriptors. It is used by the executor if no invocation
+schema is present in the \boutiques descriptor being deployed.
+
+\paragraph{Importer.} The importer takes \boutiques descriptors from older versions and
 updates them to be compliant with the most recent version of the schema. This tool can
-also create manifests from select application collections, such as BIDS
+also create descriptors from select application collections, such as BIDS
 apps~\cite{gorgolewski2017bids}.
@@
 to facilitate new users getting started with \boutiques.
-
-\subsection{Repository}
-%% gk: I think we should remove this and just have the reference in Publisher
-\todo{Gunnar, Nathaniel, John}
 
@@
 resources. \boutiques is supported through a set of templates that
-generate such classes from the application manifest. Two application
+generate such classes from the application descriptor. Two application
 integration modes are available:
 \begin{enumerate}
-  \item The manifest is stored in a CBRAIN plugin and the Ruby classes
+  \item The descriptor is stored in a CBRAIN plugin and the Ruby classes
     are generated on-the-fly when CBRAIN starts. This mode allows
@@
     customization beyond the \boutiques schema.
-  \item Ruby classes are generated from manifests
+  \item Ruby classes are generated from descriptors
     through an offline process and integrated in CBRAIN as any other
@@
     applications are difficult to maintain in the long term, in
-    particular when the manifests are updated.
+    particular when the descriptors are updated.
 \end{enumerate}
-%% gk: I think we can skip this last bit, since there is only one type of
-%% \emph{boutiques} integration with cbrain, we needn't mention the other?
 
 A custom property (\texttt{cbrain:inherits-from-class}) was added to
-the \boutiques manifest to define the Ruby class that should be used as
+the \boutiques descriptor to define the Ruby class that should be used as
 parent class for the application in CBRAIN. It provides more
@@
 
-Table~\ref{table:applications} summarizes the applications described
-with \boutiques and the main features used.
+%%Table~\ref{table:applications} summarizes the applications described
+%%with \boutiques and the main features used.
 
@@
 and \texttt{vip:miccai-challenge-team-name}) were added to the
-\boutiques manifest to help post-process results in the specific
+\boutiques descriptor to help post-process results in the specific
 context of MICCAI challenges.
@@
 in common neuroimaging pipelines, require a standardized set of input and output parameters
-by tools. A \boutiques tool has been developed which generates a manifest for any such
+by tools. A \boutiques tool has been developed which generates a descriptor for any such
 BIDS app, enabling them to be deployed within a broad range of high performance computing
@@
 
-\begin{table}
-  \begin{tabular}{l|ccccccc}
-    \rowcolor[gray]{0.9}
-    Application                      & \multicolumn{7}{c}{Input/Output properties} \\
-    \rowcolor[gray]{0.9}
-                                     & Dependencies      &Groups             & Lists             & Optional          &Default            & Type restriction             & Min/max\\
-
-    \hline
-    Niak pre-processing              &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\\
-    ICA AROMA                        &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
-    FSL tools &&&&&&\\
-    \multicolumn{1}{r|}{anat}        &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
-    \multicolumn{1}{r|}{bet}         &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}\\      
-    \multicolumn{1}{r|}{fast}        &\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
-    \multicolumn{1}{r|}{first}       &                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   \\
-    FSL Nipype interfaces     \\
-    HCP PreFreesurfer                &                   &                   &                   &                   &                   &                   &                   \\
-    ndmg                &                   &       \cellcolor{gray!75}            &                   &          \cellcolor{gray!75}         &      \cellcolor{gray!75}             & \cellcolor{gray!75} &                   \\
-    MICCAI challenge (23 pipelines)\\&                   &                   &                   &                   &                   &                   & \\
-    \end{tabular}
-  \vspace*{0.5cm}
-  \begin{tabular}{l|ccc|ccc}
-    \hline
-    \rowcolor[gray]{0.9}
-    Application                     &\multicolumn{3}{c|}{Containers}            &  \multicolumn{3}{c}{Other properties}  \\
-    \rowcolor[gray]{0.9}
-                                    &Docker              & Singularity& Rootfs  & Config. file & Env vars  & Parallelization   \\
-    \hline
-    Niak pre-processing             &\cellcolor{gray!75} &\cellcolor{gray!75}&         &              &\cellcolor{gray!75}&\cellcolor{gray!75}\\
-    ICA AROMA                       &\cellcolor{gray!75} &            &         &              &&\\
-    FSL tools &&&&&\\
-    \multicolumn{1}{r|}{anat}       &                    &            &         &              &&\\
-    \multicolumn{1}{r|}{bet}        &                    &            &         &              &&\\
-    \multicolumn{1}{r|}{fast}       &                    &            &         &              &&\\
-    \multicolumn{1}{r|}{first}      &                    &            &         &              &&\\
-    FSL Nipype interfaces           &                    &            &         &              &&\\
-    HCP PreFreesurfer               &\cellcolor{gray!75} &            &         &              &&\\
-    ndmg                            &\cellcolor{gray!75} &\cellcolor{gray!75}&         &              &&\\
-    MICCAI challenge (23 pipelines) &\cellcolor{gray!75} &            &         &              &&\\
-  \end{tabular} 
-  \caption{Applications available in Boutiques with manifest features used.}
-  \label{table:applications}
-\end{table}
+%%\begin{table}
+%%  \begin{tabular}{l|ccccccc}
+%%    \rowcolor[gray]{0.9}
+%%    Application                      & \multicolumn{7}{c}{Input/Output properties} \\
+%%    \rowcolor[gray]{0.9}
+%%                                     & Dependencies      &Groups             & Lists             & Optional          &Default            & Type restriction             & Min/max\\
+%%    
+%%    \hline
+%%    Niak pre-processing              &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\\
+%%    ICA AROMA                        &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
+%%    FSL tools &&&&&&\\
+%%    \multicolumn{1}{r|}{anat}        &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
+%%    \multicolumn{1}{r|}{bet}         &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}\\      
+%%    \multicolumn{1}{r|}{fast}        &\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
+%%    \multicolumn{1}{r|}{first}       &                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   \\
+%%    FSL Nipype interfaces     \\
+%%    HCP PreFreesurfer                &                   &                   &                   &                   &                   &                   &                   \\
+%%    ndmg                &                   &       \cellcolor{gray!75}            &                   &          \cellcolor{gray!75}         &      \cellcolor{gray!75}             & \cellcolor{gray!75} &                   \\
+%%    MICCAI challenge (23 pipelines)\\&                   &                   &                   &                   &                   &                   & \\
+%%    \end{tabular}
+%%  \vspace*{0.5cm}
+%%  \begin{tabular}{l|ccc|ccc}
+%%    \hline
+%%    \rowcolor[gray]{0.9}
+%%    Application                     &\multicolumn{3}{c|}{Containers}            &  \multicolumn{3}{c}{Other properties}  \\
+%%    \rowcolor[gray]{0.9}
+%%                                    &Docker              & Singularity& Rootfs  & Config. file & Env vars  & Parallelization   \\
+%%    \hline
+%%    Niak pre-processing             &\cellcolor{gray!75} &\cellcolor{gray!75}&         &              &\cellcolor{gray!75}&\cellcolor{gray!75}\\
+%%    ICA AROMA                       &\cellcolor{gray!75} &            &         &              &&\\
+%%    FSL tools &&&&&\\
+%%    \multicolumn{1}{r|}{anat}       &                    &            &         &              &&\\
+%%    \multicolumn{1}{r|}{bet}        &                    &            &         &              &&\\
+%%    \multicolumn{1}{r|}{fast}       &                    &            &         &              &&\\
+%%    \multicolumn{1}{r|}{first}      &                    &            &         &              &&\\
+%%    FSL Nipype interfaces           &                    &            &         &              &&\\
+%%    HCP PreFreesurfer               &\cellcolor{gray!75} &            &         &              &&\\
+%%    ndmg                            &\cellcolor{gray!75} &\cellcolor{gray!75}&         &              &&\\
+%%    MICCAI challenge (23 pipelines) &\cellcolor{gray!75} &            &         &              &&\\
+%%  \end{tabular} 
+%%  \caption{Applications available in Boutiques with descriptor features used.}
+%%  \label{table:applications}
+%%\end{table}
 
@@
 
-\subsection{\boutiques manifest}
-
-The \boutiques manifest specification allows describing a wide range
+\subsection{\boutiques descriptor}
+
+The \boutiques descriptor specification allows describing a wide range
 of applications, but it is also getting increasingly complex through
 additions such as invocation schemas and dependencies among
-inputs. Extending the \boutiques manifest has two main goals. First,
+inputs. Extending the \boutiques descriptor has two main goals. First,
 accurately describing applications improves validation: incorrect input
 values and execution results are more precisely detected when the
-application manifest is comprehensive, which is essential to improve
+application descriptor is comprehensive, which is essential to improve
 the reliability and user-friendliness of execution
-platforms. Secondly, a rich manifest schema reduces the need for
+platforms. Secondly, a rich descriptor schema reduces the need for
 custom wrappers to integrate applications as even convoluted command
@@
 
-Nonetheless, a complex manifest schema has a cost for application
+Nonetheless, a complex descriptor schema has a cost for application
 developers and platforms, which we address as follows. For developers,
-we maintain the set of mandatory manifest properties as small as
+we maintain the set of mandatory descriptor properties as small as
 possible so that simple applications can be described in a few lines
@@
 only the following steps need to be implemented to support \boutiques
-in a platform, regardless of the complexity of the manifest:
+in a platform, regardless of the complexity of the descriptor:
 \begin{enumerate}
@@
     executor and input JSON object.
-  \item Output delivery: from the manifest, identify the output files
+  \item Output delivery: from the descriptor, identify the output files
     and deliver them to the user.
@@
 at step 2, without requiring the platform to support the related
-manifest properties.
+descriptor properties.
 \begin{listing}
@@
 \end{minted}
-\caption{A minimal \boutiques manifest.} 
+\caption{A minimal \boutiques descriptor.} 
 \label{listing:minimal}
@@
 including high-performance computing platforms and individual laptops,
-to reproduce analyses. The \boutiques manifest describes the
+to reproduce analyses. The \boutiques descriptor describes the
 parameters and implementation of the application, and the invocation
@@
 the execution platform without being specified in the \boutiques
-manifest. Such runtime parameters may impact reproducibility in some
-cases. To properly cover this issue, \boutiques manifests should be
+descriptor. Such runtime parameters may impact reproducibility in some
+cases. To properly cover this issue, \boutiques descriptors should be
 complemented by a provenance framework that captures a detailed trace
@@
 provide input through a GUI. This should, however, be specified
-as an extension to the manifest since most platforms would not support
+as an extension to the descriptor since most platforms would not support
 this feature by default. Graphical output produced by applications
@@
 The Command Line Tool Description of the Common Workflow Language
-(CWL) overlaps with the \boutiques manifest. According to the history
+(CWL) overlaps with the \boutiques descriptor. According to the history
 of the CWL and \boutiques Github repositories, CWL started 6 months

### 2017-10-30 boutiquespaper.tex
Worked on introduction

@@
   application description format which relies on Linux container
-  environments such as Docker and Singularity to install applications
+  environments Docker and Singularity to install applications
   across platforms in a lightweight manner. \boutiques is currently used
@@
 
-Platforms such as web services, portals, and virtual research
-environments are becoming an common solution for increasing the
-accessibility of high performance computing. These scientific gateways and
-workflow engines often rely on third-party data processing and simulation
-applications to deliver services to their users. Such applications are
-often repeatedly and manually ported to various platforms whereas this
-porting effort would greatly benefit from being mutualized. Meanwhile,
-virtual container systems such as Docker and Singularity greatly facilitate
-the sharing of software and improve the reproducibility of analyses by defining
-immutable, reusable execution environments. However, containers need
-to be properly annotated to allow platforms to leverage the software
-they contain. The command lines that can be run in a container must be formally
-described so that they can be automatically created and executed by platforms.
-Describing command lines is not straightforward since they often contain
-implicit semantics specified in human-readable documents such as manuals.
+Platforms such as web services, portals, scientific gateways, workflow
+engines and virtual research environments commonly integrate
+third-party applications to facilitate the processing of large
+datasets on high-performance computing resources. Such applications,
+however, are often repeatedly and manually ported to various platforms
+whereas this porting effort would greatly benefit from being
+mutualized. Meanwhile, virtual container systems such as Docker and
+Singularity have emerged to facilitate the sharing of software and
+improve the reproducibility of analyses by defining immutable,
+reusable execution environments. Nevertheless, the automatic
+integration of containerized applications in platforms require formal
+annotations of container images, to describe the available command lines.
 
@@
 template comprising the inputs it requires and the outputs it
-produces. Inputs may be inter-dependent, e.g., input \emph{A} cannot be used
-if input \emph{B} is used; inputs may be passed in configuration files
-or directly on the command line. Such formal descriptions, simply referred to
-as \emph{descriptors}, allow for advanced validation of input values to
-prevent errors. A descriptor refers to a container where a given application
-and all their dependencies are installed. \boutiques also provides a
-simple parallelization framework that allows applications such as
-workflow engines to concurrently execute multiple tasks (in the
-remainder, tasks are called \emph{invocations}). A set of core tools
-facilitate the construction, validation, import, execution, and publishing
-of \boutiques descriptors. \boutiques descriptors are intented to be produced by
-scientific developers and stored alongside their application, indexed by common
-repositories, and to be consumed by execution platforms.
+produces. Inputs may be inter-dependent and passed in configuration
+files or directly on the command line. Such formal descriptions,
+simply referred to as \emph{descriptors}, link to a container image
+where the given application and all its dependencies are
+installed. \boutiques descriptors allow for automatic application
+integration in platforms, and advanced validation of input values to
+prevent errors. \boutiques also provides a simple parallelization
+framework that allows parallel applications to concurrently execute
+multiple tasks (in the remainder, tasks are called
+\emph{invocations}). \boutiques descriptors are intented to be
+produced by scientific application developers, stored alongside their
+application, indexed by common repositories, and consumed by
+execution platforms.  A set of core tools facilitate the construction,
+validation, import, execution, and publishing of \boutiques
+descriptors.
 

### 2017-10-30 boutiquespaper.tex
Updates in section 2

@@
 whereas this porting effort would greatly benefit from being
-mutualized. Meanwhile, virtual container systems such as Docker and
-Singularity have emerged to facilitate the sharing of software and
+mutualized. Meanwhile, virtual container systems such as Docker\todo{cite} and
+Singularity\todo{cite} have emerged to facilitate the sharing of software and
 improve the reproducibility of analyses by defining immutable,
@@
 
-Applications are described with a JSON descriptor that specifies the
-command-line template, inputs and outputs. The descriptor may point to a
-container where the application and all its dependencies are
-installed. It may also contain an invocation schema used for input
-validation (this will be created at runtime if it is not found). At
-runtime, the execution platform builds the command line from the descriptor
-and values entered by the user. The platform runs the command line on the
-execution infrastructure, e.g., a server, a cluster or a cloud, within a
-container whenever available. To build and run the command line, the
-platform may rely on the \boutiques core tools, packaged through the 
-\texttt{bosh} command-line utility, such as the validator and executor.
+In \boutiques, applications are described with a JSON descriptor that
+specifies the command-line template, inputs and outputs. The
+descriptor may point to a container where the application and all its
+dependencies are installed. It may also contain an invocation schema
+used for input validation (this will be created at runtime if it is
+not found). At runtime, the execution platform builds the command line
+from the descriptor and values entered by the user. The platform runs
+the command line on the execution infrastructure, e.g., a server, a
+cluster or a cloud, within a container whenever available. To build
+and run the command line, the platform may rely on the \boutiques core
+tools, in particular the validator and executor, packaged through the
+\texttt{bosh} command-line utility.
 
@@
 description represented as a simple string template complying to the
-syntax of the \texttt{bash} UNIX shell. \texttt{bash} is the default
-shell on the major Linux distributions and OS X. The command-line
-template may contain placeholders for input and output values, called
-value keys. The command line may actually encompass several commands
-separated by semicolons, pipes (\texttt{|}) or ampersands (\texttt{\&}).
-Such a flexible command-line specification is meant to facilitate the
-embedding of minor mundane operations on the command line, for instance
-directory creation, input decompression or output archival.
+syntax of the \texttt{bash} UNIX shell, the default shell on the major
+Linux distributions and OS X. The command-line template may contain
+placeholders for input and output values, called value keys. It may
+also encompass several commands separated by \texttt{bash} constructs
+such as semicolons, pipes (\texttt{|}) or ampersands (\texttt{\&}), to
+facilitate the embedding of mundane operations on the command line,
+for instance directory creation, input decompression or output
+archival.
 
@@
 \end{verbatim}
-The template contains six value keys for input and output values that
-will be replaced by values and file names according to the user input
-when the application is executed. Flags will also be added wherever
-appropriate, with customizable separators. No particular syntax is
-imposed for the value keys, they just have to be unique.  Note the use
-of the pipe (\texttt{|}) operator to chain applications, and of the
-\texttt{>>} operator to redirect the standard output to a file.
+The template contains five value keys, identified by square brackets,
+that will be replaced by values and file names according to the user
+input when the application is executed. Flags will also be added
+wherever appropriate, with customizable separators. Value keys have to
+be unique but do not have to comply to any particular syntax. Note the
+use of the pipe (\texttt{|}) operator to chain applications, and of
+the \texttt{>>} operator to redirect the standard output to a file.
 
@@
 Types can be restricted to a specific set of values (\texttt{String} 
-and \texttt{Number} only), to a specific range (\texttt{Number} only), to
-integers or to booleans (\texttt{Number} only).
+and \texttt{Number} only), to a specific range (\texttt{Number} only), or to
+integers (\texttt{Number} only).
 
@@
 with an identifier, name, and list of input identifiers. Groups may be
-used to improve the presentation in a graphical user interface and to
+used to improve the presentation in a graphical user interface, and to
 specify the following constraints among inputs:
-(1)~\texttt{mutually-exclusive}: only one member in the group may have a
-value; (2)~\texttt{one-is-required}: at least one member in the group
-must have a value; (3)~\texttt{all-or-none}: if any of the members are
-present all members must be present. Dependencies among inputs may also be
-defined regardless of a particular group: an input may (1)~require a list of
-inputs and (2)~disable a list of inputs.
+(1)~\texttt{mutually-exclusive}: only one member in the group may have
+a value; (2)~\texttt{one-is-required}: at least one member in the
+group must have a value; (3)~\texttt{all-or-none}: if any of the
+members are present all members must be present. Dependencies among
+inputs may also be defined regardless of a particular group: an input
+may (1)~require a list of inputs and (2)~disable a list of inputs.
 
@@
 transferred and delivered to the user once the execution is
-complete. Although it may not be required to distinguish inputs from
-outputs to build the application command line, this information is
-needed by computing platforms to identify the files that must be saved
-after the execution.
+complete. Outputs need to be specified so that computing platforms can
+identify the files that must be saved after the execution and raise
+errors if they are not present.
 
@@
 definition and assuming that the string input value entered by the
-user is "foo.csv", the string \texttt{[LOG]} will be
+user is ``foo.csv'', the string \texttt{[LOG]} will be
 replaced by \texttt{log-foo} on the command line at execution time.
@@
 A large number of applications rely on configuration files rather than
-command-line options to define their input and output
-parameters. Indeed, as the number of parameters increases, command
-lines rapidly become long and cumbersome whereas configuration files
-allow for better structure and documentation.
+command-line options to define their input and output parameters. As
+the number of parameters increases, command lines rapidly become long
+and cumbersome whereas configuration files allow for better structure
+and documentation.
 
 Configuration files may be complex though, and specified in any
-language.  As it does not seem reasonable to adopt a particular
-template for configuration files, the \boutiques specification allows
-application developers to specify their own template containing input
-and output value keys. Configuration files are specific types of
-output files that must have a file template in addition to a path
-template that defines how they will be named and where they will be
-written. They may also have a value key and a flag in case they need
-to be passed on the command
+language.  The \boutiques specification allows application developers
+to specify their own template containing input and output value
+keys. Configuration files are specific types of output files that must
+have a file template in addition to a path template that defines how
+they will be named and where they will be written. They may also have
+a value key and a flag in case they need to be passed on the command
 line. Listing~\ref{listing:configuration-file-example} shows an
@@
         "stringInput=foo",
+        "# This is an input file",
         "fileInput=[FILE-INPUT]",
@@
 Rigorous input validation is an important motivation for
-\boutiques. The obvious solution for JSON documents is to rely on an
-application-specific JSON schema, called \emph{invocation schema}, to
+\boutiques. For this purpose, \boutiques relies on an
+application-specific JSON schema\todo{cite}, called \emph{invocation schema}, to
 specify the input values accepted by an application. Platforms can
@@
 example of how dependencies between mutually exclusive parameters are
-defined in the invocation schema. As it does not seem realistic to
-assume that developers will easily write correct invocations schemas
-even for their own applications, invocation schemas can be generated
-automatically by the \boutiques command-line utility. The invocation
-schema is stored as an optional property of the \boutiques descriptor.
+defined in the invocation schema. To relieve application developers
+from the burden of having to write invocation schemas, invocation
+schemas can be generated automatically by the \texttt{bosh}
+command-line utility. The invocation schema is stored as an optional
+property of the \boutiques descriptor.
 
@@
 workflows expressed in a variety of languages, as detailed in
-~\cite{glatard2016fgcs}.
+~\cite{glatard2017software}.
 
@@
 
-Container images are defined from their URL (rootfs) or name in the
-Docker or Singularity indices. Manifests and container images may specify a
-working directory where the application has to be run, and a hash may also be
-reported to accurately identify container images and detect updates.
+Container images are defined from their URL (rootfs) or image name in
+a Docker or Singularity index. Descriptors and container images may
+specify a working directory where the application has to be run,
+descriptors may contain a boolean indicating if the image has an entry
+point, and a hash may also be reported to accurately identify
+container images and detect updates.
 
@@
 
-\boutiques is available on Python through the PyPi package repository as
-``\texttt{boutiques}''. The \boutiques package exposes a command-line
-utility to the user, \texttt{bosh}, which contains entrypoints for all
-core functions within \boutiques. There is help-text which should guide
-users through interacting with the \texttt{bosh} utility. The core tools
-provided by \boutiques are, the: validator, executor (for both launch and
-simulation), invocation schema handler, importer, and publisher. The tools
-exposed through the command-line API are also available consistently through
-the Python interface by importing the \boutiques package.
+\boutiques is available on Python through the PyPi package repository
+as ``\texttt{boutiques}''. The \boutiques package exposes a
+command-line utility to the user, \texttt{bosh}, which contains
+entrypoints for all core functions within \boutiques. There is
+help-text which should guide users through interacting with the
+\texttt{bosh} utility. The core tools provided by \boutiques are, the:
+validator, executor (for both launch and simulation), invocation
+schema handler, importer, and publisher. The tools exposed through the
+command-line API are also available consistently through the Python
+interface by importing the \boutiques package. Though not a component
+of the \boutiques tool chain, a Jupyter Notebook tutorial exists to
+facilitate new users getting started with \boutiques.
 
 \paragraph{Validator.} The \boutiques validator checks conformance of JSON
-descriptors to the \boutiques schema using a basic JSON validator. It also
-performs the following checks that cannot be easily implemented in JSON schema:
-value keys are unique among inputs, input and output identifiers are
-unique, input and output value keys are all included in the command
-line, identifiers with the same value key are mutually exclusive, value keys
-are not contained within each other (which would puzzle substitution), output
-path templates are unique (to avoid results to overwrite each other), inputs of
-type Flag have a command-line flag, are optional and are not lists, the default
-value of restricted types is part of the restriction, an input cannot both require
-and disable another input, required inputs cannot require or disable
-other parameters, group member identifiers must correspond to existing
-inputs and cannot appear in different groups, mutually exclusive
-groups cannot have members requiring other members, one-is-required groups
-should never have required members, and all-or-none group members must not be
-required.
+descriptors to the \boutiques schema using a basic JSON validator. It
+also performs the following checks that cannot be easily implemented
+in JSON schema: value keys are unique among inputs, input and output
+identifiers are unique, input and output value keys are all included
+in the command line, identifiers with the same value key are mutually
+exclusive, value keys are not contained within each other (which would
+puzzle substitution), output path templates are unique (to avoid
+results to overwrite each other), inputs of type Flag have a
+command-line flag, are optional and are not lists, the default value
+of restricted types is part of the restriction, an input cannot both
+require and disable another input, required inputs cannot require or
+disable other parameters, group member identifiers must correspond to
+existing inputs and cannot appear in different groups, mutually
+exclusive groups cannot have members requiring other members,
+one-is-required groups should never have required members, and
+all-or-none group members must not be required.
 
@@
 updates them to be compliant with the most recent version of the schema. This tool can
-also create descriptors from select application collections, such as BIDS
+also create descriptors from selected application collections, such as BIDS
 apps~\cite{gorgolewski2017bids}.
@@
 new repositories for other domains, such as Bioconductor for bioinformatics.
-
-Though not a component of the \boutiques tool chain, a Jupyter Notebook tutorial exists
-to facilitate new users getting started with \boutiques.
 
@@
 
+\todo{BIDS apps}
+
 The Common Workflow Language\footnote{\url{http://www.commonwl.org}}

### 2017-11-01 boutiquespaper.tex
Added resource requirements

@@
 
-\notimplementedyet{The invocation schema also has a
- \texttt{task-dependencies} array that allows specifying that the new
- invocation may start only when previously-submitted invocations have
- completed. This property is used in the parallelization model
- described hereafter.}
+%% \notimplementedyet{The invocation schema also has a
+%%  \texttt{task-dependencies} array that allows specifying that the new
+%%  invocation may start only when previously-submitted invocations have
+%%  completed. This property is used in the parallelization model
+%%  described hereafter.}
 
@@
 
-\subsection{Parallelization support}
-\label{sec:parallelization}
-\boutiques supports parallelism by allowing applications to submit new
-invocations through \notimplementedyet{the following protocol}:
-\begin{enumerate}
-\item Applications write a JSON object complying to the invocation
-  schema in their execution directory.
-\item The platform reads the JSON object, creates the corresponding
-  invocation, and writes back the invocation id in the same directory.
-% The platform could also support status requests.
-\end{enumerate}
-Several invocations can be submitted at once, possibly using the
-\texttt{task-dependencies} array to specify dependencies among
-them. This mechanism is implemented in the Boutiques schema using a
-single property, \texttt{can-submit-tasks}.
-
-Although very simple, this model allows wrapping complete
-workflows. Workflows are wrapped as any other application, and they
-can be executed using any engine provided that it is installed in the
-container image referred in the application descriptor. Such a ``task
+%% \subsection{Parallelization support}
+%% \label{sec:parallelization}
+%% \boutiques supports parallelism by allowing applications to submit new
+%% invocations through \notimplementedyet{the following protocol}:
+%% \begin{enumerate}
+%% \item Applications write a JSON object complying to the invocation
+%%   schema in their execution directory.
+%% \item The platform reads the JSON object, creates the corresponding
+%%   invocation, and writes back the invocation id in the same directory.
+%% % The platform could also support status requests.
+%% \end{enumerate}
+%% Several invocations can be submitted at once, possibly using the
+%% \texttt{task-dependencies} array to specify dependencies among
+%% them. This mechanism is implemented in the Boutiques schema using a
+%% single property, \texttt{can-submit-tasks}.
+
+%% Although very simple, this model allows wrapping complete
+%% workflows. Workflows are wrapped as any other application, and they
+%% can be executed using any engine provided that it is installed in the
+%% container image referred in the application descriptor.
+
+\subsection{Workflow support}
+
+\boutiques does not specify a particular language to compose
+descriptors in a workflow, due to the large amount of specialized
+frameworks to do so. However, workflows can be both composed from and
+described as \boutiques descriptors: workflow engines can leverage the
+\texttt{bosh} tools to call \boutiques tools from their descriptors;
+in turn, workflows can be described as \boutiques descriptors provided
+that the execution engine is containerized. Such a ``task
 encapsulation'' model allows for a scalable and reliable execution of
@@
 %% gk: I believe we only support containers now, correct?
+
+\subsection{Resource requirements}
+
+\boutiques descriptors may contain requirements regarding the number
+of CPU cores or nodes, the amount of RAM or disk storage, and the
+total walltime expected for a typical execution of the
+application. Such properties are called ``suggested resources'' as we
+are well aware that the actual resource requirements usually depend on
+the input data, parameters and hardware infrastructure.
 

### 2017-11-01 boutiquespaper.tex
Use GigaScience template

@@
-\documentclass{article}
+% GigaScience template
+\documentclass[a4paper,num-refs]{oup-contemporary}
+
+%%% Journal toggle; only specific options recognised.
+%%% (Only "gigascience" and "general" are implemented now. Support for other journals is planned.)
+\journal{gigascience}
+
+\usepackage{graphicx}
+\usepackage{siunitx}
+\papercat{Paper}
 
@@
 \usepackage{pdfcomment} % Used for notes and todos
-\usepackage[margin=1in]{geometry} % Increases margins
-\usepackage[table]{xcolor} % Color definitions
 \usepackage{algorithm} % Algorithm float
@@
 \author{Tristan Glatard, Gregory Kiar, Tristan Aumentado-Armstrong,
-  Natacha Beck, Pierre Bellec,\\ Remi Bernard, Sorina Camarasu-Pop,
-  Fr\'ed\'eric Cervenansky, Samir Das, \\ Rafael Ferreira da Silva,
-  Guillaume Flandin, John Flavin, Pascal Girard, \\ Krzysztof
+  Natacha Beck, Pierre Bellec, Remi Bernard, Sorina Camarasu-Pop,
+  Fr\'ed\'eric Cervenansky, Samir Das,  Rafael Ferreira da Silva,
+  Guillaume Flandin, John Flavin, Pascal Girard, Krzysztof
   J. Gorgolewski, Charles G. Guttmann, Val\'erie Hayot-Sasson,
-  Nathaniel Kofalt, \\ Pierre-Olivier Quirion, Pierre Rioux,
+  Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux,
   Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
 
+\begin{frontmatter}
 \maketitle
 
-\abstract{Porting applications to execution platforms such as web
+\begin{abstract}Porting applications to execution platforms such as web
   portals and workflow engines is critical to enable their sharing
@@
   by three distinct virtual research platforms, and has been used to
-  describe dozens of neuroinformatics tools.}
+  describe dozens of neuroinformatics tools.
+\end{abstract}
+
+\begin{keywords}
+Keyword1; keyword 2; keyword 3 (Three to ten keywords representing the main content of the article)
+\end{keywords}
+\end{frontmatter}
 
@@
 \label{sec:results}
+
+\boutiques has been implemented in several platforms and used to
+describe dozens of tools, as described hereafter.
 
@@
 
-\bibliographystyle{plain}
 \bibliography{biblio}

### 2017-11-01 boutiquespaper.tex
Classified ported applications

@@
 
+\todo{Intro on the relevance of pipeline sharing for reproducible and open science.}
+
+\todo{Check passive voice everywhere.}
+
 Platforms such as web services, portals, scientific gateways, workflow
@@
 
-\paragraph{General properties.} Inputs must have a name, a unique
+\paragraph{General properties} Inputs must have a name, a unique
 identifier, and a type. They may be optional, have a description, a
@@
 
-\paragraph{Types.} Inputs may be of type \texttt{String}, \texttt{Number},
+\paragraph{Types} Inputs may be of type \texttt{String}, \texttt{Number},
 \texttt{Flag} or \texttt{File}. \texttt{File} may also represent a directory.
@@
 
-\paragraph{Groups and dependencies.} Groups of inputs may be defined
+\paragraph{Groups and dependencies} Groups of inputs may be defined
 with an identifier, name, and list of input identifiers. Groups may be
@@
 
-\paragraph{Validator.} The \boutiques validator checks conformance of JSON
+\paragraph{Validator} The \boutiques validator checks conformance of JSON
 descriptors to the \boutiques schema using a basic JSON validator. It
@@
 
-\paragraph{Executor.} The executor has two modes of operation: \emph{launch}, and
+\paragraph{Executor} The executor has two modes of operation: \emph{launch}, and
 \emph{simulate}. The simulate mode can generate hypothetical command lines from
@@
 
-\paragraph{Invocation Schema Handler.} The invocation schema handler can create an
+\paragraph{Invocation Schema Handler} The invocation schema handler can create an
 invocation schema (see Section~\ref{sec:invocation-schema}) from a \boutiques descriptor
@@
 
-\paragraph{Importer.} The importer takes \boutiques descriptors from older versions and
+\paragraph{Importer} The importer takes \boutiques descriptors from older versions and
 updates them to be compliant with the most recent version of the schema. This tool can
@@
 
-\paragraph{Publisher.} As \boutiques has primarily been adopted in the neuroinformatics
+\paragraph{Publisher} As \boutiques has primarily been adopted in the neuroinformatics
 community, the publisher gets a further description (such as author, website, etc.) of
 the described tool, and adds an index to it on
-NeuroLinks\footnote{\url{https://brainhack101.github.io/neurolinks/}}, a repository
+NeuroLinks\footnote{\url{https://brainhack101.github.io/neurolinks}}, a repository
 containing links to useful resources and tools. This enables others to find and use the
@@
 \label{sec:results}
-
-\boutiques has been implemented in several platforms and used to
-describe dozens of tools, as described hereafter.
 
@@
 \begin{enumerate}
-  \item The descriptor is stored in a CBRAIN plugin and the Ruby classes
-    are generated on-the-fly when CBRAIN starts. This mode allows
-    CBRAIN developers update all \boutiques applications at once by
-    editing the templates. However, it does not allow for
-    customization beyond the \boutiques schema.
+  \item The descriptor is stored in a CBRAIN plugin and the Ruby
+    classes are generated on-the-fly when CBRAIN starts. This mode
+    allows CBRAIN developers update all \boutiques applications at
+    once by editing the templates. However, it does not allow for
+    customization beyond the \boutiques schema. To provide more
+    flexibility, we added a custom property
+    (\texttt{cbrain:inherits-from-class}) to the \boutiques descriptor
+    to define the Ruby class that should be used as parent class for
+    the application.
   \item Ruby classes are generated from descriptors
@@
 
-A custom property (\texttt{cbrain:inherits-from-class}) was added to
-the \boutiques descriptor to define the Ruby class that should be used as
-parent class for the application in CBRAIN. It provides more
-flexibility to the first integration model described previously.
-
-CBRAIN was also extended to support the parallelization mechanism
-available in \boutiques: the CBRAIN execution agent periodically scans
-the execution directories of the parallelized applications and it submits
-the invocations defined in the JSON objects written by the
-application in this directory.
-
-A new list mechanism was also introduced in CBRAIN to facilitate the
+We also extended CBRAIN to enable the parallelization of workflows
+wrapped as \boutiques descriptors. Applications with the
+\texttt{cbrain:can-submit-new-tasks} custom property may submit sub-tasks by
+creating \boutiques invocations in their working directory. CBRAIN
+periodically scans working directories, submits the requested
+sub-tasks, and writes back the invocation id in the same
+directory. This parallelization model is simple, the application only
+needs to write \boutiques invocations and communication happens
+through the file system, and it is also powerful as it enables the
+parallelization of complex workflows such as the FSL and Niak ones
+described later.
+
+We also introduced a new list mechanism in CBRAIN to facilitate the
 iteration of \boutiques applications on large sets of files. CBRAIN
@@
 
-The Virtual Imaging Platform (VIP)~\cite{GLAT-13} is a web portal for medical 
-simulation  and image data analysis. VIP allows to make applications 
-available as a service, and also to benefit from the resources available 
-to the biomed Virtual Organisation (VO) of the European Grid Infrastructure 
-\footnote{\url{https://www.egi.eu/}}, supported by approximately 65 computing
-sites world-wide offering a total of 130 computing clusters and 5 PB of storage.
-The VIP service is currently deployed at the Creatis laboratory
-\footnote{\url{https://www.creatis.insa-lyon.fr/site7/en}} and uses the DIRAC 
-French service \footnote{\url{https://dirac.france-grilles.fr}} to submit jobs on 
-EGI grid and cloud resources.  VIP counts, in October 2017, more than 1000 
-registered users and a growing number of available applications.
+The Virtual Imaging Platform (VIP)~\cite{GLAT-13} is a web portal for
+medical simulation and image data analysis. VIP makes applications
+available as services, and connects them to the biomed Virtual
+Organization (VO) in the European Grid
+Infrastructure\footnote{\url{https://www.egi.eu}}. The biomed VO
+interconnects approximately 65 computing sites world-wide and provides
+access to 130 computing clusters and 5~PB of storage.  The VIP service
+is deployed at the Creatis
+laboratory\footnote{\url{https://www.creatis.insa-lyon.fr}} in Lyon
+and it uses the DIRAC French national
+service\footnote{\url{https://dirac.france-grilles.fr}} to execute
+jobs on EGI grid and cloud resources.  As of October 2017, VIP counts
+more than 1000 registered users and a growing number of available
+applications.
 %% gk: I guessed 50 more in the last year, since 1000 is a nice number, is this accurate?
 
-Until recently, applications were manually integrated in VIP as workflows written 
-in the Gwendia\cite{MONT-09} language and managed  with the MOTEUR\cite{GLAT-08} workflow engine. 
-As of today, Boutiques is supported through a "Boutiques application importer" tool, which 
-parses the JSON file and automatically generates the corresponding application workflow
-and the wrapper script that handles, among other things, the execution of the command line itself. 
-The use of an application workflow allows to (i) iterate on input lists and generate 
-parallel tasks, as well as to (ii) concatenate multiple tools. The applications used in 
-the MICCAI challenges, for example, were followed by other tools allowing to compute 
-challenge-specific metrics. Tool concatenation is currently handeled at the importer 
-level based on pre-defined workflow templates.
-
-\note{sorina}{@tristan: some info on workflows is available in the paragraph above. Is it OK/enough or do you have other suggestions?}
+Until recently, applications were manually integrated in VIP as
+workflows written in the Gwendia~\cite{MONT-09} language and executed
+with the MOTEUR~\cite{GLAT-08} engine.  As of today, Boutiques is
+supported through a "Boutiques application importer" tool, which
+parses the JSON descriptor and automatically generates the
+corresponding application workflow and the wrapper script that
+handles, among other things, the execution of the command line.
+Application workflows enable (1) iterations on input lists, (2) the
+generation of parallel tasks, and (3) the concatenation of multiple
+tools. For examples, the applications used in the MICCAI challenges
+described below required workflows to evaluate results using the
+metrics defined by the challenge. Tool concatenation is 
+handled at the importer level based on pre-defined workflow
+templates.
 
@@
 
-\paragraph{Niak.} The Niak pre-processing pipeline, executed with the Pipeline System
+Dozens of neuroinformatics tools were ported to CBRAIN or VIP using
+\boutiques. The main ones are described below. Several \boutiques
+descriptors were published in
+Neurolinks\footnote{\url{https://brainhack101.github.io/neurolinks}}
+through the \texttt{bosh} publisher.
+
+%% \paragraph{Nipype} We instrumented the Nipype workflow engine~\cite{gorgolewski2011nipype} 
+%% to export the applications described in its internal Python
+%% description language to \boutiques. \todo{Update nipype\_cmd to
+%%   support the latest \boutiques schema and complete this paragraph.}
+
+\subsubsection{Anatomical MRI}
+
+\paragraph{FSL} Several tools from the FMRIB Software
+library (FSL) were ported to CBRAIN using \boutiques: BET, fsl\_anat, FAST,
+FIRST and probtrackx2. Descriptors are available on Neurolinks.
+
+\paragraph{Anima}
+The animaN4BiasCorrection tool from the
+Anima\footnote{\url{https://github.com/Inria-Visages/Anima-Public/wiki}} project was also made available 
+in VIP. The import process was straightforward. \todo{Sorina: reference?}
+
+\subsubsection{Functional MRI}
+
+\paragraph{Niak} The Niak pre-processing pipeline, executed with the Pipeline System
 for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
 in CBRAIN through \boutiques. The integration uses the parallelization
-mechanism described in Section~\ref{sec:parallelization} so that even
+mechanism described earlier so that even
 the invocations processing a single subject can be parallelized. It also
@@
 application. The resulting CBRAIN plugin is available at
-\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It could in
-principle be used in other platforms that support the parallelization
-property, although this has not been tested.
-
-\paragraph{Nipype.} The Nipype workflow engine~\cite{gorgolewski2011nipype} was
-instrumented to export to \boutiques applications described in its
-internal Python description language. \todo{Update nipype\_cmd to
-  support the latest \boutiques schema and complete this paragraph.}
-
-\paragraph{SPM.} The Statistical Parameter Mapping toolbox (SPM)~\cite{penny2011statistical}
-was exported to Boutiques \ldots \todo{Guillaume}
-
-\paragraph{MICCAI challenges.}
-\boutiques was used to integrate 23 pipelines in the VIP platform in
+\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It can
+be used in other platforms that support the parallelization
+property. Descriptors are available on Neurolinks.
+
+\paragraph{GinFizz}
+Ginfizz, a FMRI pipeline allowing to execute multiple 
+preprocessing steps and basic analyses such as: (i) realignment (motion registration and correction),
+(ii) slice timing (temporal interpolation to correct for the delay of acquisition between slices), 
+(iii) coregistration (registering functional and structural data), (iv) normalization (registering 
+of functional and anatomical data to a match a template that has to be specified), and (iv)
+smoothing (spatial filtering to increase Signal to Noise ratio). The use of Docker containers
+(\url{https://hub.docker.com/r/fliiam/ginfizz} based on \url{https://hub.docker.com/r/miykael/nipype_level2}) 
+greatly simplifies its deployment.
+\todo{Sorina: reference?}
+
+\subsubsection{Diffusion imaging}
+
+\paragraph{MRTrix3}
+A few tools from the MRtrix3\footnote{\url{http://www.mrtrix.org/}}
+package for diffusion MRI processing were also made available in VIP,
+as well as a script developed at Creatis, which combines MRtrix3 and
+FSL tools.
+
+\paragraph{ndmg} The NeuroData MRI to Graphs one-click connectome estimation
+pipeline~\cite{kiar2017comprehensive}, developed in Python and leveraging FSL, was
+exported to \boutiques and is available at \url{https://github.com/neurodata/boutiques-tools}.
+The ndmg pipeline is deployed in CBRAIN via its \boutiques descriptor, and is available
+both through Docker and Singularity container environments.
+
+\subsubsection{Image simulation}
+
+\paragraph{CreaPhase}
+A phase contrast simulator called CreaPhase, developed at Creatis, was
+integrated in VIP through \boutiques. The CreaPhase inputs had
+certain particularities (some nedeed to be enclosed in simple quotes,
+other were vectors of variable size enclosed in brackets) that
+required the post-processing of the wrapper script generated by the
+Boutiques importer. \todo{Sorina: reference?}
+
+\paragraph{ODIN} The Odin MRI simulator\footnote{\url{http://od1n.sourceforge.net}} was imported 
+in VIP through \boutiques. Since Odin may require important amounts of
+computing resources (the common use-case consists in iterating on a
+large number of inputs files), Odin is executed by VIP on the EGI grid
+resources, which currently do not support Docker. Consequently, the
+Odin executable was exported from the Docker image and the Odin
+wrapper script was modified accordingly.
+
+\subsubsection{General-purpose toolboxes}
+
+\paragraph{BIDS apps} BIDS apps, a recent effort for the adoption of the Brain Imaging Data Structure (BIDS)
+in common neuroimaging pipelines, require a standardized set of input
+and output parameters~\cite{gorgolewski2017bids}. We developed a tool
+as part of the \boutiques importer to generate a descriptor for any
+such BIDS app. We validated this tool by importing 
+BIDS apps containing the Statistical Parameter Mapping toolbox
+(SPM)~\cite{penny2011statistical} and the ndmg pipeline mentioned
+above. Descriptors are available on Neurolinks.
+
+\paragraph{2016 MICCAI challenges}
+We used \boutiques to integrate 23 pipelines in the VIP platform in
 the context of two challenges organized by the MICCAI conference in
@@
   \texttt{nvidia-docker}\footnote{\url{https://github.com/NVIDIA/nvidia-docker}}
-  tool not supported in Boutiques by default although it could be a possible extension.
+  tool not supported in Boutiques although it could be a possible extension.
 \item A pipeline required more than 10 GB of data dependencies (atlas
@@
 context of MICCAI challenges.
-
-\paragraph{ndmg.} The NeuroData MRI to Graphs one-click connectome estimation
-pipeline~\cite{kiar2017comprehensive}, developed in Python and leveraging FSL, was
-exported to \boutiques and is available at \url{https://github.com/neurodata/boutiques-tools}.
-The ndmg pipeline is deployed in CBRAIN via its \boutiques descriptor, and is available
-both through Docker and Singularity container environments.
-
-\paragraph{BIDS apps.} Brain Imaging Data Structure supported
-applications~\cite{gorgolewski2017bids}, a recent effort for the adoption of BIDS support
-in common neuroimaging pipelines, require a standardized set of input and output parameters
-by tools. A \boutiques tool has been developed which generates a descriptor for any such
-BIDS app, enabling them to be deployed within a broad range of high performance computing
-infrastructures, such as CBRAIN.
-
-\paragraph{Other applications.} Other neuroinformatics applications were ported to CBRAIN using \boutiques:
-FSL tools and ICA AROMA are available at
-\url{https://github.com/aces/cbrain-plugins-neuro} and some pipelines
-of the Human Connectome Project (HCP) are at
-\url{https://github.com/big-data-lab-team/cbrain-plugins-hcp}.
-
-\paragraph{VIP applications.} Several tools were imported in VIP using \boutiques. 
-The first one is a phase contrast simulator called CreaPhase and developed at Creatis. 
-The CreaPhase inputs had certain particularies (some nedeed to be enclosed 
-in simple quotes, other were vectors of variable size enclosed in brackets)
-that required the post-processing of the wrapper script generated by the
-Boutiques importer.
-
-The Odin MRI simulator\footnote{\url{http://od1n.sourceforge.net}} was also imported 
-in VIP using \boutiques. Since Odin  may require important amounts of computing resources 
-(the common use-case consists in iterating on a large number of inputs files), Odin is 
-executed by VIP on the EGI grid resources, which currently do not support Docker. Consequently, 
-the Odin executable was exported from the Docker image which allowed for his CentOS compilation
-and the Odin wrapper script was modified so that it makes no "Docker run" call.
-
-A few MRtrix3\footnote{\url{http://www.mrtrix.org/}} tools were also made available in VIP, as well as 
-a script developed at Creatis, which combines MRtrix3 and FSL tools.
-
-A few other applications from the FLI (France Life Imaging) consortium were also imported in
-VIP using \boutiques. One of them is Ginfizz, a FMRI pipeline allowing to execute multiple 
-preprocessing steps and basic analyses such as: (i) realignment (motion registration and correction),
-(ii) slice timing (temporal interpolation to correct for the delay of acquisition between slices), 
-(iii) coregistration (registering functional and structural data), (iv) normalization (registering 
-of functional and anatomical data to a match a template that has to be specified), and (iv)
-smoothing (spatial filtering to increase Signal to Noise ratio). The use of Docker containers
-(\url{https://hub.docker.com/r/fliiam/ginfizz} based on \url{https://hub.docker.com/r/miykael/nipype_level2}) 
-greatly simplifies its deployment. The animaN4BiasCorrection tool from the
-Anima\footnote{\url{https://github.com/Inria-Visages/Anima-Public/wiki}} project was also made available 
-in VIP. The import process was straightforward.
-
-\note{sorina}{@tristan: please check if you agree with the descriptions above}
-
-%%\begin{table}
-%%  \begin{tabular}{l|ccccccc}
-%%    \rowcolor[gray]{0.9}
-%%    Application                      & \multicolumn{7}{c}{Input/Output properties} \\
-%%    \rowcolor[gray]{0.9}
-%%                                     & Dependencies      &Groups             & Lists             & Optional          &Default            & Type restriction             & Min/max\\
-%%    
-%%    \hline
-%%    Niak pre-processing              &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\\
-%%    ICA AROMA                        &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
-%%    FSL tools &&&&&&\\
-%%    \multicolumn{1}{r|}{anat}        &\cellcolor{gray!75}&\cellcolor{gray!75}&                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
-%%    \multicolumn{1}{r|}{bet}         &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}\\      
-%%    \multicolumn{1}{r|}{fast}        &\cellcolor{gray!75}&                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}\\
-%%    \multicolumn{1}{r|}{first}       &                   &                   &\cellcolor{gray!75}&\cellcolor{gray!75}&\cellcolor{gray!75}&                   &                   \\
-%%    FSL Nipype interfaces     \\
-%%    HCP PreFreesurfer                &                   &                   &                   &                   &                   &                   &                   \\
-%%    ndmg                &                   &       \cellcolor{gray!75}            &                   &          \cellcolor{gray!75}         &      \cellcolor{gray!75}             & \cellcolor{gray!75} &                   \\
-%%    MICCAI challenge (23 pipelines)\\&                   &                   &                   &                   &                   &                   & \\
-%%    \end{tabular}
-%%  \vspace*{0.5cm}
-%%  \begin{tabular}{l|ccc|ccc}
-%%    \hline
-%%    \rowcolor[gray]{0.9}
-%%    Application                     &\multicolumn{3}{c|}{Containers}            &  \multicolumn{3}{c}{Other properties}  \\
-%%    \rowcolor[gray]{0.9}
-%%                                    &Docker              & Singularity& Rootfs  & Config. file & Env vars  & Parallelization   \\
-%%    \hline
-%%    Niak pre-processing             &\cellcolor{gray!75} &\cellcolor{gray!75}&         &              &\cellcolor{gray!75}&\cellcolor{gray!75}\\
-%%    ICA AROMA                       &\cellcolor{gray!75} &            &         &              &&\\
-%%    FSL tools &&&&&\\
-%%    \multicolumn{1}{r|}{anat}       &                    &            &         &              &&\\
-%%    \multicolumn{1}{r|}{bet}        &                    &            &         &              &&\\
-%%    \multicolumn{1}{r|}{fast}       &                    &            &         &              &&\\
-%%    \multicolumn{1}{r|}{first}      &                    &            &         &              &&\\
-%%    FSL Nipype interfaces           &                    &            &         &              &&\\
-%%    HCP PreFreesurfer               &\cellcolor{gray!75} &            &         &              &&\\
-%%    ndmg                            &\cellcolor{gray!75} &\cellcolor{gray!75}&         &              &&\\
-%%    MICCAI challenge (23 pipelines) &\cellcolor{gray!75} &            &         &              &&\\
-%%  \end{tabular} 
-%%  \caption{Applications available in Boutiques with descriptor features used.}
-%%  \label{table:applications}
-%%\end{table}
 
@@
 
+\todo{CBRAIN's custom property might be standardized.}
+
 \subsection{Reproducibility}
@@
 
+\section{Acknowledgments}
+
+\todo{FLI}
+
 \bibliography{biblio}

### 2017-11-02 boutiquespaper.tex
added a few TODOs
Integrated Charles' input

@@
   Guillaume Flandin, John Flavin, Pascal Girard, Krzysztof
-  J. Gorgolewski, Charles G. Guttmann, Val\'erie Hayot-Sasson,
+  J. Gorgolewski, Charles R.G. Guttmann, Val\'erie Hayot-Sasson,
   Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux,
@@
 \subsubsection{SPINE}
-%% gk: shall we check with them, and if nothing, remove?
-\todo{Charles}
+SPINE (\url{http://spinevirtuallab.org}), which stands for Structured
+Planning and Implementation of New Explorations, is a web-based,
+collaborative platform (virtual laboratory) designed to support the
+design and execution of experiments centered on specific scientific
+questions.  SPINE enables distributed data collection and management,
+as well as experiment design, execution, and review. Boutiques serves
+as SPINE's algorithm and workflow repository, and enables unequivocal
+referencing of specific workflows applied to specified datasets within
+an experiment, thereby describing the provenance and facilitating the
+reproducibility of image-derived measurements. Workflows in SPINE may
+combine human image annotation with automated image processing
+algorithms. Future development will focus on extending Boutiques
+workflow descriptors to include the identification and
+characterization of human operators and their specific historic
+performance on the required tasks. SPINE is currently hosted at
+Brigham \& Women's Hospital in Boston, and supports several
+international projects.
 
@@
 library (FSL) were ported to CBRAIN using \boutiques: BET, fsl\_anat, FAST,
-FIRST and probtrackx2. Descriptors are available on Neurolinks.
+FIRST and probtrackx2. Descriptors are available on Neurolinks. \todo{add reference.}
 
@@
 as well as a script developed at Creatis, which combines MRtrix3 and
-FSL tools.
+FSL tools. \todo{Sorina: replace footnote by citation, if available.}
 
@@
 Odin executable was exported from the Docker image and the Odin
-wrapper script was modified accordingly.
+wrapper script was modified accordingly. \todo{Sorina: replace footnote by citation, if available.}
 
@@
 context of MICCAI challenges.
+\todo{Cite Mathieu's paper.}
 

### 2017-11-02 boutiquespaper.tex
A paragraph on Nipype

@@
 
-CBRAIN~\cite{SHER-14} is a web platform that can process data
-distributed into multiple storage locations on computing clusters and
-clouds. CBRAIN offers transparent access to remote data sources,
-distributed computing sites, and an array of processing and
-visualization tools within a controlled, secure environment.  The
-CBRAIN service deployed at the Montreal Neurological Institute relies
-on the infrastructure provided by Compute Canada~\cite{das2016mni}. It
-currently provides 500+ collaborators in 22 countries with web access
-to several systems, including 6 clusters of the Compute Canada
-high-performance computing infrastructure (totaling more than 100,000
-computing cores and 40PB of disk storage) and Amazon EC2. CBRAIN
-transiently stores about 10 million files representing over 50TB
-distributed in 42 servers. 51 public data processing applications are
-integrated and over 340,000 processing batches have been submitted since 2010.
+CBRAIN (\url{http://cbrain.mcgill.ca}) ~\cite{SHER-14} is a web
+platform that can process data distributed into multiple storage
+locations on computing clusters and clouds. CBRAIN offers transparent
+access to remote data sources, distributed computing sites, and an
+array of processing and visualization tools within a controlled,
+secure environment.  The CBRAIN service deployed at the Montreal
+Neurological Institute relies on the infrastructure provided by
+Compute Canada~\cite{das2016mni}. It currently provides 500+
+collaborators in 22 countries with web access to several systems,
+including 6 clusters of the Compute Canada high-performance computing
+infrastructure (totaling more than 100,000 computing cores and 40PB of
+disk storage) and Amazon EC2. CBRAIN transiently stores about 10
+million files representing over 50TB distributed in 42 servers. 51
+public data processing applications are integrated and over 340,000
+processing batches have been submitted since 2010.
 
@@
 \todo{Gunnar, Nathanael, John}
+
+\subsubsection{Nipype}
+Nipype (\url{http://nipype.readthedocs.io/en/latest/})~\cite{gorgolewski2011nipype} is a workflow engine widely used in
+neuroinformatics. Nipype workflows can be composed from \boutiques
+tools using the Python API. As an example, we implemented NipBIDS, a
+Nipype workflow to process BIDS datasets using BIDS apps. NipBIDS
+iterates participant analyses on all the subjects found in a BIDS
+dataset and runs a group analysis if requested. It is available at
+\url{https://github.com/big-data-lab-team/sim/tree/master/sim/other_wf_examples/nipype}.
 

### 2017-11-02 boutiquespaper.tex
Fixed layout

@@
                linenos=false,
-               xleftmargin=21pt,
+               xleftmargin=0pt,
                tabsize=4]{js}
@@
 {
-    "id" : "logfile",
-    "name" : "Log file",
-    "description" : "The output log file from the example application",
-    "path-template" : "log-[STRING_INPUT]",
-    "value-key" : "[LOG]",
-    "path-template-stripped-extensions" : [".txt", ".csv"],
-    "optional" : false
+  "id": "logfile",
+  "name": "Log file",
+  "description": "The output log file",
+  "path-template": "log-[STRING_INPUT]",
+  "value-key": "[LOG]",
+  "path-template-stripped-extensions": [".txt", ".csv"],
+  "optional": false
 }
@@
 \subsubsection{Nipype}
-Nipype (\url{http://nipype.readthedocs.io/en/latest/})~\cite{gorgolewski2011nipype} is a workflow engine widely used in
-neuroinformatics. Nipype workflows can be composed from \boutiques
-tools using the Python API. As an example, we implemented NipBIDS, a
-Nipype workflow to process BIDS datasets using BIDS apps. NipBIDS
-iterates participant analyses on all the subjects found in a BIDS
-dataset and runs a group analysis if requested. It is available at
+Nipype
+(\url{http://nipype.readthedocs.io/en/latest/})~\cite{gorgolewski2011nipype}
+is a workflow engine widely used in neuroinformatics. Nipype workflows
+can be composed from \boutiques tools using the Python API. As an
+example, we implemented NipBIDS, a Nipype workflow to process BIDS
+datasets using BIDS apps imported as \boutiques tools. NipBIDS iterates
+participant analyses on all the subjects found in a BIDS dataset and
+runs a group analysis if requested. It is available at
 \url{https://github.com/big-data-lab-team/sim/tree/master/sim/other_wf_examples/nipype}.

### 2017-11-02 boutiquespaper.tex
A few fixes, added paragraph on BIDS apps, added affiliations, removed Flywheel

@@
 
-\author{Tristan Glatard, Gregory Kiar, Tristan Aumentado-Armstrong,
-  Natacha Beck, Pierre Bellec, Remi Bernard, Sorina Camarasu-Pop,
-  Fr\'ed\'eric Cervenansky, Samir Das,  Rafael Ferreira da Silva,
-  Guillaume Flandin, John Flavin, Pascal Girard, Krzysztof
-  J. Gorgolewski, Charles R.G. Guttmann, Val\'erie Hayot-Sasson,
-  Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux,
-  Marc-\'Etienne Rousseau, Gunnar Schaeffer, Alan C. Evans}
+\author[1]{Tristan Glatard}
+\author[2,3]{Gregory Kiar}
+\author[2,3]{Tristan Aumentado-Armstrong}
+\author[2,3]{Natacha Beck}
+\author[4]{Pierre Bellec}
+\author[2,3]{R\'emi Bernard}
+\author[5]{Sorina Camarasu-Pop}
+\author[5]{Fr\'ed\'eric Cervenansky}
+\author[2,3]{Samir Das}
+\author[6]{Rafael Ferreira da Silva}
+\author[7]{Guillaume Flandin}
+\author[5]{Pascal Girard}
+\author[8]{Krzysztof J. Gorgolewski}
+\author[9]{Charles R.G. Guttmann}
+\author[1]{Val\'erie Hayot-Sasson}
+\author[4]{Pierre-Olivier Quirion}
+\author[2,3]{Pierre Rioux}
+\author[10]{Marc-\'Etienne Rousseau}
+\author[2]{Alan C. Evans}
+
+\affil[1]{Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada}
+\affil[2]{McGill University, Montreal, Canada}
+\affil[3]{Montreal Neurological Institute, Montreal, Canada}
+\affil[4]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montréal, QC, Canada}
+\affil[5]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France}
+\affil[6]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA}
+\affil[7]{\todo{to be added}}
+\affil[8]{Department of Psychology, Stanford University, Stanford, California, USA}
+\affil[9]{Partners MS Center, Brigham and Women's Hospital, Brookline, Massachusetts, USA}
+\affil[10]{Compute Canada}
 
@@
 workflows expressed in a variety of languages, as detailed in
-~\cite{glatard2017software}.
+~\cite{GLATARD2017239}.
 
@@
 different purposes. For instance, Docker is well suited for developers
-and users who manipulate applications on their local workstations or the
-cloud. It is well documented, maintained and it has a rich ecosystem of
-tools to help build and run containers on most operating systems. Singularity
-is more suited for users and platforms that need to run applications
-on shared computing clusters. Bridges exist among these containers
-formats to convert container images across frameworks. For instance, a
-platform dedicated to high-performance computing may accept descriptors
-referring to Docker containers to facilitate application integration
-by developers, and internally convert container images to Singularity
-to run applications efficiently on clusters.
+and users who manipulate applications on their local workstations or
+the cloud. It is well documented, maintained and it has a rich
+ecosystem of tools to help build and run containers on most operating
+systems. Singularity is more suited for users and platforms that need
+to run applications on shared computing clusters. Bridges exist among
+these containers formats to convert container images across
+frameworks. For instance, a platform dedicated to high-performance computing may accept
+descriptors referring to Docker containers to facilitate application
+integration by developers, and run these images on clusters using
+Singularity.
 
@@
 
-\subsubsection{Flywheel}
-%% gk: shall we check with them, and if nothing, remove?
-\todo{Gunnar, Nathanael, John}
-
 \subsubsection{Nipype}
@@
 as part of the \boutiques importer to generate a descriptor for any
-such BIDS app. We validated this tool by importing 
-BIDS apps containing the Statistical Parameter Mapping toolbox
+such BIDS app. We validated this tool by importing BIDS apps
+containing the Statistical Parameter Mapping toolbox
 (SPM)~\cite{penny2011statistical} and the ndmg pipeline mentioned
@@
 inputs. Extending the \boutiques descriptor has two main goals. First,
-accurately describing applications improves validation: incorrect input
-values and execution results are more precisely detected when the
-application descriptor is comprehensive, which is essential to improve
-the reliability and user-friendliness of execution
+accurately describing applications improves validation: incorrect
+input values and execution results are more precisely detected when
+the application descriptor is comprehensive, which is essential to
+improve the reliability and user-friendliness of execution
 platforms. Secondly, a rich descriptor schema reduces the need for
-custom wrappers to integrate applications as even convoluted command
-lines can be described: this is particularly interesting for
-applications installed in containers since adding a wrapper to the
-execution stack requires an update in the container image.
+custom application wrappers, for instance to create configuration
+files from command-line options, as even convoluted command lines can
+be described. This is particularly useful for containerized
+applications since adding a wrapper requires an update of the
+container image.
 
@@
 only (see Listing~\ref{listing:minimal}). For platforms, we aim at
-supporting as many features as possible in the executor so that
-only the following steps need to be implemented to support \boutiques
-in a platform, regardless of the complexity of the descriptor:
+supporting as many features as possible in the \texttt{bosh} executor
+so that only the following steps need to be implemented to support
+\boutiques in a platform, regardless of the complexity of the
+descriptor:
 \begin{enumerate}
-  \item Input entry: generate the interface to allow users to enter
-    input values.
-  \item Input validation: create a JSON object from the interface,
+  \item Input entry: generate the interface to enter inputs.
+  \item Input validation: create a JSON invocation from the interface,
     validate it against the invocation schema.
-  \item Execution: generate and run the command line from the local
-    executor and input JSON object.
+  \item Input delivery: transfer the input files to the application
+    execution location.
+  \item Execution: pass the invocation to the \texttt{bosh} executor to run the application.
   \item Output delivery: from the descriptor, identify the output files
@@
   "command-line": "echo [PARAM] > output.txt",
-  "schema-version": "0.4",
+  "schema-version": "0.5",
   "inputs": [{
@@
       "name": "Parameter",
-      "value-key": "[INPUT_FILE]",
-      "type": "File"
+      "value-key": "[PARAM]",
+      "type": "Number"
   }],
@@
   }]
-
 }
@@
 
-The parallelization framework described in~\ref{sec:parallelization}
-supports workflows by considering them as both applications and
-platforms, which we think is a powerful but underexploited
-model. Workflows are applications because they are invoked like any
-other script or executable, possibly through a command line. Workflows
-are also platforms because they may invoke other applications that
-might themselves be described in \boutiques. This model is powerful
-because it shields the \boutiques specification from specific workflow
-constructs and allows a wide range of workflow engines to be described
-uniformly.
-
-\todo{CBRAIN's custom property might be standardized.}
+\boutiques intentionally does not provide a workflow language to
+compose tools, as this is already offered by several workflow
+engines. Workflows can be either composed from \boutiques tools, as we
+illustrated with the Nipype and MOTEUR engines, or described as
+\boutiques applications, as we demonstrated with the Niak fMRI
+pre-processing pipeline. Furthermore, the CBRAIN platform has a
+sub-tasking mechanism that allows \boutiques tools to submit tasks,
+which is used to parallelize workflows wrapped as \boutiques
+tools. Based on the CBRAIN experience, we may specify the sub-tasking
+mechanism in \boutiques so that other platforms can benefit from it.
+This model is powerful because it shields the \boutiques specification
+from specific workflow constructs and allows a wide range of workflow
+engines to be described and used uniformly.
 
@@
 
-\boutiques helps reproduce\footnote{in Peng's
-  sense~\cite{peng2011reproducible}} executions through containers and
-formal command-line descriptions. With \boutiques, complete sets of
+\boutiques helps computational reproducibility, in Peng's
+sense~\cite{peng2011reproducible}, through containers and formal
+command-line descriptions. With \boutiques, complete sets of
 applications could be easily migrated across execution platforms,
@@
 complemented by a provenance framework that captures a detailed trace
-of the execution.
-
-At a higher level, reproducibility is impacted by factors
-such as data sharing~\cite{poldrack2014making} which are out of the
-scope of \boutiques.
+of the execution. We plan to leverage the provenance format being
+defined by the Neuro-Imaging Data Model
+Workflow~\cite{ghosh2017capturing} initiative for this purpose.
 
@@
 command-line applications. While such applications cover a large
-fractions of the tools involved in scientific data processing, other
+fraction of the tools involved in scientific data processing, other
 types of programs exist such as web services, interactive tools, and
@@
 creating hundreds of richly described applications potentially usable
-by end-users. However, it remains unclear how to validate that these
-applications actually execute as expected beyond executing packaged tests
-distributed alongside a given tool. A \boutiques-specific testing
-framework could be designed and potentially fed by existing frameworks,
-to address this issue.
+by end-users. However, the automated validation of such applications
+remains challenging. A \boutiques-specific testing framework could be
+designed and potentially fed by existing frameworks, to address this
+issue.
 
@@
 possibly after a platform administrator inspected the
-container. Containers are currently bulky file archives that are
+container. Containers are bulky file archives that are
 cumbersome to inspect. Tools need to be developed to allow for an
@@
 
-\todo{BIDS apps}
-
-The Common Workflow Language\footnote{\url{http://www.commonwl.org}}
-is the piece of related work most closely related to
-\boutiques. Several other application description frameworks have been
-developed but they do not support containers.
+Several frameworks have been developed to describe and integrate
+applications in various types of platforms. \boutiques focuses on (1)
+fully-automatic integration of applications, including deployment on
+heterogeneous computing resources through containers, (2)
+comprehensive input validation through a strict JSON schema, (3)
+flexible application description through a rich JSON
+schema.
 
@@
 
-The Command Line Tool Description of the Common Workflow Language
-(CWL) overlaps with the \boutiques descriptor. According to the history
-of the CWL and \boutiques Github repositories, CWL started 6 months
-before \boutiques (September 2014 vs. May 2015). The differences
-highlighted below were identified from version 1.0 of the CWL Command
-Line Tool
-Description\footnote{\url{http://www.commonwl.org/v1.0/CommandLineTool.html}}.
+The Common Workflow Language
+(CWL\footnote{\url{http://www.commonwl.org}}) is the piece of related
+work most closely related to \boutiques as it provides a formal way to
+describe containerized applications. In particular, CWL's Command Line
+Tool Description overlaps with the \boutiques descriptor. This Section
+highlights the main differences between CWL and \boutiques, based on
+version 1.0 of the CWL Command Line Tool
+Description\footnote{\url{http://www.commonwl.org/v1.0/CommandLineTool.html}}. According
+to GitHub, CWL started 6 months before \boutiques (September 2014
+vs. May 2015).
 
@@
 
-CWL has a formal command-line definition while \boutiques is more
+CWL imposes a strict command-line format while \boutiques is more
 flexible. CWL specifies command lines using an array containing an
@@
 instance, a \boutiques command line may easily include input
-decompression using, e.g., the tar command in addition to the main
-tool command. On the contrary, CWL applications may only have a single
-executable. Importantly, \boutiques' template system allows supporting
-configuration files (see Section~\ref{sec:configuration-files}).
+decompression using the \texttt{tar} command in addition to the main
+tool command. Importantly, \boutiques' template system allows
+supporting configuration files.
 
@@
 \item  CWL has a rich set of types whereas \boutiques only has simple
-types. Again, this may be seen as a feature or as an overhead
+types. This may again be seen as a feature or as an overhead
 depending on the context. \boutiques tries to limit the complexity of
@@
 properties for basename, dirname, location, path, checksum, etc.
-\item \boutiques supports various types of containers (Docker, Singularity,
-rootfs) while CWL supports only Docker. Both tools have rich requirements:
-for instance, they may include RAM, disk usage, and walltime estimate; 
-whereas \boutiques also supports number of requested CPU cores. CWL has hints, i.e.,
-recommendations that only lead to warnings when not respected.
+\item \boutiques supports various types of containers (Docker,
+  Singularity, rootfs) while CWL supports only Docker. Both tools have
+  rich requirements: for instance, they may include RAM, disk usage,
+  and walltime estimate. CWL has hints, i.e., recommendations that
+  only lead to warnings when not respected.
 \item In \boutiques, dependencies can be defined among inputs, e.g., to
@@
 
-\subsubsection{Other minor differences}
-
-The \boutiques and CWL schemas have other minor differences that we
-summarized in a public
-spreadsheet\footnote{\url{https://docs.google.com/spreadsheets/d/1eLhpxzaaMPCIUFmV5Trs-WtaMpBO-c6QDQsmhwP7QIE/edit\#gid=0}}.
+\subsection{BIDS apps}
+
+BIDS apps~\cite{gorgolewski2017bids} specify a framework for
+neuroimaging applications to process datasets complying to the Brain
+Imaging Data Structure (BIDS). They share common goals with
+\boutiques, in particular reusability across platforms through
+containerization. Conceptually, however, BIDS apps and \boutiques are
+different since BIDS apps intend to standardize application interfaces
+while \boutiques intends to describe them as flexibly as
+possible. BIDS apps have a specific set of inputs and outputs, for
+instance the input dataset, that have to be present in a specific
+order on the command line for the application to be valid. The
+specification adopted by BIDS apps simplifies the integration of
+applications in platforms as they all comply to the same
+interface. However, it is also limited to the subset of neuroimaging
+applications that process BIDS datasets and it does not formally
+describe application-specific inputs. All in all, BIDS apps and
+\boutiques complement each other: BIDS apps provide a practical way to
+integrate neuroimaging applications, while \boutiques offers a formal
+description of their specific parameters.  \boutiques descriptors can
+be generated from BIDS apps using the \texttt{bosh} importer.
 
@@
 
-Several frameworks have been created to facilitate the
+Several other frameworks have been created to facilitate the
 integration of command-line applications in platforms. In
@@
 remains tightly bound to the Common Toolkit's C++ implementation which
-limits its adoption, e.g., in web platforms \todo{Fr\'ed\'eric, maybe
-  you have more to say on this}.
+limits its adoption, e.g., in web platforms.
 
@@
 platform~\cite{terstyanszky2014enabling}, in a similar but different
-way than proposed by \boutiques (see Section~\ref{sec:parallelization}).
+way than proposed by \boutiques.
 
@@
 
-\todo{FLI}
+Pipeline integration for the MICCAI 2016 challenge was funded by the
+French National Agency for Research (ANR) through ``France
+Life-Imaging''. We also thank Compute Canada and Calcul Québec for
+providing a computing infrastructure supporting Docker and Singularity
+containers. This research was undertaken thanks in part to funding
+from the Canada First Research Excellence Fund, awarded to McGill
+University for the Healthy Brains for Healthy Lives initiative.
 

### 2017-11-03 boutiquespaper.tex
Fixed a few TODOs, added Fig 1

@@
 \affil[6]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA}
-\affil[7]{\todo{to be added}}
+\affil[7]{Wellcome Trust Centre for Neuroimaging}
 \affil[8]{Department of Psychology, Stanford University, Stanford, California, USA}
@@
   interoperability would save cost and improve the quality of the ported
-  applications. We describe \boutiques, a descriptive framework and
+  applications. We present \boutiques, a descriptive framework and
   command line utility that allows for automatic import and exchange of
   applications across platforms. \boutiques adopts a rich, flexible
-  application description format which relies on Linux container
+  application description format that relies on Linux container
   environments Docker and Singularity to install applications
   across platforms in a lightweight manner. \boutiques is currently used
-  by three distinct virtual research platforms, and has been used to
+  by several distinct virtual research platforms, and has been used to
   describe dozens of neuroinformatics tools.
@@
 \begin{keywords}
-Keyword1; keyword 2; keyword 3 (Three to ten keywords representing the main content of the article)
+Application integration; Containers; Neuroinformatics.
 \end{keywords}
@@
 
-\todo{Intro on the relevance of pipeline sharing for reproducible and open science.}
-
 \todo{Check passive voice everywhere.}
@@
 engines and virtual research environments commonly integrate
-third-party applications to facilitate the processing of large
-datasets on high-performance computing resources. Such applications,
-however, are often repeatedly and manually ported to various platforms
-whereas this porting effort would greatly benefit from being
-mutualized. Meanwhile, virtual container systems such as Docker\todo{cite} and
-Singularity\todo{cite} have emerged to facilitate the sharing of software and
-improve the reproducibility of analyses by defining immutable,
-reusable execution environments. Nevertheless, the automatic
-integration of containerized applications in platforms require formal
-annotations of container images, to describe the available command lines.
-
-\boutiques is a system to share command-line applications across
-platforms. In \boutiques, a command line is described using a flexible
-template comprising the inputs it requires and the outputs it
-produces. Inputs may be inter-dependent and passed in configuration
-files or directly on the command line. Such formal descriptions,
-simply referred to as \emph{descriptors}, link to a container image
-where the given application and all its dependencies are
-installed. \boutiques descriptors allow for automatic application
+third-party applications to enable data processing on various types of
+computing resources. Such applications, however, are often repeatedly
+and manually ported to different platforms whereas sharing this
+porting effort would contribute to Open Science and improve
+computational reproducibility. Meanwhile, software container systems
+such as Docker\footnote{\url{https://www.docker.com}} and
+Singularity~\cite{kurtzer2017singularity} have emerged to facilitate
+the sharing and migration of software by defining immutable, reusable
+execution environments.
+
+We present \boutiques, a system to publish, integrate and execute
+containerized command-line applications across platforms (see
+Figure~\ref{fig:diagram}).  In \boutiques, a command line is described
+using a flexible template comprising the inputs it requires and the
+outputs it produces. Inputs may be inter-dependent and passed in
+configuration files or directly on the command line. Such formal
+descriptions, simply referred to as \emph{descriptors}, link to a
+container image where the given application and all its dependencies
+are installed. \boutiques descriptors allow for automatic application
 integration in platforms, and advanced validation of input values to
@@
 produced by scientific application developers, stored alongside their
-application, indexed by common repositories, and consumed by
-execution platforms.  A set of core tools facilitate the construction,
+application, indexed by common repositories, and consumed by execution
+platforms.  A set of core tools facilitate the construction,
 validation, import, execution, and publishing of \boutiques
 descriptors.
+\begin{figure}
+  \includegraphics[width=\columnwidth]{boutiquesfig1.pdf}
+  \caption{Publication, integration and execution of applications with \boutiques.}
+  \label{fig:diagram}
+\end{figure}
 
@@
 \boutiques. For this purpose, \boutiques relies on an
-application-specific JSON schema\todo{cite}, called \emph{invocation schema}, to
+application-specific JSON schema, called \emph{invocation schema}, to
 specify the input values accepted by an application. Platforms can
@@
 
-\subsubsection{Pegasus}
-%% gk: to what extent does this exist?
-\todo{Rafael}
-
 \subsubsection{SPINE}
@@
 
-\subsubsection{Anatomical MRI}
-
-\paragraph{FSL} Several tools from the FMRIB Software
-library (FSL) were ported to CBRAIN using \boutiques: BET, fsl\_anat, FAST,
-FIRST and probtrackx2. Descriptors are available on Neurolinks. \todo{add reference.}
+\subsubsection{Anatomical imaging}
+
+\paragraph{FSL} Several MRI analysis tools from the FMRIB Software
+library (FSL~\cite{jenkinson2012fsl}) were ported to CBRAIN using
+\boutiques: BET, fsl\_anat, FAST and FIRST. Descriptors are available
+on Neurolinks.
 
 \paragraph{Anima}
-The animaN4BiasCorrection tool from the
-Anima\footnote{\url{https://github.com/Inria-Visages/Anima-Public/wiki}} project was also made available 
-in VIP. The import process was straightforward. \todo{Sorina: reference?}
-
-\subsubsection{Functional MRI}
-
-\paragraph{Niak} The Niak pre-processing pipeline, executed with the Pipeline System
+\texttt{animaN4BiasCorrection}, an ITK-based bias field correction tool from the
+Anima\footnote{\url{https://github.com/Inria-Visages/Anima-Public/wiki}}
+project was also made available in VIP. The import process was
+straightforward.
+
+\subsubsection{Functional MRI (fMRI)}
+
+\paragraph{Niak} The Niak fMRI pre-processing pipeline, executed with the Pipeline System
 for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
 in CBRAIN through \boutiques. The integration uses the parallelization
-mechanism described earlier so that even
-the invocations processing a single subject can be parallelized. It also
-allows CBRAIN to leverage the efficient agent model used in PSOM, as
-described in~\cite{GLAT-16}. The integration required some work in
-PSOM to facilitate its invocation as a non-interactive command-line
+mechanism described earlier so that even the invocations processing a
+single subject can be parallelized. It also allows CBRAIN to leverage
+the efficient agent model used in PSOM, as described
+in~\cite{GLAT-16}. The integration required some work in PSOM to
+facilitate its invocation as a non-interactive command-line
 application. The resulting CBRAIN plugin is available at
-\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It can
-be used in other platforms that support the parallelization
-property. Descriptors are available on Neurolinks.
+\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It can be used in
+other platforms that support the parallelization property. Descriptors
+are available on Neurolinks.
 
 \paragraph{GinFizz}
-Ginfizz, a FMRI pipeline allowing to execute multiple 
-preprocessing steps and basic analyses such as: (i) realignment (motion registration and correction),
-(ii) slice timing (temporal interpolation to correct for the delay of acquisition between slices), 
-(iii) coregistration (registering functional and structural data), (iv) normalization (registering 
-of functional and anatomical data to a match a template that has to be specified), and (iv)
-smoothing (spatial filtering to increase Signal to Noise ratio). The use of Docker containers
-(\url{https://hub.docker.com/r/fliiam/ginfizz} based on \url{https://hub.docker.com/r/miykael/nipype_level2}) 
-greatly simplifies its deployment.
-\todo{Sorina: reference?}
+We integrated the
+GinFizz\footnote{\url{https://github.com/thomashirsch/ginfizz}}
+Nipype-based fMRI pre-processing pipeline in VIP with \boutiques. \todo{Sorina, Pascal, a few lines on the technical deployment?}
+%% Ginfizz performs multiple
+%% preprocessing steps and basic analyses such as: (i) realignment
+%% (motion registration and correction), (ii) slice timing (temporal
+%% interpolation to correct for the delay of acquisition between slices),
+%% (iii) coregistration (registering functional and structural data),
+%% (iv) normalization (registering of functional and anatomical data to a
+%% match a template that has to be specified), and (iv) smoothing
+%% (spatial filtering to increase Signal to Noise ratio).
 
@@
 \paragraph{MRTrix3}
-A few tools from the MRtrix3\footnote{\url{http://www.mrtrix.org/}}
-package for diffusion MRI processing were also made available in VIP,
+A few tools from the MRtrix3
+package~\cite{tournier2012mrtrix}\footnote{\url{http://www.mrtrix.org/}} for diffusion MRI processing were also made available in VIP,
 as well as a script developed at Creatis, which combines MRtrix3 and
-FSL tools. \todo{Sorina: replace footnote by citation, if available.}
+FSL tools.
 
@@
 
-\paragraph{ODIN} The Odin MRI simulator\footnote{\url{http://od1n.sourceforge.net}} was imported 
+\paragraph{ODIN} The Odin MRI simulator~\cite{jochimsen2004odin}\footnote{\url{http://od1n.sourceforge.net}} was imported 
 in VIP through \boutiques. Since Odin may require important amounts of
@@
 Odin executable was exported from the Docker image and the Odin
-wrapper script was modified accordingly. \todo{Sorina: replace footnote by citation, if available.}
-
-\subsubsection{General-purpose toolboxes}
-
-\paragraph{BIDS apps} BIDS apps, a recent effort for the adoption of the Brain Imaging Data Structure (BIDS)
+wrapper script was modified accordingly.
+
+\subsubsection{BIDS apps}
+
+BIDS apps, an effort for the adoption of the Brain Imaging Data Structure (BIDS)
 in common neuroimaging pipelines, require a standardized set of input
@@
 
-\paragraph{2016 MICCAI challenges}
+\subsubsection{2016 MICCAI challenges}
+
 We used \boutiques to integrate 23 pipelines in the VIP platform in
@@
 and of tumor volumes in PET images (PETSEG
-challenge\footnote{\url{https://portal.fli-iam.irisa.fr/petseg-challenge/overview}}). The
+challenge~\cite{hatt2018}). The
 pipelines were ported to VIP and executed on 205 subjects in a few
@@
 context of MICCAI challenges.
-\todo{Cite Mathieu's paper.}
 
@@
 
-The recent advent of virtual containers requires a new generation of
+The recent advent of software containers requires a new generation of
 application description frameworks that are independent from any

### 2017-11-03 boutiquespaper.tex
formatting
Various fixes, including issue #2

@@
 descriptors.
+
+The remainder of this paper describes the \boutiques system
+(Section~\ref{sec:system}) and reports on its adoption by platforms
+and applications in the neuroinformatics domain, our primary field of
+interest (Section~\ref{sec:results}). It closes on a discussion and
+comparison with related systems.
+
 \begin{figure}
@@
 \end{figure}
-
-The remainder of this paper describes the \boutiques system
-(Section~\ref{sec:system}) and reports on its adoption by platforms
-and applications in the neuroinformatics domain, our primary field of
-interest (Section~\ref{sec:results}). It closes on a discussion and
-comparison with related systems.
-
-%\note{Tristan}{Vocabulary (check consistency across the paper): tool vs. application}
-%\note{Tristan}{Targeted journals: Scientific Data, Giga Science, Frontiers in Neuroinformatics, Future Generation Computer Systems}
 
@@
 
-CBRAIN (\url{http://cbrain.mcgill.ca}) ~\cite{SHER-14} is a web
+CBRAIN (\url{http://github.com/aces/cbrain}) ~\cite{SHER-14} is a web
 platform that can process data distributed into multiple storage
@@
 through the file system, and it is also powerful as it enables the
-parallelization of complex workflows such as the FSL and Niak ones
+parallelization of complex workflows such as the Niak ones
 described later.

### 2017-11-06 boutiquespaper.tex

@@
 GinFizz\footnote{\url{https://github.com/thomashirsch/ginfizz}}
-Nipype-based fMRI pre-processing pipeline in VIP with \boutiques. \todo{Sorina, Pascal, a few lines on the technical deployment?}
+Nipype-based fMRI pre-processing pipeline in VIP with \boutiques. 
+A few technical issues were encountered with the Docker container. 
+They were mainly due to the re-use of multiple containers build 
+with specific users, leading to file access issues. We therefore 
+had to override the permissions on the files and folders necessary
+to the correct execution of the GinFizz application.
+
 %% Ginfizz performs multiple
@@
 \paragraph{CreaPhase}
-A phase contrast simulator called CreaPhase, developed at Creatis, was
+A phase contrast simulator called CreaPhase\cite{WEBE-17}, developed at Creatis, was
 integrated in VIP through \boutiques. The CreaPhase inputs had
@@
 required the post-processing of the wrapper script generated by the
-Boutiques importer. \todo{Sorina: reference?}
+Boutiques importer.
 

### 2017-11-06 boutiquespaper.tex
Several fixes following a complete proof-read

--- 2017-11-03 boutiquespaper.tex
@@
 \author[2,3]{R\'emi Bernard}
+\author[5]{Axel Bonnet}
 \author[5]{Sorina Camarasu-Pop}
@@
 \author[10]{Marc-\'Etienne Rousseau}
-\author[2]{Alan C. Evans}
+\author[2,3]{Alan C. Evans}
 
@@
 
-\begin{abstract}Porting applications to execution platforms such as web
+\begin{abstract}Integrating applications in execution platforms such as web
   portals and workflow engines is critical to enable their sharing
   among scientific communities and their efficient execution on
-  computing infrastructures. However, application porting remains a
-  costly human effort that consists of 1)~installing the application
+  computing infrastructures. However, application integration remains
+  a costly human effort that consists of 1)~installing the application
   on the target infrastructure, 2)~describing the application in a
   format compatible with the execution platform, and 3)~generating
-  proper user interfaces. Due to the variety of computing platforms,
-  application porting efforts are often replicated several times while
-  interoperability would save cost and improve the quality of the ported
-  applications. We present \boutiques, a descriptive framework and
-  command line utility that allows for automatic import and exchange of
-  applications across platforms. \boutiques adopts a rich, flexible
-  application description format that relies on Linux container
-  environments Docker and Singularity to install applications
-  across platforms in a lightweight manner. \boutiques is currently used
-  by several distinct virtual research platforms, and has been used to
-  describe dozens of neuroinformatics tools.
+  proper user interfaces. Application integration efforts are often
+  replicated several times while interoperability would improve the
+  quality of the integrations. We present \boutiques, a descriptive
+  framework and command line utility to automatically import and
+  exchange applications across platforms. \boutiques adopts a rich,
+  flexible application description format that relies on Linux
+  container environments Docker and Singularity to install
+  applications across platforms in a lightweight manner. \boutiques is
+  currently used by several distinct virtual research platforms, and
+  has been used to describe dozens of applications in the
+  neuroinformatics domain.
 \end{abstract}
@@
 
-\todo{Check passive voice everywhere.}
-
 Platforms such as web services, portals, scientific gateways, workflow
@@
 third-party applications to enable data processing on various types of
-computing resources. Such applications, however, are often repeatedly
-and manually ported to different platforms whereas sharing this
-porting effort would contribute to Open Science and improve
-computational reproducibility. Meanwhile, software container systems
-such as Docker\footnote{\url{https://www.docker.com}} and
+computing resources. Such applications, however, are often manually
+and repeatedly integrated in different platforms whereas automating
+and sharing this effort would improve computational
+reproducibility~\cite{peng2011reproducible,Stodden1240} and contribute to Open
+Science. Meanwhile, software container systems such as
+Docker\footnote{\url{https://www.docker.com}} and
 Singularity~\cite{kurtzer2017singularity} have emerged to facilitate
@@
 We present \boutiques, a system to publish, integrate and execute
-containerized command-line applications across platforms (see
+command-line applications across platforms (see
 Figure~\ref{fig:diagram}).  In \boutiques, a command line is described
 using a flexible template comprising the inputs it requires and the
-outputs it produces. Inputs may be inter-dependent and passed in
-configuration files or directly on the command line. Such formal
-descriptions, simply referred to as \emph{descriptors}, link to a
-container image where the given application and all its dependencies
-are installed. \boutiques descriptors allow for automatic application
-integration in platforms, and advanced validation of input values to
-prevent errors. \boutiques also provides a simple parallelization
-framework that allows parallel applications to concurrently execute
-multiple tasks (in the remainder, tasks are called
-\emph{invocations}). \boutiques descriptors are intented to be
+outputs it produces. Inputs may be passed directly on the command line
+or through configuration files. They may also be inter-dependent, for
+instance mutually exclusive. Such formal descriptions, simply referred
+to as \emph{descriptors}, link to a container image where the given
+application is installed. \boutiques descriptors allow for automatic
+application integration in platforms, and advanced validation of input
+values to prevent errors.  \boutiques descriptors are intended to be
 produced by scientific application developers, stored alongside their
@@
 The remainder of this paper describes the \boutiques system
-(Section~\ref{sec:system}) and reports on its adoption by platforms
+and reports on its adoption by platforms
 and applications in the neuroinformatics domain, our primary field of
-interest (Section~\ref{sec:results}). It closes on a discussion and
+interest. It closes on a discussion and
 comparison with related systems.
@@
 \section{System description}
-\label{sec:system}
 
@@
 not found). At runtime, the execution platform builds the command line
-from the descriptor and values entered by the user. The platform runs
+from the descriptor and the values entered by the user. The platform runs
 the command line on the execution infrastructure, e.g., a server, a
@@
 
-The core component of the descriptor is a flexible command-line
-description represented as a simple string template complying to the
-syntax of the \texttt{bash} UNIX shell, the default shell on the major
-Linux distributions and OS X. The command-line template may contain
-placeholders for input and output values, called value keys. It may
-also encompass several commands separated by \texttt{bash} constructs
-such as semicolons, pipes (\texttt{|}) or ampersands (\texttt{\&}), to
-facilitate the embedding of mundane operations on the command line,
-for instance directory creation, input decompression or output
-archival.
+The core component of the descriptor is a command-line template
+complying to the syntax of the \texttt{bash} UNIX shell, the default
+shell on most of the Linux distributions and on OS X. The command-line
+template is a single string that may contain placeholders for input and output values, called
+value keys. It may also encompass several commands separated by
+\texttt{bash} constructs such as semicolons, pipes (\texttt{|}) or
+ampersands (\texttt{\&}), to facilitate the embedding of basic
+operations on the command line, for instance directory creation, input
+decompression or output archival.
 
@@
 \begin{verbatim}
-exampleTool_1 [STRING_INPUT] [FILE_INPUT]  | \
+exampleTool_1 [CONFIG_FILE] [STRING_INPUT] [FILE_INPUT]  | \
   exampleTool_2 [FLAG_INPUT] [NUMBER_INPUT] >> [LOG].txt
@@
 be unique but do not have to comply to any particular syntax. Note the
-use of the pipe (\texttt{|}) operator to chain applications, and of
+use of the \texttt{|} operator to chain applications, and of
 the \texttt{>>} operator to redirect the standard output to a file.
@@
 \texttt{Flag} or \texttt{File}. \texttt{File} may also represent a directory.
-Types can be restricted to a specific set of values (\texttt{String} 
-and \texttt{Number} only), to a specific range (\texttt{Number} only), or to
-integers (\texttt{Number} only).
+Types can be restricted to a specific set of values or to a specific range.
 
@@
 group must have a value; (3)~\texttt{all-or-none}: if any of the
-members are present all members must be present. Dependencies among
+members have a value then all members must have a value. Dependencies among
 inputs may also be defined regardless of a particular group: an input
@@
 \texttt{[NUMBER\_INPUT]} will be replaced by \texttt{-n=0.3} on the command
-line at execution time.
+line.
 \begin{listing}
@@
 Application outputs are files and directories that need to be
-transferred and delivered to the user once the execution is
+delivered to the user once the execution is
 complete. Outputs need to be specified so that computing platforms can
@@
 produced only when a particular flag is activated. They may also be
-(un-ordered) lists: in this case, the path template must contain a
+lists: in this case, the path template must contain a
 wildcard (\texttt{*}) matching any string of characters and defining
@@
 definition and assuming that the string input value entered by the
-user is ``foo.csv'', the string \texttt{[LOG]} will be
-replaced by \texttt{log-foo} on the command line at execution time.
+user is \texttt{foo.csv}, the string \texttt{[LOG]} will be
+replaced by \texttt{log-foo} on the command line.
 
@@
 \subsection{Configuration files}
-\label{sec:configuration-files}
+
 A large number of applications rely on configuration files rather than
@@
 Configuration files may be complex though, and specified in any
-language.  The \boutiques specification allows application developers
+language.  For this reason, \boutiques allows application developers
 to specify their own template containing input and output value
 keys. Configuration files are specific types of output files that must
-have a file template in addition to a path template that defines how
+have a file template that defines how
 they will be named and where they will be written. They may also have
@@
     "name": "Configuration file",
-    "type": "Configuration File",
+    "type": "File",
     "value-key": "[CONFIG_FILE]",
@@
         "# And here is the result",
-        "fileOutput=[OUTPUT-FILE]",
+        "fileOutput=[OUTPUT-FILE-NAME]",
         ""
@@
 Algorithm~\ref{algo:command-line} shows how the command line is
-constructed from the descriptor and the values entered by the user. It
+constructed from the descriptor and the values entered. It
 substitutes all the value keys in the command line, output path
@@
 \subsection{Invocation schema}
-\label{sec:invocation-schema}
 
@@
 defined in the invocation schema. To relieve application developers
-from the burden of having to write invocation schemas, invocation
+from the burden of having to write JSON schemas, invocation
 schemas can be generated automatically by the \texttt{bosh}
@@
 
-\boutiques does not specify a particular language to compose
-descriptors in a workflow, due to the large amount of specialized
-frameworks to do so. However, workflows can be both composed from and
-described as \boutiques descriptors: workflow engines can leverage the
-\texttt{bosh} tools to call \boutiques tools from their descriptors;
-in turn, workflows can be described as \boutiques descriptors provided
-that the execution engine is containerized. Such a ``task
-encapsulation'' model allows for a scalable and reliable execution of
-workflows expressed in a variety of languages, as detailed in
-~\cite{GLATARD2017239}.
+\boutiques does not specify a particular language to build workflows
+from descriptors, due to the large amount of specialized frameworks to
+do so. However, workflows can be both composed from and described as
+\boutiques descriptors: workflow engines can leverage the
+\texttt{bosh} tools to call \boutiques applications from their
+descriptors; in turn, workflows can be described as \boutiques
+descriptors. Such a ``task encapsulation'' model allows for a scalable
+and reliable execution of workflows expressed in a variety of
+languages, as detailed in ~\cite{GLATARD2017239}.
 
@@
 
-Applications are installed in a container image complying to the
+Applications may be installed in a container image complying to the
 Docker, Singularity or rootfs format. We intentionally support
@@
 the cloud. It is well documented, maintained and it has a rich
-ecosystem of tools to help build and run containers on most operating
+ecosystem of tools to build and run containers on most operating
 systems. Singularity is more suited for users and platforms that need
@@
 Container images are defined from their URL (rootfs) or image name in
-a Docker or Singularity index. Descriptors and container images may
-specify a working directory where the application has to be run,
-descriptors may contain a boolean indicating if the image has an entry
-point, and a hash may also be reported to accurately identify
-container images and detect updates.
+a Docker or Singularity index. Descriptors may specify a working
+directory where the application has to be run, they may indicate if
+the image has an entry point, and they may also report a hash to
+accurately identify container images and detect updates.
 
@@
 command-line utility to the user, \texttt{bosh}, which contains
-entrypoints for all core functions within \boutiques. There is
-help-text which should guide users through interacting with the
-\texttt{bosh} utility. The core tools provided by \boutiques are, the:
-validator, executor (for both launch and simulation), invocation
-schema handler, importer, and publisher. The tools exposed through the
-command-line API are also available consistently through the Python
-interface by importing the \boutiques package. Though not a component
-of the \boutiques tool chain, a Jupyter Notebook tutorial exists to
-facilitate new users getting started with \boutiques.
+entrypoints for all core functions within \boutiques. The core tools
+provided by \boutiques are, the: validator, executor (for both launch
+and simulation), invocation schema handler, importer, and
+publisher. The tools exposed through the command-line interface are also
+available consistently through a Python API by importing the
+\boutiques package. Though not a component of the \boutiques tool
+chain, a Jupyter Notebook tutorial exists to facilitate new users
+getting started with \boutiques.
 
@@
 
-\paragraph{Executor} The executor has two modes of operation: \emph{launch}, and
-\emph{simulate}. The simulate mode can generate hypothetical command lines from
-random values given the descriptor (and corresponding invocation schema) for debugging
-purposes, or display the command that would be executed given a provided valid
-invocation. The launch mode can execute command line from a \boutiques descriptor and
-a set of input values represented in JSON file complying with the invocation schema.
-It runs the command in a container provided that the required framework (e.g. Docker)
-is already installed. The executor can be used by application users to run
-applications locally, or by platforms to generate command lines to be run on the
-execution infrastructure.
+\paragraph{Executor} The executor has two modes of operation: \emph{simulate} and \emph{launch}. The simulate mode can generate hypothetical command lines from
+random values given the descriptor (and corresponding invocation
+schema) for debugging purposes, or display the command that would be
+executed given a provided valid invocation. The launch mode can
+execute command line from a \boutiques descriptor and a set of input
+values represented in JSON file complying with the invocation schema.
+It runs the command in a container provided that the required
+framework (e.g. Docker) is installed. The executor can be used
+by application users to run applications locally, or by platforms to
+generate command lines to be run on the execution infrastructure.
 
 \paragraph{Invocation Schema Handler} The invocation schema handler can create an
-invocation schema (see Section~\ref{sec:invocation-schema}) from a \boutiques descriptor
-and validate input data against it using a regular JSON validator. It can be used to add
-invocation schemas to existing descriptors. It is used by the executor if no invocation
-schema is present in the \boutiques descriptor being deployed.
+invocation schema from a \boutiques descriptor and validate input data
+against it using a regular JSON validator. It can be used to add
+invocation schemas to existing descriptors. It is used by the executor
+if no invocation schema is present in the \boutiques descriptor being
+deployed.
 
@@
 \paragraph{Publisher} As \boutiques has primarily been adopted in the neuroinformatics
-community, the publisher gets a further description (such as author, website, etc.) of
-the described tool, and adds an index to it on
-NeuroLinks\footnote{\url{https://brainhack101.github.io/neurolinks}}, a repository
-containing links to useful resources and tools. This enables others to find and use the
-descriptor for your tool easily. This functionality could be extended to
-new repositories for other domains, such as Bioconductor for bioinformatics.
+community, the publisher gets a further description (such as author,
+website, etc.) of the described application, and adds an index to it
+on
+NeuroLinks\footnote{\url{https://brainhack101.github.io/neurolinks}},
+a repository containing links to neuroinformatics resources and
+tools. This functionality could be extended to new repositories for
+other domains, such as Bioconductor for
+bioinformatics~\cite{gentleman2004bioconductor}.
 
 \section{Results}
-\label{sec:results}
 
@@
 
-\boutiques application import and execution is currently supported by several
-platforms which aid in execution and tool management. These are enumerated
-below.
+The import and execution of \boutiques applications is currently
+supported in the platforms enumerated below.
 
@@
 CBRAIN (\url{http://github.com/aces/cbrain}) ~\cite{SHER-14} is a web
-platform that can process data distributed into multiple storage
+platform to process data distributed into multiple storage
 locations on computing clusters and clouds. CBRAIN offers transparent
@@
 including 6 clusters of the Compute Canada high-performance computing
-infrastructure (totaling more than 100,000 computing cores and 40PB of
+infrastructure (totaling more than 100,000 computing cores and 40~PB of
 disk storage) and Amazon EC2. CBRAIN transiently stores about 10
-million files representing over 50TB distributed in 42 servers. 51
+million files representing over 50~TB distributed in 42 servers. 51
 public data processing applications are integrated and over 340,000
@@
     classes are generated on-the-fly when CBRAIN starts. This mode
-    allows CBRAIN developers update all \boutiques applications at
+    allows CBRAIN developers to update all \boutiques applications at
     once by editing the templates. However, it does not allow for
@@
 periodically scans working directories, submits the requested
-sub-tasks, and writes back the invocation id in the same
+sub-tasks, and writes back an invocation identifier in the same
 directory. This parallelization model is simple, the application only
@@
 Nipype
-(\url{http://nipype.readthedocs.io/en/latest/})~\cite{gorgolewski2011nipype}
+(\url{http://nipype.readthedocs.io/en/latest})~\cite{gorgolewski2011nipype}
 is a workflow engine widely used in neuroinformatics. Nipype workflows
-can be composed from \boutiques tools using the Python API. As an
-example, we implemented NipBIDS, a Nipype workflow to process BIDS
-datasets using BIDS apps imported as \boutiques tools. NipBIDS iterates
-participant analyses on all the subjects found in a BIDS dataset and
-runs a group analysis if requested. It is available at
-\url{https://github.com/big-data-lab-team/sim/tree/master/sim/other_wf_examples/nipype}.
+can be composed from \boutiques applications using the Python API. As
+an example, we implemented
+NipBIDS\footnote{\url{https://github.com/big-data-lab-team/sim/tree/master/sim/other_wf_examples/nipype}},
+a Nipype workflow to process BIDS datasets using BIDS apps imported as
+\boutiques applications. NipBIDS iterates participant analyses on all
+the subjects found in a BIDS dataset and runs a group analysis if
+requested.
 
@@
 jobs on EGI grid and cloud resources.  As of October 2017, VIP counts
-more than 1000 registered users and a growing number of available
+more than 1,000 registered users and a growing number of available
 applications.
-%% gk: I guessed 50 more in the last year, since 1000 is a nice number, is this accurate?
 
@@
 with the MOTEUR~\cite{GLAT-08} engine.  As of today, Boutiques is
-supported through a "Boutiques application importer" tool, which
-parses the JSON descriptor and automatically generates the
-corresponding application workflow and the wrapper script that
-handles, among other things, the execution of the command line.
-Application workflows enable (1) iterations on input lists, (2) the
-generation of parallel tasks, and (3) the concatenation of multiple
-tools. For examples, the applications used in the MICCAI challenges
-described below required workflows to evaluate results using the
-metrics defined by the challenge. Tool concatenation is 
-handled at the importer level based on pre-defined workflow
-templates.
-
-\subsection{Ported applications}
+supported through an importer tool that parses the JSON descriptor and
+automatically generates the corresponding application workflow and the
+wrapper script that handles, among other things, the execution of the
+command line.  In VIP, application workflows enable (1) iterations on
+input lists, (2) the generation of parallel tasks, and (3) the
+concatenation of multiple applications. For example, the applications
+used in the MICCAI challenges described below required workflows to
+evaluate results using the metrics defined by the
+challenge. Application concatenation is handled at the importer level
+based on pre-defined workflow templates.
+
+\subsection{Integrated applications}
 
@@
 
-Dozens of neuroinformatics tools were ported to CBRAIN or VIP using
-\boutiques. The main ones are described below. Several \boutiques
-descriptors were published in
-Neurolinks\footnote{\url{https://brainhack101.github.io/neurolinks}}
-through the \texttt{bosh} publisher.
+Dozens of neuroinformatics applications were integrated in CBRAIN or
+VIP using \boutiques. The main ones are described below. Several
+\boutiques descriptors were published in Neurolinks through the
+\texttt{bosh} publisher.
 
@@
 
-\paragraph{FSL} Several MRI analysis tools from the FMRIB Software
-library (FSL~\cite{jenkinson2012fsl}) were ported to CBRAIN using
-\boutiques: BET, fsl\_anat, FAST and FIRST. Descriptors are available
+\paragraph{FSL} Several MRI analysis applications from the FMRIB Software
+library (FSL~\cite{jenkinson2012fsl}) were integrated in CBRAIN using
+\boutiques: BET, fsl\_anat, FAST and FIRST. Descriptors are
 on Neurolinks.
@@
 \paragraph{Anima}
-\texttt{animaN4BiasCorrection}, an ITK-based bias field correction tool from the
+\texttt{animaN4BiasCorrection}, an ITK-based bias field correction application from the
 Anima\footnote{\url{https://github.com/Inria-Visages/Anima-Public/wiki}}
-project was also made available in VIP. The import process was
-straightforward.
+project was made available in VIP through \boutiques.
 
@@
 
-\paragraph{Niak} The Niak fMRI pre-processing pipeline, executed with the Pipeline System
+\paragraph{Niak} The Niak fMRI pre-processing pipeline~\cite{bellec2011neuroimaging}, executed with the Pipeline System
 for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
-in CBRAIN through \boutiques. The integration uses the parallelization
-mechanism described earlier so that even the invocations processing a
-single subject can be parallelized. It also allows CBRAIN to leverage
-the efficient agent model used in PSOM, as described
+in CBRAIN through \boutiques. The integration uses the CBRAIN
+sub-tasking mechanism described earlier so that even the invocations
+processing a single subject can be parallelized. It also allows CBRAIN
+to leverage the efficient agent model used in PSOM, as described
 in~\cite{GLAT-16}. The integration required some work in PSOM to
@@
 application. The resulting CBRAIN plugin is available at
-\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It can be used in
-other platforms that support the parallelization property. Descriptors
-are available on Neurolinks.
+\url{https://github.com/SIMEXP/cbrain-plugins-psom}.  Descriptors
+are on Neurolinks.
 
@@
 GinFizz\footnote{\url{https://github.com/thomashirsch/ginfizz}}
-Nipype-based fMRI pre-processing pipeline in VIP with \boutiques. \todo{Sorina, Pascal, a few lines on the technical deployment?}
+Nipype-based fMRI pre-processing pipeline in VIP with \boutiques.  A
+few technical issues coming from the management of users in Docker
+containers had to be addressed: to enable the execution in \boutiques
+we had to override the permissions of files and folders in the GinFizz
+container. We also had to install all the GinFizz pipeline components
+in a single container while multiple ones were used by the application
+initially.
+
 %% Ginfizz performs multiple
@@
 \paragraph{MRTrix3}
-A few tools from the MRtrix3
+A few applications from the MRtrix3
 package~\cite{tournier2012mrtrix}\footnote{\url{http://www.mrtrix.org/}} for diffusion MRI processing were also made available in VIP,
-as well as a script developed at Creatis, which combines MRtrix3 and
-FSL tools.
+as well as a pipeline developed at Creatis, which combines MRtrix3 and
+FSL applications.
 
@@
 \paragraph{CreaPhase}
-A phase contrast simulator called CreaPhase, developed at Creatis, was
-integrated in VIP through \boutiques. The CreaPhase inputs had
-certain particularities (some nedeed to be enclosed in simple quotes,
+The CreaPhase phase-contrast simulator~\cite{WEBE-17}, developed at
+Creatis, was integrated in VIP through \boutiques. The inputs had
+certain particularities (some needed to be enclosed in simple quotes,
 other were vectors of variable size enclosed in brackets) that
 required the post-processing of the wrapper script generated by the
-Boutiques importer. \todo{Sorina: reference?}
-
-\paragraph{ODIN} The Odin MRI simulator~\cite{jochimsen2004odin}\footnote{\url{http://od1n.sourceforge.net}} was imported 
-in VIP through \boutiques. Since Odin may require important amounts of
-computing resources (the common use-case consists in iterating on a
-large number of inputs files), Odin is executed by VIP on the EGI grid
-resources, which currently do not support Docker. Consequently, the
-Odin executable was exported from the Docker image and the Odin
-wrapper script was modified accordingly.
+\boutiques importer.
+
+\paragraph{ODIN} The Odin MRI simulator~\cite{jochimsen2004odin}\footnote{\url{http://od1n.sourceforge.net}} was integrated
+in VIP through \boutiques. Since Odin requires important amounts of
+computing resources it is executed by VIP on the EGI grid that
+currently does not support Docker. The Docker image was used just for
+compilation and testing; the Odin executable was extracted from the
+Docker image and the Odin wrapper script was modified accordingly.
 
@@
 
-BIDS apps, an effort for the adoption of the Brain Imaging Data Structure (BIDS)
-in common neuroimaging pipelines, require a standardized set of input
-and output parameters~\cite{gorgolewski2017bids}. We developed a tool
-as part of the \boutiques importer to generate a descriptor for any
-such BIDS app. We validated this tool by importing BIDS apps
-containing the Statistical Parameter Mapping toolbox
-(SPM)~\cite{penny2011statistical} and the ndmg pipeline mentioned
-above. Descriptors are available on Neurolinks.
+BIDS apps~\cite{gorgolewski2017bids}, an effort for the adoption of
+the Brain Imaging Data Structure (BIDS) in common neuroimaging
+pipelines, require a standardized set of input and output
+parameters. We developed a tool as part of the \boutiques importer to
+generate a descriptor for any such BIDS app. We validated this tool by
+importing BIDS apps containing the Statistical Parameter Mapping
+toolbox (SPM)~\cite{penny2011statistical} and the ndmg pipeline
+mentioned above. Descriptors are on Neurolinks.
 
@@
 challenge~\cite{hatt2018}). The
-pipelines were ported to VIP and executed on 205 subjects in a few
-weeks only, which would not have been possible without
-\boutiques. Some pipelines had to be adjusted manually once ported to
+pipelines were integrated in VIP and executed on 205 subjects in a few
+weeks only. Some pipelines had to be adjusted manually once integrated in
 the platform for the following reasons:
@@
   in our setup. We solved the issue by installing the data in a
-  directory of the host server that we mounted in the container. Such
-  platform-specific configurations could not possibly be supported by
-  \boutiques.
+  directory of the host server that we mounted in the container.
 \item A pipeline wrote more than 10 GB of intermediate data in a
@@
 \end{enumerate}
-The pipelines did not use much of the expressiveness features offered
-by \boutiques since the instructions given to the challengers required
-that all the pipeline parameters were hardcoded, to facilitate
-execution. Two custom properties (\texttt{vip:miccai-challenger-email}
-and \texttt{vip:miccai-challenge-team-name}) were added to the
+Two custom properties (\texttt{vip:miccai-challenger-email}
+and \texttt{vip:miccai-challenge-team-name}) were also added to the
 \boutiques descriptor to help post-process results in the specific
@@
 \section{Discussion}
-With \boutiques, developers can port their applications once and execute
-them in several platforms. \boutiques removes the technological dependency to a
-particular platform and facilitates migration.  Although the
-motivating use cases were taken from neuroinformatics, our primary
-field of interest, nothing prevents the system from being used in a
-variety of other domains.
+With \boutiques, developers can integrate their applications once and
+execute them in several platforms. \boutiques removes the
+technological dependency to a particular platform and facilitates
+application migration.  Although the motivating use cases were taken
+from neuroinformatics, our primary field of interest, nothing prevents
+the system from being used in other domains.
 
@@
 additions such as invocation schemas and dependencies among
-inputs. Extending the \boutiques descriptor has two main goals. First,
-accurately describing applications improves validation: incorrect
-input values and execution results are more precisely detected when
-the application descriptor is comprehensive, which is essential to
-improve the reliability and user-friendliness of execution
-platforms. Secondly, a rich descriptor schema reduces the need for
-custom application wrappers, for instance to create configuration
-files from command-line options, as even convoluted command lines can
-be described. This is particularly useful for containerized
-applications since adding a wrapper requires an update of the
-container image.
+inputs. Extending the \boutiques descriptor has two main goals: (1)
+validation: incorrect input values and execution results are more
+precisely detected when the application descriptor is
+comprehensive; (2) automation: a rich descriptor schema reduces the need for
+custom application wrappers, which is particularly useful for containerized
+applications.
 
@@
 supporting as many features as possible in the \texttt{bosh} executor
-so that only the following steps need to be implemented to support
-\boutiques in a platform, regardless of the complexity of the
-descriptor:
+so that only the following steps need to be implemented in a platform,
+regardless of the complexity of the descriptor:
 \begin{enumerate}
@@
 \end{enumerate}
-In particular, advanced validation features such as dependencies
-between inputs are embedded in the regular JSON validation performed
-at step 2, without requiring the platform to support the related
+In particular, command-line generation and advanced validation
+features such as dependencies between inputs are embedded in
+\texttt{bosh}, without requiring the platform to support the related
 descriptor properties.
@@
 \boutiques intentionally does not provide a workflow language to
-compose tools, as this is already offered by several workflow
-engines. Workflows can be either composed from \boutiques tools, as we
-illustrated with the Nipype and MOTEUR engines, or described as
-\boutiques applications, as we demonstrated with the Niak fMRI
-pre-processing pipeline. Furthermore, the CBRAIN platform has a
-sub-tasking mechanism that allows \boutiques tools to submit tasks,
-which is used to parallelize workflows wrapped as \boutiques
-tools. Based on the CBRAIN experience, we may specify the sub-tasking
-mechanism in \boutiques so that other platforms can benefit from it.
-This model is powerful because it shields the \boutiques specification
-from specific workflow constructs and allows a wide range of workflow
-engines to be described and used uniformly.
+compose applications, as this is already possible with numerous
+workflow engines. Workflows can be either composed from \boutiques
+applications, as we illustrated with the Nipype and MOTEUR engines, or
+described as \boutiques applications, as we demonstrated with the Niak
+fMRI pre-processing pipeline. Furthermore, the CBRAIN platform has a
+sub-tasking mechanism that allows \boutiques applications to submit
+tasks, which is used to parallelize workflows wrapped as \boutiques
+applications. Based on the CBRAIN experience, we may specify the
+sub-tasking mechanism in \boutiques so that other platforms can
+benefit from it.  This model is powerful because it shields the
+\boutiques specification from specific workflow constructs and it allows
+a wide range of workflow engines to be described and used uniformly.
 
@@
 
-\boutiques helps computational reproducibility, in Peng's
-sense~\cite{peng2011reproducible}, through containers and formal
+\boutiques helps computational reproducibility through containers and formal
 command-line descriptions. With \boutiques, complete sets of
 applications could be easily migrated across execution platforms,
-including high-performance computing platforms and individual laptops,
+including high-performance computing clusters and individual laptops,
 to reproduce analyses. The \boutiques descriptor describes the
@@
 versions or hardware platforms. For instance, containers may not
-execute consistently on different x86\_64 CPUs (e.g. Intel and AMD)
-when software compiled with architecture-specific flags is involved
-(see for instance GCC's \texttt{-march} flag).
+execute consistently on different CPUs complying to the x86\_64
+architecture (e.g. Intel and AMD) when the application is compiled with
+architecture-specific flags such as GCC's \texttt{-march}.
 
@@
 command-line applications. While such applications cover a large
-fraction of the tools involved in scientific data processing, other
-types of programs exist such as web services, interactive tools, and
+fraction of the applications involved in scientific data processing, other
+types of programs exist such as web services, interactive applications, and
 applications with a graphical user interface (GUI).  Such application
@@
 non-interactive ones through configuration files. Finally, nothing
-prevents a \boutiques tool from popping up a window for a user to
+prevents a \boutiques application from popping up a window for a user to
 provide input through a GUI. This should, however, be specified
@@
 A few limitations remain that should be addressed in the
-future. First, \boutiques moves the application porting bottleneck
+future. First, \boutiques moves the application integration bottleneck
 from integration to validation. Using \boutiques, functions can be
@@
 developers rather than platforms, which is a good thing to reduce
-application porting bottlenecks, nothing prevents developers to embed
+application integration bottlenecks, nothing prevents developers to embed
 malicious code in their container at any stage of the process,
@@
 The Common Workflow Language
-(CWL\footnote{\url{http://www.commonwl.org}}) is the piece of related
-work most closely related to \boutiques as it provides a formal way to
-describe containerized applications. In particular, CWL's Command Line
-Tool Description overlaps with the \boutiques descriptor. This Section
+(CWL\footnote{\url{http://www.commonwl.org}}) is the work most closely
+related to \boutiques as it provides a formal way to describe
+containerized applications. In particular, CWL's Command Line Tool
+Description overlaps with the \boutiques descriptor. This Section
 highlights the main differences between CWL and \boutiques, based on
@@
 The following differences are conceptual in the sense that they may
-not be easily addressed by CWL or \boutiques developers without deeply
+not be easily addressed in CWL or \boutiques without deeply
 refactoring the frameworks.
@@
 decompression using the \texttt{tar} command in addition to the main
-tool command. Importantly, \boutiques' template system allows
+application command. Importantly, \boutiques' template system allows
 supporting configuration files.
@@
 depending on the context. \boutiques tries to limit the complexity of
-the specification to facilitate its support by platforms where tools
+the specification to facilitate its support by platforms where applications
 will be integrated.
@@
 \begin{itemize}
-\item CWL tools have to write in a specific set of directories called
+\item CWL applications have to write in a specific set of directories called
 ``designated output directory'', ``designated temporary directory''
-and ``system temporary directory''. Tools are informed of the location
+and ``system temporary directory''. Applications are informed of the location
 of such directories through environment variables. Having to write in
-specific directories is problematic because tools have to be modified
+specific directories is problematic because applications have to be modified
 to enable that. In \boutiques, the path of output files is defined
@@
 activated. This is a very useful feature to improve validation, in
-particular when tools have a lot of options.
+particular for applications with a lot of options.
 \item In \boutiques, named
@@
 from the Canada First Research Excellence Fund, awarded to McGill
-University for the Healthy Brains for Healthy Lives initiative.
+University for the Healthy Brains for Healthy Lives initiative. We also
+thank the developers of all the applications described with \boutiques.
 

### 2017-11-07 boutiquespaper.tex
Rewrote abstract

@@
 
-\begin{abstract}Integrating applications in execution platforms such as web
-  portals and workflow engines is critical to enable their sharing
-  among scientific communities and their efficient execution on
-  computing infrastructures. However, application integration remains
-  a costly human effort that consists of 1)~installing the application
-  on the target infrastructure, 2)~describing the application in a
-  format compatible with the execution platform, and 3)~generating
-  proper user interfaces. Application integration efforts are often
-  replicated several times while interoperability would improve the
-  quality of the integrations. We present \boutiques, a descriptive
-  framework and command line utility to automatically import and
-  exchange applications across platforms. \boutiques adopts a rich,
-  flexible application description format that relies on Linux
-  container environments Docker and Singularity to install
-  applications across platforms in a lightweight manner. \boutiques is
-  currently used by several distinct virtual research platforms, and
-  has been used to describe dozens of applications in the
-  neuroinformatics domain.
+\begin{abstract}
+We present \boutiques, a system to automatically publish, integrate
+and execute applications across computational platforms. \boutiques
+applications are installed through software containers described in a
+rich and flexible JSON language. A set of core tools facilitate the
+construction, validation, import, execution, and publishing of
+applications. \boutiques is currently supported by several distinct virtual
+research platforms, and it has been used to describe dozens of
+applications in the neuroinformatics domain. We expect \boutiques to
+improve the quality of application integration in computational
+platforms, to reduce redundancy of effort, to contribute to computational
+reproducibility, and to foster Open Science.
 \end{abstract}
@@
 
-Platforms such as web services, portals, scientific gateways, workflow
-engines and virtual research environments commonly integrate
-third-party applications to enable data processing on various types of
-computing resources. Such applications, however, are often manually
-and repeatedly integrated in different platforms whereas automating
-and sharing this effort would improve computational
-reproducibility~\cite{peng2011reproducible,Stodden1240} and contribute to Open
-Science. Meanwhile, software container systems such as
+Computational platforms such as web services, portals, scientific
+gateways, workflow engines and virtual research environments commonly
+integrate third-party applications to enable various types of data
+processing. Applications, however, are often manually and
+repeatedly integrated whereas automating and
+sharing this effort would improve computational
+reproducibility~\cite{peng2011reproducible,Stodden1240} and contribute
+to Open Science. Meanwhile, container systems such as
 Docker\footnote{\url{https://www.docker.com}} and
@@
 \end{figure}
+
+%% Application integration consists of 1)~installing the application
+%%   on the target infrastructure, 2)~describing the application in a
+%%   format compatible with the execution platform, and 3)~generating
+%%   proper user interfaces.
 

### 2017-11-07 boutiquespaper.tex
Updates following comments from Charles and Guillaume

@@
 \affil[6]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA}
-\affil[7]{Wellcome Trust Centre for Neuroimaging}
+\affil[7]{Wellcome Trust Centre for Neuroimaging, London, UK}
 \affil[8]{Department of Psychology, Stanford University, Stanford, California, USA}
-\affil[9]{Partners MS Center, Brigham and Women's Hospital, Brookline, Massachusetts, USA}
+\affil[9]{Center for Neurological Imaging, Department of Radiology, Brigham and Women's Hospital, Boston, Massachusetts, USA}
 \affil[10]{Compute Canada}
@@
 Custom properties may be added to the Boutiques specification without
-restriction. Custom properties are corralled in a specific JSON object
+restriction. Custom properties are grouped together in a specific JSON object
 to facilitate validation. They may be useful to implement
@@
 questions.  SPINE enables distributed data collection and management,
-as well as experiment design, execution, and review. Boutiques serves
+as well as experiment design, execution, and review. Boutiques will serve
 as SPINE's algorithm and workflow repository, and enables unequivocal
@@
 \paragraph{CreaPhase}
-The CreaPhase phase-contrast simulator~\cite{WEBE-17}, developed at
+The CreaPhase phase-contrast simulator, developed at
 Creatis, was integrated in VIP through \boutiques. The inputs had
@@
 generate a descriptor for any such BIDS app. We validated this tool by
-importing BIDS apps containing the Statistical Parameter Mapping
+importing BIDS apps containing the Statistical Parametric Mapping
 toolbox (SPM)~\cite{penny2011statistical} and the ndmg pipeline
@@
 of the execution. We plan to leverage the provenance format being
-defined by the Neuro-Imaging Data Model
-Workflow~\cite{ghosh2017capturing} initiative for this purpose.
+defined by the NeuroImaging Data
+Model-Workflow~\cite{ghosh2017capturing} initiative for this purpose.
 

### 2018-01-19 answers.tex
Added Shawn as co-author, implemented reviewers' comments, drafted letter.

\documentclass[a4]{article}
\usepackage{graphicx}
\usepackage{color}
\usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links

\usepackage{hyperref}
\usepackage{xspace}
\newenvironment{review}%          environment name
{\textbf{Reviewer comment:}\begin{quote}}% begin code
{\end{quote}}%  
\newenvironment{answer}%          environment name
{\textbf{Answer:}\begin{small}\begin{quote}}% begin code
{\end{quote}\end{small}}%

\newcommand{\todo}[1]{\color{red}#1\color{black}}

\newcommand{\revised}[1]{\color{blue} #1\color{black}\xspace}

\begin{document}

\section{Reviewer \#1}

\begin{review}
This is a very clear and readable manuscript describing an apparently
simple but powerful framework for constructing automated workflows
from command-line-driven applications. While reading the manuscript, I
kept thinking "this sounds very like CWL", so I was pleased to see an
extended discussion of the similarities and differences between the
two systems. In general I think it is helpful to have a few competing
solutions in domains such as this. My only, small, critique, is that
the words "command-line" should be present in (ideally) the title and
(at least) the abstract in order to qualify "application(s)".
\end{review}

\begin{answer}
We thank the reviewer for his valuable feedback. We added the words
``command-line'' in the title and abstract as per the reviewer's
suggestion.
\end{answer}

\section{Reviewer \#2}

\begin{review}
Boutiques is a specification that abstracts command-line programs,
their arguments, and their expected outputs. The goal of the
specification is to simplify application integration, reduce code
redundancy, and support reproducible science. Following the Boutiques
specification, a command-line program is documented in a descriptor
file, which follows the JSON specification. A descriptor file
specifies the arguments and expected outputs of a program. The file
includes a template of the command-line instruction, the location of a
software container that includes the program, information about each
parameter (e.g., what type of argument it expects, if any), and
relationships within groups of related parameters (e.g., parameters in
a group might be mutually exclusive). Potential arguments to a program
may be validated against the information in that program's
descriptor. Early validation is a key feature of the Boutiques
specification, because it identifies issues to the user prior to the
program's runtime. The command-line template is also advantageous
because it allows for shell functions to be used, like piping and
input/output redirection. As a side effect of the level of detail
required by the Boutiques specification, the descriptor file also
serves as precise usage documentation of a program.

Moreover, the paper describes the Boutiques core tools, which I
understood to be a Python implementation of the Boutiques
specification. These core tools include `bosh`, a command-line program
which enables users to validate Boutiques descriptors and invocation
files, easily run containerized programs, convert BIDS apps to
Boutiques descriptors, and publish their Boutiques descriptors to the
NeuroLinks website. `bosh` can run containers in Docker, Singularity,
or rootfs format. This feature satisfies the needs of both local
execution and execution on high performance computing clusters. The
inclusion of containers in the Boutiques specification lessens the
burden on the user because the user does not have to install the
software on the host machine.
\end{review}

\begin{answer}
  We thank the reviewer for his thorough review of our manuscript.
\end{answer}

\begin{review}
I would recommend a minor revision of this paper. The descriptor
schema and core tools were described and reasoned well, but the
overall goal of Boutiques could be clarified earlier in the paper. I
found the last sentence of the section titled "Other frameworks" to be
a concise and informative summary of the goal of Boutiques and the
advantages of Boutiques over other, similar frameworks. The authors
might consider adding a similar sentence to the introduction of the
paper.
\end{review}

\begin{answer}
The main features of Boutiques mentioned in section ``Other
frameworkds'' are (1) its support for containerized applications and
(2) its independence from any particular programming language. While
(1) was already mentioned in the introduction, (2) indeed wasn't. We
updated the following sentence in the introduction, to address this
issue:

\begin{quote}
  Such formal descriptions, simply referred
to as \emph{descriptors}, \color{blue}can be parsed in any programming language,
can describe command lines regardless of their implementation,\color{black}
and link to a container image where the application is
installed.
\end{quote}

\end{answer}

\begin{review}
The work could also benefit from minor clarification of the descriptor
specification and minor updates to the core tools. Overall, I found
the descriptor schema well-documented, intuitive, modular, and
relatively complete (keeping in mind the ability to specify custom
parameters). I also found the various listings in the paper and the
online documentation helpful. I installed Boutiques version 0.5.3
(Python 3.6.3 on OS X 10.11.6) to test the minimal Boutiques
descriptor in Listing 5. This descriptor passed the validation step,
but when attempting to conduct a simulation run of the descriptor
given an input JSON file with the contents `{"param": 5}`, a
`KeyError` was produced, indicating that the key "container-image" was
not present in the descriptor. This led me to wonder whether
"container-image" is a required field in the descriptor schema and
whether `bosh` can execute on bare metal.
\end{review}

\begin{answer}
  The reviewer is right that no test of bare metal tools were included
  in the code base and that `bosh exec` was crashing on them. This was
  fixed in \url{https://github.com/boutiques/boutiques/pull/200}. We
  thank the reviewer for spotting this.
\end{answer}

\begin{review}
When trying to run the example in
https://github.com/boutiques/boutiques/tree/https://github.com/boutiques/boutiques/tree/5678d6f84bc1863741c6b8c595f037f622936c5c/tools/python/boutiques/schema/examples/example1/tools/python/boutiques/schema/examples/example1,
validation of the invocation and descriptor functioned as
described. For instance, validation failed when `enum\_input` and
`num\_input`, mutually exclusive parameters, were both
specified. Validation did not fail, however, when attempting to run a
'simulation' or 'launch' with the same invalid inputs. The authors
should consider including the validation step when running the program
with `bosh exec`.
\end{review}

\begin{answer}
  We thank the reviewer for spotting this. This was fixed in
  \url{https://github.com/boutiques/boutiques/pull/200}.
\end{answer}

\begin{review}
Furthermore, the feature of `bosh` to upload descriptors to a common repository (i.e., NeuroLinks) is an important one. While links to descriptors are certainly useful, I would like to suggest a common repository of the descriptor files themselves. Such a repository would reduce redundancy in coding effort because contributors could see which programs already have descriptors. Upstream software maintainers could maintain their Boutiques descriptors in such a repository. The burden on new users might also be reduced because they would have access to all Boutiques descriptors in a single location. If a common repository like this were in place, `bosh` might be extended to accept descriptors in a more abstract form. For example, instead of specifying a path to an existing descriptor file, a user could specify "fsl.bet.v1", and `bosh` could pull the relevant descriptor from the common repository. A similar effect might be achieved using the existing JSON file that lists the
tools on the NeuroLinks website.

\todo{to be discussed.}

In conclusion, Boutiques is a framework that flexibly specifies command-line programs and improves reproducibility through the use of software containers. Other frameworks can leverage Boutiques to create self-contained workflows that only depend on a container implementation like Docker or Singularity.
\end{review}

\todo{Make a new release and push it to pip}

\end{document}

### 2018-01-19 boutiquespaper.tex
Declaration of conflicts of interests
Adjusted style for pre-print
Added Shawn as co-author, implemented reviewers' comments, drafted letter.

@@
 
-%%% Journal toggle; only specific options recognised.
-%%% (Only "gigascience" and "general" are implemented now. Support for other journals is planned.)
-\journal{gigascience}
-
 \usepackage{graphicx}
 \usepackage{siunitx}
-\papercat{Paper}
 
@@
 
-\title{Boutiques: a flexible framework for automated application integration in computing platforms}
+\title{Boutiques: a flexible framework to integrate command-line applications in computing platforms}
 
@@
 \author[5]{Axel Bonnet}
+\author[2,3]{Shawn T Brown}
 \author[5]{Sorina Camarasu-Pop}
@@
 
-\begin{frontmatter}
 \maketitle
@@
 We present \boutiques, a system to automatically publish, integrate
-and execute applications across computational platforms. \boutiques
-applications are installed through software containers described in a
-rich and flexible JSON language. A set of core tools facilitate the
-construction, validation, import, execution, and publishing of
-applications. \boutiques is currently supported by several distinct virtual
-research platforms, and it has been used to describe dozens of
-applications in the neuroinformatics domain. We expect \boutiques to
-improve the quality of application integration in computational
-platforms, to reduce redundancy of effort, to contribute to computational
-reproducibility, and to foster Open Science.
+and execute command-line applications applications across
+computational platforms. \boutiques applications are installed through
+software containers described in a rich and flexible JSON language. A
+set of core tools facilitate the construction, validation, import,
+execution, and publishing of applications. \boutiques is currently
+supported by several distinct virtual research platforms, and it has
+been used to describe dozens of applications in the neuroinformatics
+domain. We expect \boutiques to improve the quality of application
+integration in computational platforms, to reduce redundancy of
+effort, to contribute to computational reproducibility, and to foster
+Open Science.
 \end{abstract}
@@
 \end{keywords}
-\end{frontmatter}
 
@@
 
-Computational platforms such as web services, portals, scientific
+Computational platforms such as web services, portals, science
 gateways, workflow engines and virtual research environments commonly
@@
 processing. Applications, however, are often manually and
-repeatedly integrated whereas automating and
+repeatedly integrated in such platforms whereas automating and
 sharing this effort would improve computational
@@
 instance mutually exclusive. Such formal descriptions, simply referred
-to as \emph{descriptors}, link to a container image where the given
-application is installed. \boutiques descriptors allow for automatic
-application integration in platforms, and advanced validation of input
-values to prevent errors.  \boutiques descriptors are intended to be
-produced by scientific application developers, stored alongside their
-application, indexed by common repositories, and consumed by execution
-platforms.  A set of core tools facilitate the construction,
-validation, import, execution, and publishing of \boutiques
-descriptors.
+to as \emph{descriptors}, can be parsed in any programming language,
+can describe command lines regardless of their implementation,
+and link to a container image where the application is
+installed. \boutiques descriptors allow for automatic application
+integration in platforms, and advanced validation of input values to
+prevent errors.  \boutiques descriptors are intended to be produced by
+application developers, stored alongside their application, indexed by
+common repositories, and consumed by execution platforms.  A set of
+core tools facilitate the construction, validation, import, execution,
+and publishing of \boutiques descriptors.
 
@@
 
+\section{Competing interests}
+
+The authors declare that they have no competing interests.
+
 \bibliography{biblio}

### 2018-01-26 answers.tex
More revisions and answers

@@
 \newcommand{\revised}[1]{\color{blue} #1\color{black}\xspace}
+
+\usepackage{hyperref}
+\hypersetup{colorlinks=true,linkcolor=blue,citecolor=black,filecolor=black,urlcolor=black}
 
@@
 When trying to run the example in
-https://github.com/boutiques/boutiques/tree/https://github.com/boutiques/boutiques/tree/5678d6f84bc1863741c6b8c595f037f622936c5c/tools/python/boutiques/schema/examples/example1/tools/python/boutiques/schema/examples/example1,
+\url{https://github.com/boutiques/boutiques/tree/5678d6f84bc1863741c6b8c595f037f622936c5c/tools/python/boutiques/schema/examples/example1/tools/python/boutiques/schema/examples/example1},
 validation of the invocation and descriptor functioned as
@@
 tools on the NeuroLinks website.
+\end{review}
 
-\todo{to be discussed.}
-
-In conclusion, Boutiques is a framework that flexibly specifies command-line programs and improves reproducibility through the use of software containers.
-Other frameworks can leverage Boutiques to create self-contained workflows that only depend on a container implementation like Docker or Singularity.
-\end{review}
+\begin{answer}
+We thank the reviewer for such a relevant comment. We deliberately
+chose to not publish tools to a common repository because we believe
+that they should be stored in the tool repositories so that they can
+be updated and tested by the tool developers as their implementation
+evolves. Instead, the NeuroLinks index provides a way to reference
+descriptors without having to store them directly. The search
+functionality, however, is a great one and we have discussed the following 
+implementation in \texttt{bosh} over the NeuroLinks index.
+\begin{enumerate}
+\item Make Boutiques entries in Neurolinks immutable:
+  \begin{enumerate}
+  \item modify bosh publish to commit a URL of a particular commit of the descriptor instead of the latest version.
+  \item add the version number of the identifier.
+  \item write guidelines to accept PRs about Boutiques descriptors in Neurolinks.
+  \end{enumerate}
+\item Extend \texttt{bosh} by implementing:
+  \begin{enumerate}
+  \item \texttt{bosh search} that greps in \url{https://raw.githubusercontent.com/brainhack101/neurolinks/master/jsons/tool.json} and returns a table containing the identifiers and (truncated) descriptions of the Boutiques tools.
+  \item \texttt{bosh get} that downloads a descriptor based on its neurolinks identifier.
+  \item \texttt{bosh launch <simulate|exec> <neurolinks\_id>}  that calls \texttt{bosh get} to get the descriptor and launch the tool.
+  \end{enumerate}
+\end{enumerate}
+\end{answer}
+This will be implemented as part of issue \href{https://github.com/boutiques/boutiques/issues/34}{34}. We have updated the manuscript accordingly. The paragraph on the Publisher now reads as follows:
+\begin{quote}
+\emph{Publisher} As Boutiques has primarily been adopted in the neuroinformatics
+community, the publisher gets a further description (such as author,
+website, etc.) of the described application, and adds an index to it
+on
+NeuroLinks\footnote{\url{https://brainhack101.github.io/neurolinks}},
+a repository containing links to neuroinformatics resources and
+tools. We intentionally opted for publishing descriptors to an index
+such as NeuroLinks rather than to a common Boutiques repository so
+that descriptors remain in the tool repositories and can be maintained
+by the tool developers directly. We are currently extending
+\texttt{bosh} to enable users to search for Boutiques tools in
+NeuroLinks using the command-line or Python API. The publishing
+functionality could be extended to new repositories for other domains,
+such as Bioconductor for
+bioinformatics~\cite{gentleman2004bioconductor}.
+\end{quote}
 

### 2018-01-26 boutiquespaper.tex
More revisions and answers

@@
 a repository containing links to neuroinformatics resources and
-tools. This functionality could be extended to new repositories for
-other domains, such as Bioconductor for
+tools. We intentionally opted for publishing descriptors to an index
+such as NeuroLinks rather than to a common \boutiques repository so
+that descriptors remain in the tool repositories and can be maintained
+by the tool developers directly. We are currently extending
+\texttt{bosh} to enable users to search for \boutiques tools in
+NeuroLinks using the command-line or Python API. The publishing
+functionality could be extended to new repositories for other domains,
+such as Bioconductor for
 bioinformatics~\cite{gentleman2004bioconductor}.

### 2018-01-26 answers.tex
Minor fixes

@@
 \newenvironment{answer}%          environment name
-{\textbf{Answer:}\begin{small}\begin{quote}}% begin code
-{\end{quote}\end{small}}%
+{\textbf{Answer:}\begin{quote}}% begin code
+{\end{quote}}%
 
@@
 functionality, however, is a great one and we have discussed the following 
-implementation in \texttt{bosh} over the NeuroLinks index.
+implementation in \texttt{bosh} over the NeuroLinks index:
 \begin{enumerate}
-\item Make Boutiques entries in Neurolinks immutable:
+\item Make Boutiques entries in NeuroLinks immutable:
   \begin{enumerate}
@@
 \begin{quote}
-\emph{Publisher} As Boutiques has primarily been adopted in the neuroinformatics
+\emph{Publisher.} As Boutiques has primarily been adopted in the neuroinformatics
 community, the publisher gets a further description (such as author,
@@
 such as Bioconductor for
-bioinformatics~\cite{gentleman2004bioconductor}.
+bioinformatics.
 \end{quote}

### 2018-02-01 answers.tex
Minor fixes

@@
 The main features of Boutiques mentioned in section ``Other
-frameworkds'' are (1) its support for containerized applications and
+frameworks'' are (1) its support for containerized applications and
 (2) its independence from any particular programming language. While
@@
   \url{https://github.com/boutiques/boutiques/pull/200}.
+
+  We released a
+  new version of Boutiques (0.5.5) that contains the fixes 
+  mentioned above.
 \end{answer}
@@
 \emph{Publisher.} As Boutiques has primarily been adopted in the neuroinformatics
-community, the publisher gets a further description (such as author,
-website, etc.) of the described application, and adds an index to it
+community, the publisher gets more information about the described application (such as author,
+website, etc.), and adds an index to it
 on
@@
 
-\todo{Make a new release and push it to pip}
-
 \end{document}

### 2018-02-01 boutiquespaper.tex
Fixed conflict (II)
Minor fixes

--- 2018-01-26 boutiquespaper.tex
@@
 We present \boutiques, a system to automatically publish, integrate
-and execute command-line applications applications across
+and execute command-line applications across
 computational platforms. \boutiques applications are installed through
@@
 \paragraph{Publisher} As \boutiques has primarily been adopted in the neuroinformatics
-community, the publisher gets a further description (such as author,
-website, etc.) of the described application, and adds an index to it
+community, the publisher gets more information about the described application (such as author,
+website, etc.), and adds an index to it
 on

### 2018-02-01 letter.tex

\documentclass[a4]{letter}
\usepackage{graphicx}
\name{Tristan Glatard}
\signature{\includegraphics[width=4cm]{signature.png}}
\date{January 26th, 2018}

\begin{document}

\begin{letter}{}

\opening{Dear Editors,}

We would like to thank you for offering us the opportunity to revise
and re-submit our paper entitled ``Boutiques: a flexible framework to integrate
command-line applications in computing platforms'' to GigaScience.

The remainder of this letter contains our answer to the 2 reviews that
we received following the evaluation process. A revised version of the
manuscript follows.

We thank you again for your time and consideration, and we remain
available for any further information you may need.

\vspace{0.5cm}

Sincerely yours,

\vspace{1cm}

\closing{Tristan Glatard, PhD.\\On behalf of the authors.}

\end{letter}
\end{document}

### 2018-02-02 boutiquespaper.tex
Changed shell to sh

@@
 The core component of the descriptor is a command-line template
-complying to the syntax of the \texttt{bash} UNIX shell, the default
+complying to the syntax of the \texttt{sh} UNIX shell, the default
 shell on most of the Linux distributions and on OS X. The command-line
@@
 value keys. It may also encompass several commands separated by
-\texttt{bash} constructs such as semicolons, pipes (\texttt{|}) or
+\texttt{sh} constructs such as semicolons, pipes (\texttt{|}) or
 ampersands (\texttt{\&}), to facilitate the embedding of basic

### 2018-02-20 boutiquespaper.tex
Updates requested by the editor

@@
 \documentclass[a4paper,num-refs]{oup-contemporary}
+
+\journal{gigascience}
 
@@
 %%%% Commands %%%%
-
-\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
-\newcommand{\todo}[1]{\color{red}TODO: #1\color{black}}
 \newcommand{\boutiques}{Boutiques\xspace}
-\newcommand{\notimplementedyet}[1]{\color{blue}\emph{#1}\footnote{Still needs to be implemented}\color{black}\xspace}
 \algrenewcommand{\algorithmiccomment}[1]{\# \textit{#1}} % so that comments in algorithms are preprended by '#' instead of a small right arrow
@@
 
+\section{Availability of supporting source code and requirements}
+\begin{itemize}
+\item Project name: Boutiques
+\item Project RRID: SCR\_016073
+\item Project home page: \url{http://boutiques.github.io}
+\item Operating system(s): Platform independent
+\item Programming language: Python and JSON schema
+\item Other requirements: None
+\item License: MIT
+\end{itemize}
+
+\section{Availability of supporting data}
+
+Boutiques is available on GitHub (\url{https://github.com/boutiques/boutiques}) and Zenodo~\cite{tristan_glatard_2018_1164290}. This paper is based on release 0.5.5.
+
 \section{Competing interests}

### 2018-05-09 boutiquespaper.tex
Removed all footnotes, as per the editor's request

@@
 to Open Science. Meanwhile, container systems such as
-Docker\footnote{\url{https://www.docker.com}} and
+Docker (\url{https://www.docker.com}) and
 Singularity~\cite{kurtzer2017singularity} have emerged to facilitate
@@
 on
-NeuroLinks\footnote{\url{https://brainhack101.github.io/neurolinks}},
+NeuroLinks (\url{https://brainhack101.github.io/neurolinks}),
 a repository containing links to neuroinformatics resources and
@@
 an example, we implemented
-NipBIDS\footnote{\url{https://github.com/big-data-lab-team/sim/tree/master/sim/other_wf_examples/nipype}},
+NipBIDS (\url{https://github.com/big-data-lab-team/sim/tree/master/sim/other_wf_examples/nipype}),
 a Nipype workflow to process BIDS datasets using BIDS apps imported as
@@
 Organization (VO) in the European Grid
-Infrastructure\footnote{\url{https://www.egi.eu}}. The biomed VO
+Infrastructure (\url{https://www.egi.eu}). The biomed VO
 interconnects approximately 65 computing sites world-wide and provides
@@
 is deployed at the Creatis
-laboratory\footnote{\url{https://www.creatis.insa-lyon.fr}} in Lyon
+laboratory in Lyon
 and it uses the DIRAC French national
-service\footnote{\url{https://dirac.france-grilles.fr}} to execute
+service (\url{https://dirac.france-grilles.fr}) to execute
 jobs on EGI grid and cloud resources.  As of October 2017, VIP counts
@@
 \texttt{animaN4BiasCorrection}, an ITK-based bias field correction application from the
-Anima\footnote{\url{https://github.com/Inria-Visages/Anima-Public/wiki}}
+Anima (\url{https://github.com/Inria-Visages/Anima-Public/wiki})
 project was made available in VIP through \boutiques.
@@
 We integrated the
-GinFizz\footnote{\url{https://github.com/thomashirsch/ginfizz}}
+GinFizz (\url{https://github.com/thomashirsch/ginfizz})
 Nipype-based fMRI pre-processing pipeline in VIP with \boutiques.  A
@@
 A few applications from the MRtrix3
-package~\cite{tournier2012mrtrix}\footnote{\url{http://www.mrtrix.org/}} for diffusion MRI processing were also made available in VIP,
+package~\cite{tournier2012mrtrix} (\url{http://www.mrtrix.org}) for diffusion MRI processing were also made available in VIP,
 as well as a pipeline developed at Creatis, which combines MRtrix3 and
@@
 
-\paragraph{ODIN} The Odin MRI simulator~\cite{jochimsen2004odin}\footnote{\url{http://od1n.sourceforge.net}} was integrated
+\paragraph{ODIN} The Odin MRI simulator~\cite{jochimsen2004odin} (\url{http://od1n.sourceforge.net}) was integrated
 in VIP through \boutiques. Since Odin requires important amounts of
@@
 images (MSSEG
-challenge\footnote{\url{https://portal.fli-iam.irisa.fr/msseg-challenge/overview}})
+challenge -- \url{https://portal.fli-iam.irisa.fr/msseg-challenge/overview})
 and of tumor volumes in PET images (PETSEG
@@
   required a GPU, which we enabled through the
-  \texttt{nvidia-docker}\footnote{\url{https://github.com/NVIDIA/nvidia-docker}}
+  \texttt{nvidia-docker} (\url{https://github.com/NVIDIA/nvidia-docker})
   tool not supported in Boutiques although it could be a possible extension.
@@
 The Common Workflow Language
-(CWL\footnote{\url{http://www.commonwl.org}}) is the work most closely
+(CWL -- \url{http://www.commonwl.org}) is the work most closely
 related to \boutiques as it provides a formal way to describe
@@
 version 1.0 of the CWL Command Line Tool
-Description\footnote{\url{http://www.commonwl.org/v1.0/CommandLineTool.html}}. According
+Description (\url{http://www.commonwl.org/v1.0/CommandLineTool.html}). According
 to GitHub, CWL started 6 months before \boutiques (September 2014
@@
 command-line applications. Among them, the Common
-Toolkit\footnote{\url{http://www.commontk.org}} interoperates with
+Toolkit (\url{http://www.commontk.org}) interoperates with
 several platforms such as 3D Slicer~\cite{pieper20043d},

### 2018-05-09 boutiquespaper.tex
Fixed other editor's comments

@@
 to Open Science. Meanwhile, container systems such as
-Docker (\url{https://www.docker.com}) and
+\href{https://www.docker.com}{Docker} and
 Singularity~\cite{kurtzer2017singularity} have emerged to facilitate
@@
 on
-NeuroLinks (\url{https://brainhack101.github.io/neurolinks}),
+\href{https://brainhack101.github.io/neurolinks}{NeuroLinks},
 a repository containing links to neuroinformatics resources and
@@
 
-CBRAIN (\url{http://github.com/aces/cbrain}) ~\cite{SHER-14} is a web
+\href{http://github.com/aces/cbrain}{CBRAIN}~\cite{SHER-14} is a web
 platform to process data distributed into multiple storage
@@
 \subsubsection{Nipype}
-Nipype
-(\url{http://nipype.readthedocs.io/en/latest})~\cite{gorgolewski2011nipype}
+\href{http://nipype.readthedocs.io/en/latest}{Nipype}
+~\cite{gorgolewski2011nipype}
 is a workflow engine widely used in neuroinformatics. Nipype workflows
@@
 an example, we implemented
-NipBIDS (\url{https://github.com/big-data-lab-team/sim/tree/master/sim/other_wf_examples/nipype}),
+\href{https://github.com/big-data-lab-team/sim/tree/master/sim/other_wf_examples/nipype}{NipBIDS},
 a Nipype workflow to process BIDS datasets using BIDS apps imported as
@@
 \subsubsection{SPINE}
-SPINE (\url{http://spinevirtuallab.org}), which stands for Structured
+\href{http://spinevirtuallab.org}{SPINE}, which stands for Structured
 Planning and Implementation of New Explorations, is a web-based,
@@
 Organization (VO) in the European Grid
-Infrastructure (\url{https://www.egi.eu}). The biomed VO
+Infrastructure (\href{http://www.egi.eu}{EGI}). The biomed VO
 interconnects approximately 65 computing sites world-wide and provides
@@
 laboratory in Lyon
-and it uses the DIRAC French national
-service (\url{https://dirac.france-grilles.fr}) to execute
+and it uses the \href{https://dirac.france-grilles.fr}{DIRAC} French national
+service to execute
 jobs on EGI grid and cloud resources.  As of October 2017, VIP counts
@@
 \texttt{animaN4BiasCorrection}, an ITK-based bias field correction application from the
-Anima (\url{https://github.com/Inria-Visages/Anima-Public/wiki})
+\href{https://github.com/Inria-Visages/Anima-Public/wiki}{Anima}
 project was made available in VIP through \boutiques.
@@
 facilitate its invocation as a non-interactive command-line 
-application. The resulting CBRAIN plugin is available at
-\url{https://github.com/SIMEXP/cbrain-plugins-psom}.  Descriptors
-are on Neurolinks.
+application. The resulting CBRAIN plugin is available on 
+\href{https://github.com/SIMEXP/cbrain-plugins-psom}{GitHub}.  
+Descriptors are on Neurolinks.
 
@@
 We integrated the
-GinFizz (\url{https://github.com/thomashirsch/ginfizz})
+\href{https://github.com/thomashirsch/ginfizz}{GinFizz}
 Nipype-based fMRI pre-processing pipeline in VIP with \boutiques.  A
@@
 \paragraph{MRTrix3}
-A few applications from the MRtrix3
-package~\cite{tournier2012mrtrix} (\url{http://www.mrtrix.org}) for diffusion MRI processing were also made available in VIP,
+A few applications from the \href{http://www.mrtrix.org}{MRtrix3}
+package~\cite{tournier2012mrtrix} for diffusion MRI processing were also made available in VIP,
 as well as a pipeline developed at Creatis, which combines MRtrix3 and
@@
 pipeline~\cite{kiar2017comprehensive}, developed in Python and leveraging FSL, was
-exported to \boutiques and is available at \url{https://github.com/neurodata/boutiques-tools}.
+exported to \boutiques and is available \href{online}{https://github.com/neurodata/boutiques-tools}.
 The ndmg pipeline is deployed in CBRAIN via its \boutiques descriptor, and is available
@@
 
-\paragraph{ODIN} The Odin MRI simulator~\cite{jochimsen2004odin} (\url{http://od1n.sourceforge.net}) was integrated
+\paragraph{ODIN} The \href{http://od1n.sourceforge.net}{Odin} MRI simulator~\cite{jochimsen2004odin} was integrated
 in VIP through \boutiques. Since Odin requires important amounts of
@@
 2016, related to the segmentation of multiple-sclerosis lesions in MR
-images (MSSEG
-challenge -- \url{https://portal.fli-iam.irisa.fr/msseg-challenge/overview})
+images (\href{https://portal.fli-iam.irisa.fr/msseg-challenge/overview}{MSSEG}
+challenge)
 and of tumor volumes in PET images (PETSEG
@@
   required a GPU, which we enabled through the
-  \texttt{nvidia-docker} (\url{https://github.com/NVIDIA/nvidia-docker})
+  \href{https://github.com/NVIDIA/nvidia-docker}{nvidia-docker}
   tool not supported in Boutiques although it could be a possible extension.
@@
 The Common Workflow Language
-(CWL -- \url{http://www.commonwl.org}) is the work most closely
+(\href{http://www.commonwl.org}{CWL}) is the work most closely
 related to \boutiques as it provides a formal way to describe
@@
 highlights the main differences between CWL and \boutiques, based on
-version 1.0 of the CWL Command Line Tool
-Description (\url{http://www.commonwl.org/v1.0/CommandLineTool.html}). According
+version 1.0 of the CWL \href{http://www.commonwl.org/v1.0/CommandLineTool.html}{Command Line Tool
+Description}. According
 to GitHub, CWL started 6 months before \boutiques (September 2014
@@
 neuroinformatics, many platforms define a formal interface to embed
-command-line applications. Among them, the Common
-Toolkit (\url{http://www.commontk.org}) interoperates with
+command-line applications. Among them, the \href{http://www.commontk.org}{Common
+Toolkit} interoperates with
 several platforms such as 3D Slicer~\cite{pieper20043d},
